---
title: "01. Getting Started with ToolsRTM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Getting Started with ToolsRTM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
```

The goal of this page is the smallest possible example: produce one
vegetation reflectance spectrum with the minimum configuration required.
Everything else in this tutorial series (LUT inversion, parallel
simulation, machine learning) builds on the four-step chain below.

```text
Vegetation parameters
        |
        v
     PROSPECT
        |
        v
Leaf reflectance / transmittance
        |
        v
     fourSAIL
        |
        v
  Canopy reflectance
```

## 1. Leaf traits and canopy structure: one row of parameters

A simulation always starts from one row of trait values -- a `data.frame`
with one column per parameter the leaf and canopy model need. Here,
`PROSPECT-D` needs the leaf biochemistry (`N`, `Cab`, `Car`, `Anth`,
`Cbrown`, `EWT`, `LMA`); `fourSAIL` needs the canopy structure (`LAI`,
`LIDFa`/`LIDFb`/`TypeLidf`, `hspot`) and the sun-view geometry
(`tts`/`tto`/`psi`):

```{r row}
row <- data.frame(
  # Canopy structure and geometry (fourSAIL)
  LAI = 3, hspot = 0.01, LIDFa = -0.35, LIDFb = -0.15, TypeLidf = 1,
  tts = 30, tto = 0, psi = 0,
  # Leaf biochemistry (PROSPECT-D)
  N = 1.5, Cab = 40, Car = 8, Anth = 2, Cbrown = 0, EWT = 0.009, LMA = 0.009,
  alpha = 40
)
```

A soil spectrum is also required -- `foursail()` mixes leaf optics with
whatever shows through the canopy gaps. A 50/50 blend of the package's own
bundled dry/wet reference soil spectra is a reasonable, physically-motivated
default (Tutorial 05 covers building LUTs properly; Tutorial 09's sibling
article on soil/MARMIT integration covers a moisture-realistic alternative):

```{r soil}
rsoil <- 0.5 * dataSpec_PDB[, 11] + 0.5 * dataSpec_PDB[, 12]  # dry/wet soil blend
```

## 2. PROSPECT: leaf reflectance and transmittance

`foursail()` runs PROSPECT internally (via its `LeafModel` argument) and
returns the canopy-level outputs directly -- there's no need to call the
leaf model on its own for this minimal example. `LeafModel = "PROSPECT-D"`
selects the leaf model; Tutorial 02 shows every leaf model available and
how to call PROSPECT on its own when only leaf-level output is needed.

## 3. fourSAIL: canopy reflectance

```{r foursail}
sail <- foursail(inputLUT = row, rsoil = rsoil, LeafModel = "PROSPECT-D")
names(sail)
```

`sail$rdot` (hemispherical-directional reflectance factor) and `sail$rsot`
(bi-directional reflectance factor) are fourSAIL's two raw outputs.
`Compute_BRF()` blends them using the sun/view geometry into the single
top-of-canopy (TOC) reflectance spectrum a sensor actually observes:

```{r brf}
reflectance <- Compute_BRF(rdot = sail$rdot, rsot = sail$rsot,
                            tts = row$tts, data.light = dataSpec_PDB)
length(reflectance)  # one value per nm, 400-2500nm
```

## 4. Plot the result

```{r plot}
plot(dataSpec_PDB[, 1], reflectance, type = "l", col = "#2E8B57", lwd = 2,
     xlab = "Wavelength (nm)", ylab = "Reflectance",
     main = "One simulation: fourSAIL + PROSPECT-D")
```

The familiar vegetation signature is already visible: low reflectance in
the visible (400-700nm, chlorophyll absorption), a sharp rise at the red
edge (~700-750nm), a high NIR plateau (leaf/canopy structural scattering),
and two water-absorption dips in the SWIR (~1450nm, ~1950nm).

## What's next

This page deliberately stopped at one simulation. The rest of this series
builds outward from here:

- **Tutorial 02** -- every leaf model (PROSPECT-D/-PRO, Fluspect, Liberty)
  and every canopy model (fourSAIL, fourSAIL2, INFORM), and how trait
  changes propagate from leaf to canopy.
- **Tutorial 03** -- SPART, when a single TOC simulation like this one
  isn't enough and you need the full soil-plant-atmosphere chain to
  top-of-atmosphere.
- **Tutorial 05** -- building a LUT (many rows of trait values at once)
  instead of one hand-written row.
