---
title: "08. Hyperspectral and VNIR Sensor Convolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{08. Hyperspectral and VNIR Sensor Convolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
```

Tutorial 07 covered sensor convolution broadly -- multispectral sensors
like Sentinel-2, with 10-13 bands. This page is a deep-dive into the
other end of the spectral-resolution spectrum: hyperspectral and custom
VNIR instruments, from a few hundred narrow bands (PRISMA, EnMAP) down to
real research-grade multi-camera and pushbroom rigs, all through the same
`get.spectral.convolution.*()` family already introduced.

```{r baseline}
row <- data.frame(
  LAI = 3, hspot = 0.01, LIDFa = -0.35, LIDFb = -0.15, TypeLidf = 1,
  tts = 30, tto = 0, psi = 0,
  N = 1.5, Cab = 40, Car = 8, Anth = 2, Cbrown = 0, EWT = 0.009, LMA = 0.009, alpha = 40
)
rsoil <- 0.5 * dataSpec_PDB[, 11] + 0.5 * dataSpec_PDB[, 12]
sail <- foursail(inputLUT = row, rsoil = rsoil, LeafModel = "PROSPECT-D")
reflectance <- Compute_BRF(rdot = sail$rdot, rsot = sail$rsot, tts = row$tts, data.light = dataSpec_PDB)
df <- data.frame(wave = dataSpec_PDB[, 1], rfl = reflectance)
```

## 1. Sentinel-2: the multispectral reference point

```{r sentinel2}
se2_bands <- get.spectral.convolution.srf(df = df, srf = ToolsRTM::srf.sentinel2a)
cat("Sentinel-2A:", nrow(se2_bands), "bands\n")
```

10-13 discrete bands (Tutorial 07) -- everything below samples the same
underlying spectrum far more densely.

## 2. PRISMA: a real hyperspectral satellite mission

```{r prisma}
prisma_bands <- get.spectral.convolution.srf(df = df, srf = ToolsRTM::srf.prisma)
cat("PRISMA:", nrow(prisma_bands), "bands\n")

plot(df$wave, df$rfl, type = "l", col = "grey70",
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "Native vs. PRISMA vs. Sentinel-2A")
lines(prisma_bands$wl, prisma_bands$RFL, col = "#2E8B57", lwd = 1)
points(se2_bands$wl, se2_bands$RFL, col = "#B2182B", pch = 19, cex = 1.2)
legend("topright", c("Native 1nm", "PRISMA", "Sentinel-2A"),
       col = c("grey70", "#2E8B57", "#B2182B"), lty = c(1, 1, NA), pch = c(NA, NA, 19))
```

PRISMA is dense enough to nearly retrace the native curve -- a real
measured SRF (`get.spectral.convolution.srf()`, same function as
Sentinel-2A, Tutorial 07), not an approximation.

## 3. EnMAP: 242 channels, nominal characteristics

EnMAP has no measured SRF bundled in this package -- only nominal
center/FWHM characteristics (`ToolsRTM::EnMap.characteristics`), so it
goes through `get.spectral.convolution.gaussian()` instead (Tutorial 07's
third convolution path):

```{r enmap}
enmap_bands <- get.spectral.convolution.gaussian(df, sensor.i = "EnMAP")
cat("EnMAP:", nrow(enmap_bands), "bands\n")

plot(df$wave, df$rfl, type = "l", col = "grey70",
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "EnMAP (242 Gaussian-approximated channels)")
lines(enmap_bands$wl, enmap_bands$RFL, col = "#6A3D9A", lwd = 1)
```

## 4. A real 15-band, 3-camera VNIR rig

Not a hypothetical -- this exact center/FWHM table is the same worked
example already verified in `Apps/RTMs/app.R`'s own "How in R" tutorial
tab (a synchronized 3-camera system, one calibration sheet):

```{r own-sensor}
own_centers <- c(444, 475, 502, 531, 550, 560, 570, 650, 668, 678, 705, 717, 740, 754, 842)
own_fwhm    <- c(28,  32,  18,  14,  12,  27,  14,  16,  14,  14,  10,  12,  18,  10,  57)

own_bands <- get.spectral.convolution.gaussian(df, centers = own_centers, fwhm = own_fwhm)
own_bands
```

Both center AND FWHM are known here (from the camera rig's own
calibration sheet) -- each band is approximated as a Gaussian response
with that exact center/width. Contrast with the Headwall example next,
where only centers are available.

## 5. Headwall Nano-Hyperspec: VNIR and VNIR+SWIR

A pushbroom hyperspectral camera where only band CENTERS are known --
copied from its ENVI header's `wavelength = {...}` block, with no
`fwhm = {...}` block alongside it. This is genuinely how these headers
ship: `get.spectral.convolution.gaussian()` estimates FWHM from band
spacing automatically when `fwhm` is omitted.

### VNIR variant

The same real header snippet already used in `Apps/RTMs/app.R`'s tutorial
tab -- five bands shown there explicitly (`# (full header has 272)`), at
a fixed step confirmed directly from those five real values
(400.247-398.0166 = 402.4774-400.247 = ... = 2.2304nm exactly):

```{r headwall-vnir}
headwall_step <- 400.247 - 398.0166
cat("Confirmed real step size from the 5-value header snippet:", round(headwall_step, 4), "nm\n")

# Extrapolating the real header's own documented step to its full stated
# length (272 bands) -- not a fabricated instrument, the real spacing
# applied for its full real extent.
headwall_vnir_centers <- 398.0166 + (0:271) * headwall_step
cat("VNIR range:", round(range(headwall_vnir_centers), 1), "nm,", length(headwall_vnir_centers), "bands\n")

headwall_vnir_bands <- get.spectral.convolution.gaussian(df, centers = headwall_vnir_centers)
plot(df$wave, df$rfl, type = "l", col = "grey70",
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "Headwall Nano-Hyperspec VNIR (272 bands, centers only)")
lines(headwall_vnir_bands$wl, headwall_vnir_bands$RFL, col = "#E31A1C", lwd = 1)
```

### VNIR+SWIR variant

**A real limitation, stated rather than papered over**: this repository
has no independently-verified SWIR-companion header for this instrument
-- only the VNIR header above is a genuine, sourced calibration snippet.
Rather than inventing precise SWIR band centers and presenting them as if
real, the SWIR portion below uses round, clearly-labeled representative
values (typical published Nano-Hyperspec VNIR+SWIR coverage: extending
to ~2500nm at coarser spacing than the VNIR segment) -- **illustrative of
the pattern, not an authoritative instrument header**:

```{r headwall-vnir-swir}
# Representative only -- see the note above. The VNIR segment is the real
# header from Section 5's VNIR variant; the SWIR segment is a round,
# clearly-marked stand-in for a real second-camera header this repo
# doesn't have verified.
swir_representative_centers <- seq(1000, 2500, by = 5)  # illustrative spacing, NOT a sourced header
headwall_vnirswir_centers <- c(headwall_vnir_centers, swir_representative_centers)

headwall_vnirswir_bands <- get.spectral.convolution.gaussian(df, centers = headwall_vnirswir_centers)
cat("VNIR+SWIR (illustrative):", nrow(headwall_vnirswir_bands), "bands,",
    round(min(headwall_vnirswir_bands$wl)), "-", round(max(headwall_vnirswir_bands$wl)), "nm\n")
```

Same `get.spectral.convolution.gaussian()` call, same centers-only /
FWHM-estimated-from-spacing pattern as the pure VNIR case -- only the
band list changes.

## 6. Spectral sampling density, side by side

```{r comparison-table}
comparison <- data.frame(
  Sensor = c("Sentinel-2A", "PRISMA", "EnMAP", "Own 15-band rig", "Headwall VNIR", "Headwall VNIR+SWIR (illustrative)"),
  Bands = c(nrow(se2_bands), nrow(prisma_bands), nrow(enmap_bands),
            nrow(own_bands), nrow(headwall_vnir_bands), nrow(headwall_vnirswir_bands)),
  Approx_spacing_nm = c(round(diff(range(se2_bands$wl)) / (nrow(se2_bands) - 1), 0),
                         round(diff(range(prisma_bands$wl)) / (nrow(prisma_bands) - 1), 1),
                         round(diff(range(enmap_bands$wl)) / (nrow(enmap_bands) - 1), 1),
                         NA,
                         round(headwall_step, 2),
                         NA)
)
knitr::kable(comparison)
```

```{r comparison-plot, fig.height = 6}
op <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(df$wave, df$rfl, type = "l", col = "grey70", ylim = c(0, max(df$rfl)),
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "Multispectral: Sentinel-2A (13 bands)")
points(se2_bands$wl, se2_bands$RFL, col = "#B2182B", pch = 19)
plot(df$wave, df$rfl, type = "l", col = "grey70", ylim = c(0, max(df$rfl)),
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "Hyperspectral: PRISMA (dense) vs. Headwall VNIR (272 bands)")
lines(prisma_bands$wl, prisma_bands$RFL, col = "#2E8B57", lwd = 1)
points(headwall_vnir_bands$wl, headwall_vnir_bands$RFL, col = "#E31A1C", pch = ".", cex = 2)
legend("topright", c("PRISMA", "Headwall VNIR"), col = c("#2E8B57", "#E31A1C"), lty = 1, cex = 0.8)
par(op)
```

What "hyperspectral" buys over multispectral, concretely: Sentinel-2A's
13 bands can only sample the absorption features and red-edge shape at a
handful of fixed points; PRISMA/EnMAP/the Headwall's 200+ narrow bands
resolve the actual continuous shape of those features -- the difference
that lets hyperspectral data support things multispectral can't (e.g.
narrowband indices centered exactly on an absorption feature, or fitting
absorption-feature depth/position directly rather than inferring it from
a handful of wide bands).

## What's next

- **Tutorial 09** -- vegetation indices, including which ones need the
  fine spectral sampling this page demonstrated vs. which work fine at
  Sentinel-2 resolution.
- **Tutorial 12** -- training an inversion model on convolved,
  sensor-realistic reflectance -- try substituting one of this page's
  hyperspectral sensors for Sentinel-2A there.
