---
title: "09. Understanding RTM Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{09. Understanding RTM Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(ggplot2)
library(dplyr)
```

Tutorial 02 already swept one trait at a time (Cab, EWT, LAI) and watched
the spectrum respond -- one-at-a-time (OAT), the simplest and most
intuitive sensitivity method. This page goes further: **global**
sensitivity analysis, where every trait varies simultaneously (capturing
interactions OAT can't), using three different methods that -- when they
agree -- give real confidence the result reflects the physics, not one
method's quirks.

| Method | Function | Scope |
|---|---|---|
| One-at-a-time (OAT) | plain sweep (Tutorial 02) | Sweep one trait, hold others fixed |
| Sobol total index, per wavelength | `get.spectral.sensitivity()` | Global, full spectrum |
| Sobol Si/Ti, per band | `sensobol::sobol_matrices()` + `foursail()` + `sensobol::sobol_indices()` | Global, discrete sensor bands |
| Johnson relative importance | `sensitivity::johnson()` | Regression-based, per band |

## 1. Global Sobol sensitivity, across the full spectrum

`get.spectral.sensitivity()` wraps a Sobol total-index estimator, run once
per wavelength -- every trait varies simultaneously per simulation, unlike
Tutorial 02's OAT sweeps:

```{r sobol-spectral}
si <- get.spectral.sensitivity(n.samples = 200, distribution = "Uniform",
                                traits = c("Cab", "EWT", "LAI"), wl.step = 10,
                                seed = 1, chunk.size = 200, save.path = tempfile())
head(si)
```

```{r sobol-spectral-plot}
ggplot(si, aes(x = wavelength, y = STi_pct, fill = trait)) +
  geom_area(position = "stack") +
  labs(title = "Spectral sensitivity of TOC reflectance (fourSAIL + PROSPECT-D)",
       subtitle = sprintf("Sobol total sensitivity index, %d simulations", 200),
       x = "Wavelength (nm)", y = "Total SI [%]") +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
```

`Cab` dominates the visible, `EWT` the SWIR water-absorption region --
consistent with Tutorial 02's OAT sweeps, but now with every trait varying
jointly rather than one at a time.

## 2. Global Sobol sensitivity, per Sentinel-2A band

A proper quasi-random Sobol design (`sensobol::sobol_matrices()`), giving
both first-order (Si) and total (Ti) indices per band rather than per
native wavelength:

```{r sobol-perband}
library(sensobol)
params <- c("Cab", "Car", "LAI", "EWT")
N <- 40  # base sample size; total simulations = N * (length(params) + 2)
mat01 <- sobol_matrices(N = N, params = params)
ranges <- list(Cab = c(5, 90), Car = c(0.5, 25), LAI = c(0.5, 7), EWT = c(0.001, 0.035))

LUT_sobol <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = nrow(mat01), setseed = 1234))
for (p in params) LUT_sobol[[p]] <- ranges[[p]][1] + mat01[, p] * diff(ranges[[p]])

rsoil <- rep(0.15, 2101)
refl <- t(sapply(seq_len(nrow(LUT_sobol)), function(i) {
  foursail(inputLUT = LUT_sobol[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D")$rsot
}))
colnames(refl) <- paste0("R.", 400:2500)
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", 400:2500); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

sobol_df <- do.call(rbind, lapply(c("B4", "B8", "B11"), function(band) {
  ind <- sobol_indices(Y = se2a[[band]], N = N, params = params)
  data.frame(Band = band, Parameter = ind$results$parameters, Index = ind$results$original, Type = ind$results$sensitivity)
}))
```

```{r sobol-perband-plot}
ggplot(subset(sobol_df, Type == "Ti"), aes(x = Band, y = Index, fill = Parameter)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Total Sobol indices (Ti) by Sentinel-2A band", x = "Band", y = "Sobol index") +
  theme_bw()
```

`LAI` dominates the NIR band (B8); pigments (`Cab`/`Car`) dominate the red
band (B4) -- exactly what leaf/canopy optics theory predicts.

## 3. Johnson relative importance

A regression-based alternative that doesn't need a special quasi-random
design -- an ordinary random LUT is enough:

```{r johnson}
library(sensitivity)
LUT_j <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 100, setseed = 1234))
LUT_j$Cs <- 0; LUT_j$fqe <- 0.01; LUT_j$Cx <- 0
LUT_j$cell.d <- 40; LUT_j$inter.c <- 0.045; LUT_j$baseline.abs <- 0.0006
LUT_j$leaf.thick <- 1.6; LUT_j$albino.abs <- 0; LUT_j$lign.cell <- 2; LUT_j$Nitrogen <- 1

refl_j <- t(sapply(seq_len(100), function(i) {
  sim <- foursail(inputLUT = LUT_j[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")
  Compute_BRF(rdot = sim$rdot, rsot = sim$rsot, tts = LUT_j$tts[i], data.light = dataSpec_PDB)
}))
refl_X_j <- as.data.frame(refl_j); colnames(refl_X_j) <- paste0("X", 400:2500); refl_X_j <- cbind(id = 1:100, refl_X_j)
se2a_j <- suppressMessages(get.spectra.convolved(rfl = refl_X_j, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_j) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

candidate_traits <- c("Cab", "Car", "Anth", "LMA", "Cbrown", "N", "LIDFa", "LAI", "EWT")
X <- LUT_j[, candidate_traits]
# PROSPECT-PRO leaves LMA constant (uses Prot/CBC for dry matter instead) --
# johnson()'s correlation-matrix step fails outright on a zero-variance
# column, so drop any before calling it.
zero_var <- sapply(X, function(v) sd(v) == 0)
if (any(zero_var)) { cat("Dropping zero-variance trait(s):", paste(names(X)[zero_var], collapse = ", "), "\n"); X <- X[, !zero_var, drop = FALSE] }

johnson_df <- do.call(rbind, lapply(c("B4", "B8", "B11"), function(band) {
  j <- johnson(X, se2a_j[[band]])
  data.frame(Band = band, Parameter = rownames(j$johnson), Index = j$johnson$original)
}))
```

```{r johnson-plot}
ggplot(johnson_df, aes(x = Band, y = Index, fill = Parameter)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Johnson relative-importance indices by Sentinel-2A band", x = "Band", y = "Johnson index") +
  theme_bw()
```

`EWT` dominating B11 (SWIR, a classic water-absorption region) agrees with
Section 2's Sobol result -- two different statistical methods converging
on the same physical story is reassuring, not redundant.

## A practical note on zero-variance predictors

Building the Johnson section above surfaced a real, generically-useful
gotcha: `johnson()`'s correlation-matrix step fails outright
(`eigen(): infinite or missing values`) if any predictor column has zero
variance -- e.g. `LMA` is constant when a LUT is built with
`LeafModel = "PROSPECT-PRO"`, since that variant uses `CBC`/`Prot` for dry
matter instead. Always check for and drop constant columns before feeding
a LUT-derived data.frame into a correlation- or regression-based
sensitivity method.

## What's next

- **Tutorial 10** -- turning "which trait drives which band" into an
  actual trait retrieval: hybrid inversion.
- **Tutorial 04** -- if the question is "which model" rather than "which
  trait", see the model-comparison agreement/RMSE table there instead.
