---
title: "09. Vegetation Indices and Spectral Features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{09. Vegetation Indices and Spectral Features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
```

`ToolsRTM` ships **five** functions with "indices" in the name. They are
not interchangeable, and the package's own naming doesn't make the
distinction obvious -- this page exists to clear that up before using any
of them.

| Function | Input shape | Spectral basis | Index set | Typical use |
|---|---|---|---|---|
| `getIndices()` | Tabular: many rows, `R.<wavelength>` columns (native 1nm) | Native spectrum, domain-limited (`"VNIR"`/`"SWIR"`/`"VNIR-SWIR"`) | ~75 indices, full domain-appropriate set | LUTs at native resolution (Tutorials 05-06) |
| `getIndicesSE2()` | Tabular: many rows, `B1`-`B12` columns (Sentinel-2 bands) | Sentinel-2A/2B band centers | **70** named indices -- the full literature set (NDVI, RDVI, MCARI, REIP, SIPI, ...) | Exploratory analysis: which *specific*, citable index correlates with a trait |
| `getIndicesSE2.ML()` | Same as `getIndicesSE2()` | Same | **31** indices -- a curated subset of the 70, decorrelated/simplified for regression, plus two ML-oriented additions (`kNDVI`, `NDVIv`) not in the full set | Feature engineering for `get.inversion()`/ML models (Tutorials 10-12) |
| `getSpectraIndices()` | **Spatial**: a folder of GeoTIFF images (not a data.frame of spectra) | Sentinel-2A only (currently) | Any one index, or `'All'`, computed per pixel across every image in the folder | Batch/time-series index maps from real imagery |
| `getSpatial_index()` | **Spatial**: one GeoTIFF image at a time | Sentinel-2A only (currently) | Any one index, or `'All'`, computed per pixel | A single-scene version of the above, used internally by `getSpatialTrait()`'s trait-mapping workflow |

Verified directly (same 20-row LUT, convolved to Sentinel-2A, Tutorial
05-07's own pipeline): `getIndicesSE2()` returned 66 non-constant columns
and `getIndicesSE2.ML()` returned 30 on that particular sample -- close to
their full 70/31-index design sets, with a couple dropped as constant or
non-finite for this specific draw (both functions drop those
automatically, same defensive pattern used throughout this package).

## 1. `getIndices()`: native-resolution, tabular

```{r indices-native}
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 20, setseed = 1))
rsoil <- rep(0.15, 2101)
refl <- t(sapply(seq_len(20), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D")$rsot
}))
colnames(refl) <- paste0("R.", 400:2500)

idx_native <- suppressMessages(getIndices(as.data.frame(refl), pattern.rfl = "R.", spectral.domain = "VNIR"))
cat("Columns after adding indices:", ncol(idx_native), "(", ncol(refl), "reflectance +", ncol(idx_native) - ncol(refl), "indices)\n")
```

## 2. `getIndicesSE2()` vs. `getIndicesSE2.ML()`: sensor-band, full vs. curated

Both take Sentinel-2 band-named columns (`B1`...`B12`), not the native
spectrum -- convolve first (Tutorial 07):

```{r convolve}
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", 400:2500); refl_X <- cbind(id = 1:20, refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
```

```{r se2-full}
idx_full <- suppressMessages(getIndicesSE2(df = se2a[, -1], sensor = "Sentinel-2a", df.data = NULL, fast.process = TRUE))
cat("getIndicesSE2() (full set):", ncol(idx_full), "indices\n")
```

```{r se2-ml}
idx_ml <- suppressMessages(getIndicesSE2.ML(df = se2a[, -1], sensor = "Sentinel-2a", df.data = NULL, fast.process = TRUE))
cat("getIndicesSE2.ML() (curated):", ncol(idx_ml), "indices\n")
cat("In the ML set but NOT the full set:", setdiff(names(idx_ml), names(idx_full)), "\n")
```

`kNDVI` (kernel NDVI) and `NDVIv` are the two ML-oriented additions --
`kNDVI` in particular is documented in the remote-sensing literature as
behaving better as a regression predictor than plain NDVI (less
saturation at high biomass), which is exactly the kind of index worth
having in a curated ML feature set even though it isn't one of the
"classic" 70.

```{r cab-relationship, fig.width=6}
plot(LUT$Cab, idx_full$NDVI, pch = 19, col = "#0072B2",
     xlab = "Cab", ylab = "NDVI", main = "One of the 70: NDVI vs. Cab")
```

## 3. Spatial indices: `getSpectraIndices()` and `getSpatial_index()`

Both work on real GeoTIFF imagery, not simulated spectra -- not runnable
here without real Sentinel-2 scenes on disk, but their signatures show the
distinction clearly:

```r
# getSpectraIndices(): a whole FOLDER of images (e.g. a time series),
# looped per file, with a progress bar and optional export -- Sentinel-2A
# only for now.
getSpectraIndices(rasterFiles = "path/to/tif_folder/", Sensor = "Sentinel2a",
                   SpectraltoCompute = "NDVI", factorR = 1/10000,
                   path.export = "path/to/output/")

# getSpatial_index(): ONE image in, one (or 'All') index map(s) out -- no
# folder loop, no progress bar; the building block getSpatialTrait() itself
# calls internally when producing a spatial trait map.
getSpatial_index(rasterFiles = "path/to/one_scene.tif", Sensor = "Sentinel2a",
                  SpectraltoCompute = "NDVI", factorR = 1/10000)
```

Tutorial 15 uses this spatial path for real, on an actual Sentinel-2 image
retrieved via STAC, to produce a genuine 2D index/trait map.

## Which one should you use?

```text
What are you computing indices from?
          |
          +-- A LUT / many simulated spectra at native (1nm) resolution
          |      |
          |      v
          |   getIndices()
          |
          +-- Sentinel-2 band-level reflectance (simulated or real)
          |      |
          |      +-- Exploring which named index matters
          |      |      -> getIndicesSE2()  (full 70-index set)
          |      |
          |      +-- Feeding indices into an ML/inversion model
          |             -> getIndicesSE2.ML()  (31-index curated set)
          |
          +-- Real GeoTIFF imagery (a spatial map, not a spectrum table)
                 |
                 +-- One scene -> getSpatial_index()
                 +-- Many scenes / a time series -> getSpectraIndices()
```

## What's next

- **Tutorial 10** -- formal sensitivity analysis: which traits actually
  drive which indices and bands, and by how much.
- **Tutorial 12** -- using `getIndicesSE2.ML()`'s curated set as ML
  predictor features alongside raw bands.
