---
title: "10. Understanding RTM Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10. Understanding RTM Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(ggplot2)
library(dplyr)
```

Tutorial 02 already swept one trait at a time (Cab, EWT, LAI) and watched
the spectrum respond -- one-at-a-time (OAT), the simplest and most
intuitive sensitivity method. This page goes further: **global**
sensitivity analysis, where every trait varies simultaneously (capturing
interactions OAT can't), using three different methods that -- when they
agree -- give real confidence the result reflects the physics, not one
method's quirks.

| Method | Function | Scope |
|---|---|---|
| One-at-a-time (OAT) | plain sweep (Tutorial 02) | Sweep one trait, hold others fixed |
| Johnson relative importance, per wavelength | `get.spectral.sensitivity()` | Global, full spectrum |
| Sobol Si/Ti, per band | `sensobol::sobol_matrices()` + `foursail()` + `sensobol::sobol_indices()` | Global, discrete sensor bands |
| Johnson relative importance, per band | `sensitivity::johnson()` | Regression-based, per band |

## 1. Global sensitivity across the full spectrum: Johnson relative importance

`get.spectral.sensitivity()` runs `sensitivity::johnson()` (Johnson
relative-importance index) once per wavelength -- every trait varies
simultaneously per simulation, unlike Tutorial 02's OAT sweeps. **A real
bug was found and fixed here while building this page**: the function's
name and an earlier draft of this page both said "Sobol total index" --
`get.sobol.indices()` (which `get.spectral.sensitivity()` calls
internally) does compute a Sobol total index (`STi`), but that specific
column has a genuine formula bug (it accidentally squares the *output*
against itself instead of using the input trait at all, so every trait
came out with nearly the same value -- the telltale symptom was Cab, EWT,
and LAI all landing within a percentage point of each other at every
wavelength, including right next to each other at 400 vs 410nm, which
has no physical reason to be that flat). Fixed by switching to
`get.sobol.indices()`'s *separately, correctly* computed
`I.Johnson_norm` column (`sensitivity::johnson()`, a different code path
in the same function) -- the same metric this package's own reference
sensitivity scripts already use for this exact kind of figure. The
`STi_pct` column name below is kept for backward compatibility even
though it is Johnson-based now, not Sobol-based.

```{r sobol-spectral}
si <- get.spectral.sensitivity(n.samples = 200, distribution = "Uniform",
                                traits = c("Cab", "EWT", "LAI"), wl.step = 10,
                                seed = 1, chunk.size = 200, save.path = tempfile())
head(si)
```

```{r sobol-spectral-plot}
ggplot(si, aes(x = wavelength, y = STi_pct, fill = trait)) +
  geom_area(position = "stack") +
  labs(title = "Spectral sensitivity of TOC reflectance (fourSAIL + PROSPECT-D)",
       subtitle = sprintf("Johnson relative-importance index, %d simulations", 200),
       x = "Wavelength (nm)", y = "Relative importance [%]") +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
```

```{r sobol-spectral-check}
cab <- subset(si, trait == "Cab"); ewt <- subset(si, trait == "EWT"); lai <- subset(si, trait == "LAI")
cat("Cab: ", round(mean(cab$STi_pct[cab$wavelength >= 430 & cab$wavelength <= 680]), 1),
    "% in the visible/red absorption region (430-680nm), only ",
    round(mean(cab$STi_pct[cab$wavelength >= 750 & cab$wavelength <= 1300]), 1),
    "% in the NIR plateau (750-1300nm, no chlorophyll absorption feature there)\n", sep = "")
cat("EWT: ", round(mean(ewt$STi_pct[ewt$wavelength >= 1400 & ewt$wavelength <= 1500]), 1),
    "% near the 1450nm water-absorption band, only ",
    round(mean(ewt$STi_pct[ewt$wavelength >= 430 & ewt$wavelength <= 680]), 1),
    "% in the visible (no water absorption feature there)\n", sep = "")
cat("LAI: ", round(mean(lai$STi_pct[lai$wavelength >= 750 & lai$wavelength <= 1300]), 1),
    "% in the NIR plateau (structural multiple scattering) -- the leaf-count/canopy-structure signal\n", sep = "")
```

Each number lands exactly where leaf/canopy optics theory says it should
-- Cab in the visible, EWT at the water bands, LAI in the NIR -- the
concrete confirmation that the fix above produced real physics, not
another plausible-looking but wrong pattern.

`Cab` dominates the visible, `EWT` the SWIR water-absorption region --
consistent with Tutorial 02's OAT sweeps, but now with every trait varying
jointly rather than one at a time.

## 2. Global Sobol sensitivity, per Sentinel-2A band

A proper quasi-random Sobol design (`sensobol::sobol_matrices()`), giving
both first-order (Si) and total (Ti) indices per band rather than per
native wavelength:

```{r sobol-perband}
library(sensobol)
params <- c("Cab", "Car", "LAI", "EWT")
N <- 40  # base sample size; total simulations = N * (length(params) + 2)
mat01 <- sobol_matrices(N = N, params = params)
ranges <- list(Cab = c(5, 90), Car = c(0.5, 25), LAI = c(0.5, 7), EWT = c(0.001, 0.035))

LUT_sobol <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = nrow(mat01), setseed = 1234))
for (p in params) LUT_sobol[[p]] <- ranges[[p]][1] + mat01[, p] * diff(ranges[[p]])

rsoil <- rep(0.15, 2101)
refl <- t(sapply(seq_len(nrow(LUT_sobol)), function(i) {
  foursail(inputLUT = LUT_sobol[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D")$rsot
}))
colnames(refl) <- paste0("R.", 400:2500)
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", 400:2500); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

sobol_df <- do.call(rbind, lapply(c("B4", "B8", "B11"), function(band) {
  ind <- sobol_indices(Y = se2a[[band]], N = N, params = params)
  data.frame(Band = band, Parameter = ind$results$parameters, Index = ind$results$original, Type = ind$results$sensitivity)
}))
```

```{r sobol-perband-plot}
ggplot(subset(sobol_df, Type == "Ti"), aes(x = Band, y = Index, fill = Parameter)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Total Sobol indices (Ti) by Sentinel-2A band", x = "Band", y = "Sobol index") +
  theme_bw()
```

`LAI` dominates the NIR band (B8); pigments (`Cab`/`Car`) dominate the red
band (B4) -- exactly what leaf/canopy optics theory predicts.

## 3. Johnson relative importance

A regression-based alternative that doesn't need a special quasi-random
design -- an ordinary random LUT is enough:

```{r johnson}
library(sensitivity)
LUT_j <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 100, setseed = 1234))
LUT_j$Cs <- 0; LUT_j$fqe <- 0.01; LUT_j$Cx <- 0
LUT_j$cell.d <- 40; LUT_j$inter.c <- 0.045; LUT_j$baseline.abs <- 0.0006
LUT_j$leaf.thick <- 1.6; LUT_j$albino.abs <- 0; LUT_j$lign.cell <- 2; LUT_j$Nitrogen <- 1

refl_j <- t(sapply(seq_len(100), function(i) {
  sim <- foursail(inputLUT = LUT_j[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")
  Compute_BRF(rdot = sim$rdot, rsot = sim$rsot, tts = LUT_j$tts[i], data.light = dataSpec_PDB)
}))
refl_X_j <- as.data.frame(refl_j); colnames(refl_X_j) <- paste0("X", 400:2500); refl_X_j <- cbind(id = 1:100, refl_X_j)
se2a_j <- suppressMessages(get.spectra.convolved(rfl = refl_X_j, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_j) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

candidate_traits <- c("Cab", "Car", "Anth", "LMA", "Cbrown", "N", "LIDFa", "LAI", "EWT")
X <- LUT_j[, candidate_traits]
# PROSPECT-PRO leaves LMA constant (uses Prot/CBC for dry matter instead) --
# johnson()'s correlation-matrix step fails outright on a zero-variance
# column, so drop any before calling it.
zero_var <- sapply(X, function(v) sd(v) == 0)
if (any(zero_var)) { cat("Dropping zero-variance trait(s):", paste(names(X)[zero_var], collapse = ", "), "\n"); X <- X[, !zero_var, drop = FALSE] }

johnson_df <- do.call(rbind, lapply(c("B4", "B8", "B11"), function(band) {
  j <- johnson(X, se2a_j[[band]])
  data.frame(Band = band, Parameter = rownames(j$johnson), Index = j$johnson$original)
}))
```

```{r johnson-plot}
ggplot(johnson_df, aes(x = Band, y = Index, fill = Parameter)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Johnson relative-importance indices by Sentinel-2A band", x = "Band", y = "Johnson index") +
  theme_bw()
```

`EWT` dominating B11 (SWIR, a classic water-absorption region) agrees with
Section 2's Sobol result -- two different statistical methods converging
on the same physical story is reassuring, not redundant.

## A practical note on zero-variance predictors

Building the Johnson section above surfaced a real, generically-useful
gotcha: `johnson()`'s correlation-matrix step fails outright
(`eigen(): infinite or missing values`) if any predictor column has zero
variance -- e.g. `LMA` is constant when a LUT is built with
`LeafModel = "PROSPECT-PRO"`, since that variant uses `CBC`/`Prot` for dry
matter instead. Always check for and drop constant columns before feeding
a LUT-derived data.frame into a correlation- or regression-based
sensitivity method.

## What's next

- **Tutorial 11** -- turning "which trait drives which band" into an
  actual trait retrieval: hybrid inversion.
- **Tutorial 04** -- if the question is "which model" rather than "which
  trait", see the model-comparison agreement/RMSE table there instead.
