---
title: "12. Comparing ML Algorithms for RTM Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{12. Comparing ML Algorithms for RTM Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

Tutorial 11 introduced `get.inversion()` in passing (Random Forest only).
This page compares several of its 12 supported algorithms (via `caret`) on
the same LUT/train-test split, so the comparison is apples-to-apples.

## 1. Same LUT, same split as Tutorial 11

```{r simulate}
n_samples <- 150
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
names(se2a) <- c("id", band_names)

set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
test_idx  <- setdiff(seq_len(n_samples), train_idx)
train_df <- cbind(LUT[train_idx, ], se2a[train_idx, band_names])
test_df  <- cbind(LUT[test_idx, ],  se2a[test_idx,  band_names])

r2_f   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))
```

## 2. Fit several algorithms, predict Cab on the held-out test set

`get.inversion()` supports 12 algorithms: `"PLSR"`, `"SVM"`, `"RF"`,
`"GB"`, `"NN"`, `"Bayesian"`, `"AdaBag"`, `"BRNN"`, `"xGB"`, `"RVM"`,
`"qLASSO"`, `"Ensemble"`. `"NN"` (caret's `nnet` tuning grid) is skipped
here -- impractically slow against this many predictors at only 105
training rows, the same reason this package's own course pipeline scripts
(`Scripts/R/*/2-inversion_ML.R`) skip it too. A representative subset run
for real below; the rest of the call is identical for any of the others.

**`"Ensemble"` (stacks SVM + Gradient Boosting + a neural net via
`caretEnsemble`) is left out of the comparison below on purpose**: while
verifying this page, it surfaced two real bugs in `get.inversion()`'s
source (both fixed directly in the package for this release --
`fmla.n` was referenced without being defined in that branch, and its
internal `stackControl` passed the whole training data.frame to
`caret::createFolds()` instead of the response column, misaligning fold
sizes) -- but even after both fixes, `caretEnsemble::caretStack()` still
fails with `"pred_rows == pred_rows[1L] are not all TRUE"`, a deeper
row-alignment mismatch across the three stacked sub-models' predictions
that needs further investigation into how their tuning grids interact,
not something safe to guess-fix here.

```{r fit-algorithms, results='hide'}
algorithms <- c("PLSR", "SVM", "RF", "GB")
fits <- lapply(algorithms, function(algo) {
  get.inversion(data = train_df, depVar = "Cab", inputs = band_names,
                algorithm = algo, n.samples = nrow(train_df), seed = 42)
})
names(fits) <- algorithms
```

```{r evaluate}
metrics <- do.call(rbind, lapply(algorithms, function(algo) {
  pred <- as.numeric(predict(fits[[algo]]$model, newdata = test_df[, c("Cab", band_names)]))
  data.frame(algorithm = algo, R2 = r2_f(test_df$Cab, pred), RMSE = rmse_f(test_df$Cab, pred))
}))
knitr::kable(metrics[order(-metrics$R2), ], digits = 3, row.names = FALSE)
```

```{r best-plot}
best_algo <- metrics$algorithm[which.max(metrics$R2)]
pred_best <- as.numeric(predict(fits[[best_algo]]$model, newdata = test_df[, c("Cab", band_names)]))
plot(test_df$Cab, pred_best, pch = 19, col = "#2166AC",
     xlab = "Observed Cab", ylab = paste("Predicted Cab (", best_algo, ")"),
     main = paste("Best algorithm on this LUT:", best_algo))
abline(0, 1, col = "grey40", lty = 2)
```

## 3. Comparing against Tutorial 11's merit-function matching

```{r vs-opt, results='hide'}
se2a_mat <- as.matrix(se2a[, band_names])
opt_rmse <- get.inversionOpt(rfl.sensor = se2a_mat[test_idx, ], rfl.rtm = se2a_mat[train_idx, ],
                              LUT = LUT[train_idx, ], wave = as.numeric(sub("B", "", band_names)),
                              method = "merit-RMSE", nOpt = 5)
```

```{r vs-opt-table}
comparison <- rbind(metrics[, c("algorithm", "R2", "RMSE")],
                     data.frame(algorithm = "inversionOpt (merit-RMSE)",
                                R2 = r2_f(LUT[test_idx, ]$Cab, opt_rmse[[2]]$Cab),
                                RMSE = rmse_f(LUT[test_idx, ]$Cab, opt_rmse[[2]]$Cab)))
knitr::kable(comparison[order(-comparison$R2), ], digits = 3, row.names = FALSE)
```

At this LUT size (150 rows, 105 training), a fitted ML model and pure
LUT-search inversion are often close -- the gap widens in the ML model's
favour with more training data, and in `get.inversionOpt()`'s favour when
too little data is available to fit a model reliably (Tutorial 11's
discussion of when to use which).

## What's next

- **Tutorial 13** -- deep learning (`getMLmodel()`) on the same kind of
  data, and when it's worth the extra complexity over the algorithms here.
- **Tutorial 14** -- this whole simulate-convolve-invert chain as one
  coherent pipeline, across every canopy model this package supports.
