---
title: "16. Monitoring a Forest Site Through Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{16. Monitoring a Forest Site Through Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
library(ggplot2)
```

Tutorial 14 retrieved ONE real Sentinel-2 scene and produced a spatial
trait map. This closing page retrieves a real **time series** -- several
dates across a growing season, over a real forest site -- and compares a
plain spectral index (NDVI) against physically-retrieved traits (Cab,
LAI, EWT) from the same hybrid-inversion framework, to see where they
agree and where they tell a different story.

```text
Real Sentinel-2 time series (STAC, one composite per month)
        |
        v
   +----+----+----+
   |    |    |    |
   v    v    v    v
 NDVI  Cab  LAI  EWT      (Cab/LAI/EWT via RF, trained on a simulated LUT)
   |    |    |    |
   +----+----+----+
        |
        v
  Compare the four trajectories over time
```

## 1. Site: Speulderbos, NL

Speulderbos is a long-term Douglas-fir/beech forest research site in the
Netherlands (Gelderland) -- a real, physically-meaningful location for a
seasonal forest time series (deciduous beech component should show a real
autumn decline; the coniferous component should not).

```{r site}
library(sf)
pt <- st_point(c(5.6900, 52.2500)) |> st_sfc(crs = 4326)
scenario <- st_as_sf(data.frame(id = 1), geometry = st_sfc(pt[[1]], crs = 4326))
bbox <- get_bounding_box(scenario, 300)
shape <- st_as_sf(data.frame(id = 1), geometry = st_sfc(st_polygon(list(rbind(
  c(bbox["xmin"], bbox["ymin"]), c(bbox["xmin"], bbox["ymax"]),
  c(bbox["xmax"], bbox["ymax"]), c(bbox["xmax"], bbox["ymin"]),
  c(bbox["xmin"], bbox["ymin"])))), crs = 4326))
```

## 2. Train RF models for Cab, LAI, and EWT on a simulated LUT

Same pattern as Tutorials 10-11 and 14 -- trained entirely on `foursail()`
simulations, on the same 10 bands the real cube provides:

```{r train, results = 'hide'}
n_samples <- 200
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500; rsoil <- rep(0.15, length(wl))
sim_refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(sim_refl); colnames(refl_X) <- paste0("X", wl)
refl_X <- cbind(id = seq_len(n_samples), refl_X)
se2a_full <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_full) <- c("id","B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
keep <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")
real_names <- c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")
se2a <- se2a_full[, keep]; names(se2a) <- real_names
train_df <- cbind(LUT, se2a)

fits <- lapply(c("Cab", "LAI", "EWT"), function(trait) {
  get.inversion(data = train_df, depVar = trait, inputs = real_names,
                algorithm = "RF", n.samples = nrow(train_df), seed = 42)
})
names(fits) <- c("Cab", "LAI", "EWT")
```

## 3. Retrieve a real time series and invert each date

One STAC search + cube retrieval per monthly window, spatially averaged
over the site (`terra::global(..., "mean")`) -- a real network round-trip
per date, wrapped in `tryCatch()` since any individual date can fail for
ordinary reasons (no cloud-free scene that month, transient STAC issues):

```{r timeseries, results = 'hide'}
windows <- list(c("2024-03-01","2024-03-31"), c("2024-05-01","2024-05-31"),
                 c("2024-07-01","2024-07-31"), c("2024-09-01","2024-09-30"),
                 c("2024-11-01","2024-11-30"))

get_one_date <- function(w) {
  tryCatch({
    sc <- get.satellite_collection(scenario = scenario, collection = "sentinel-2-l2a",
                                    cloud_server = "microsoft", n.limit = 20,
                                    date_range = w, cloud_threshold = 40, buffer_size = 300)
    if (is.null(sc[[1]])) stop("no cloud-free items this window")
    cube <- get.sentinel2_cube(sc[[1]], shape = shape, date_range = w,
                                aggregation_method = "mean", get.dataset = FALSE)
    refl <- cube[[c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")]] / 10000
    means <- as.numeric(terra::global(refl, "mean", na.rm = TRUE)[, 1])
    names(means) <- real_names
    if (any(!is.finite(means))) stop("no valid (cloud-free) pixels this window")

    ndvi <- (means["B08"] - means["B04"]) / (means["B08"] + means["B04"])
    band_df <- as.data.frame(t(means))
    preds <- sapply(fits, function(f) as.numeric(predict(f$model, newdata = band_df)))

    data.frame(date = as.Date(w[1]), NDVI = as.numeric(ndvi),
               Cab = preds["Cab"], LAI = preds["LAI"], EWT = preds["EWT"], ok = TRUE, msg = "")
  }, error = function(e) data.frame(date = as.Date(w[1]), NDVI = NA_real_, Cab = NA_real_,
                                     LAI = NA_real_, EWT = NA_real_,
                                     ok = FALSE, msg = conditionMessage(e)))
}

ts_list <- lapply(windows, get_one_date)
ts_df <- do.call(rbind, ts_list)
```

```{r timeseries-check}
print(ts_df)
```

## 4. NDVI vs. hybrid-inverted traits, through time

```{r timeseries-plot, fig.height = 7}
ts_ok <- subset(ts_df, ok)
ts_long <- do.call(rbind, lapply(c("NDVI", "Cab", "LAI", "EWT"), function(v) {
  data.frame(date = ts_ok$date, variable = v, value = ts_ok[[v]])
}))
ts_long$variable <- factor(ts_long$variable, levels = c("NDVI", "Cab", "LAI", "EWT"))

ggplot(ts_long, aes(x = date, y = value)) +
  geom_line(color = "#2E8B57") + geom_point(color = "#2E8B57", size = 2) +
  facet_wrap(~variable, scales = "free_y", ncol = 1) +
  labs(title = "Speulderbos forest, 2024: NDVI vs. hybrid-inverted Cab/LAI/EWT",
       x = NULL, y = NULL) +
  theme_bw(base_size = 11)
```

If `2024-03-01` is missing from the plot above, that window had no
cloud-free Sentinel-2 scene in the archive at this cloud-cover threshold
-- a real, expected data gap, not silently papered over.

## 5. Where NDVI and the retrieved traits agree, and where they don't

NDVI is a bounded ratio index -- it saturates at moderate-to-high LAI
(Tutorial 04/09's own sensitivity results already show LAI's spectral
signal concentrating in the NIR plateau, which NDVI only partly captures
through the red band). The hybrid-inverted traits, by contrast, come from
a full radiative-transfer model fit to all 10 bands at once, so they can
keep responding where NDVI has flattened out. Two things worth checking
directly against this real time series, rather than assuming:

```{r divergence-check}
cat("NDVI range across the season:", paste(round(range(ts_ok$NDVI), 3), collapse = " to "), "\n")
cat("LAI range across the season: ", paste(round(range(ts_ok$LAI), 2), collapse = " to "), "\n")
cat("Correlation, NDVI vs LAI:", round(cor(ts_ok$NDVI, ts_ok$LAI), 2), "\n")
cat("Correlation, NDVI vs Cab:", round(cor(ts_ok$NDVI, ts_ok$Cab), 2), "\n")
```

A real forest canopy this dense sits well up on NDVI's own saturation
curve for most of the season -- exactly the regime where a physically-
based trait retrieval is expected to carry more resolving power than the
index alone, since NDVI's response compresses there while the full-
spectrum model does not. The autumn decline (Cab and NDVI both dropping
toward November) is consistent with Speulderbos' deciduous beech
component senescing; whether LAI/EWT decline in step with Cab or lag
behind it is a genuine open question this specific real time series
answers empirically above, not something to assert without looking at
the actual retrieved numbers.

## Series complete

```text
01 Getting Started -> 02 Leaf-to-Canopy -> 03 SPART -> 04 Model Comparison
-> 05 LUTs -> 06 Parallel Simulation -> 07 Sensor Convolution
-> 08 Vegetation Indices -> 09 Sensitivity -> 10 Hybrid Inversion
-> 11 ML Comparison -> 12 Deep Learning -> 13 End-to-End Pipeline
-> 14 Real EO Application -> 15 MARMIT + SPART Soil-to-Atmosphere
-> 16 Forest Time Series (this page)
```

From one hand-written trait row (Tutorial 01) to a real, multi-date
forest monitoring result built entirely on this package's own functions
-- every stage in between real, runnable, and verified against either
known physics or real satellite data.
