---
title: "Trait inversion: merit-function matching, ML, and deep learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trait inversion: merit-function matching, ML, and deep learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ToolsRTM)
```

## The problem: given a sensor spectrum, retrieve a trait

Every method on this page solves the same task -- given ONLY the sensor-band
reflectance a satellite or field spectrometer would measure, retrieve a
biophysical trait (chlorophyll content, `Cab`, here) without knowing the
ground truth -- but they get there in fundamentally different ways:

- **`get.inversionOpt()`** (this page's main subject): no model is fit at
  all. It ranks every spectrum in a reference LUT by how closely it matches
  each observed spectrum under a chosen merit function, then averages the
  trait values of the `nOpt` best matches. Pure library search.
- **`get.inversion()`**: fits a statistical/ML model (Random Forest, PLSR,
  SVM, ... 12 algorithms via `caret`) on a training LUT, then predicts on
  new spectra.
- **`getMLmodel()`**: fits a deep-learning model (dense or 1D-CNN, via
  TensorFlow/Keras) on the same kind of data.

To compare them fairly, all three are evaluated on the exact same held-out
test spectra below.

## 1. Simulate a LUT and convolve to Sentinel-2A

```{r simulate}
n_samples <- 150
inputs <- ToolsRTM::inputsPROSAIL
LUT <- as.data.frame(getLUT(inputs = inputs, nLUT = n_samples, setseed = 1))

wl <- 400:2500
rsoil <- rep(0.15, length(wl))

refl <- t(sapply(seq_len(n_samples), function(i) {
  sim <- simulate_RTM(inputLUT = LUT[i, ], rsoil = rsoil,
                       leaf.model = "PROSPECT-PRO", canopy.model = "fourSAIL")
  sim$rsot
}))

refl_X <- as.data.frame(refl)
colnames(refl_X) <- paste0("X", wl)
refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)

se2a <- get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE)
wl_bands <- as.numeric(names(se2a)[-1])
band_names <- paste0("B", seq_along(wl_bands))
names(se2a) <- c("id", band_names)
```

150 simulated spectra, convolved down to Sentinel-2A's 13 real bands --
this is what all three inversion methods below actually see, not the
native 1nm spectrum.

## 2. A train/test split, shared by every method

The same 70/30 split feeds `get.inversionOpt()` (as its reference LUT vs.
its "observed" test spectra), `get.inversion()` (as training data vs.
prediction targets), and `getMLmodel()` -- so the comparison at the end is
apples-to-apples.

```{r split}
set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
test_idx  <- setdiff(seq_len(n_samples), train_idx)

LUT_train <- LUT[train_idx, ]
LUT_test  <- LUT[test_idx, ]
se2a_mat  <- as.matrix(se2a[, band_names])
se2a_train_mat <- se2a_mat[train_idx, ]
se2a_test_mat  <- se2a_mat[test_idx, ]

train_df <- cbind(LUT_train, se2a[train_idx, band_names])
test_df  <- cbind(LUT_test,  se2a[test_idx,  band_names])

r2_f   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))

cat("Train:", length(train_idx), "spectra. Test (held out):", length(test_idx), "spectra.\n")
```

## 3. `get.inversionOpt()`: LUT merit-function matching

`method` picks how "closeness" between two spectra is measured. `nOpt`
controls how many of the closest training spectra get averaged together
for the final trait estimate (`nOpt = 5` below -- a compromise between a
single nearest neighbour, which is noisy, and averaging too many, which
smooths out real trait variation):

```{r inversionopt-methods, results='hide'}
opt_rmse <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-RMSE", nOpt = 5)
opt_fge  <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-FGE", nOpt = 5)
opt_dwt  <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-DWT", nOpt = 5)
```

Each call returns a 2-element list: `[[1]]` (`rfl.b`) is the matched
reflectance itself, `[[2]]` (`LUT.best`) is the corresponding trait table
-- including the merit-function's own error value (`merit-RMSE`,
`merit-FGE`, ...) and the `Cab` estimate we're evaluating below:

```{r inversionopt-check}
opt_metrics <- data.frame(
  method = c("merit-RMSE", "merit-FGE", "merit-DWT"),
  R2   = c(r2_f(LUT_test$Cab, opt_rmse[[2]]$Cab), r2_f(LUT_test$Cab, opt_fge[[2]]$Cab), r2_f(LUT_test$Cab, opt_dwt[[2]]$Cab)),
  RMSE = c(rmse_f(LUT_test$Cab, opt_rmse[[2]]$Cab), rmse_f(LUT_test$Cab, opt_fge[[2]]$Cab), rmse_f(LUT_test$Cab, opt_dwt[[2]]$Cab))
)
knitr::kable(opt_metrics, digits = 3)
```

`merit-RMSE`/`merit-FGE` compare raw band reflectance; `merit-DWT`
compares each spectrum's discrete wavelet transform coefficients instead
(sensitive to the overall shape of the spectrum rather than each band's
exact value -- can be more robust to band-to-band noise). `merit-NRMSE`,
`merit-MAE`, `merit-NMB` and `merit-1stD` (first-derivative matching, more
sensitive to absorption-feature position/shape than absolute reflectance)
are the remaining built-in options; `custom_stat` accepts your own
`function(sim, obs)` returning a single error value in place of any of
them.

```{r inversionopt-plot, fig.width=5, fig.height=5}
plot(LUT_test$Cab, opt_rmse[[2]]$Cab, pch = 19, col = "#2166AC",
     xlab = "Observed Cab", ylab = "Predicted Cab (merit-RMSE, nOpt=5)",
     main = "LUT merit-function matching")
abline(0, 1, col = "grey40", lty = 2)
```

## 4. `get.inversion()`: machine learning (Random Forest)

Same train/test split, but now a model is actually fit on `train_df`
first, then applied to `test_df` -- no per-observation LUT search at
prediction time:

```{r ml-inversion, results='hide'}
ml_res <- get.inversion(data = train_df, depVar = "Cab", inputs = band_names,
                         algorithm = "RF", n.samples = nrow(train_df), seed = 42)
pred_ml_test <- as.numeric(predict(ml_res$model, newdata = test_df[, c("Cab", band_names)]))
```

```{r ml-check}
cat("ML (RF)  R2:", round(r2_f(LUT_test$Cab, pred_ml_test), 3),
    " RMSE:", round(rmse_f(LUT_test$Cab, pred_ml_test), 3), "\n")
```

`algorithm = "RF"` can be swapped for any of `get.inversion()`'s 12
supported values (`"PLSR"`, `"SVM"`, `"GB"`, `"NN"`, `"Bayesian"`,
`"AdaBag"`, `"BRNN"`, `"xGB"`, `"RVM"`, `"qLASSO"`, `"Ensemble"`) -- the
rest of the call is identical.

## 5. `getMLmodel()`: deep learning (TensorFlow/Keras)

Needs a working Python/TensorFlow/tf-keras stack reachable through
`reticulate` -- genuinely machine-specific setup (see
`ToolsRTM_PROSAIL_tutorial.Rmd`'s "Deep learning" section for the full
install sequence), so this is wrapped in `tryCatch()` to keep the rest of
this vignette buildable on a machine without that stack configured, while
still showing the real, correct calling code:

```{r dl-inversion}
dl_result <- tryCatch({
  Sys.setenv(TF_USE_LEGACY_KERAS = "1")
  library(reticulate); library(keras)
  dl_model <- getMLmodel(dataset = train_df, depVar = "Cab", model = "Hidden-layers",
                          optimizer = "adam", n.epochs = 5)
  list(ok = TRUE, model = dl_model)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))

if (dl_result$ok) {
  cat("DL model trained. Final training loss:",
      round(tail(dl_result$model$history$metrics$loss, 1), 3), "\n")
} else {
  first_line <- strsplit(dl_result$msg, "\n", fixed = TRUE)[[1]][1]
  cat("Skipped -- no working Python/TensorFlow/tf-keras stack reachable through",
      "reticulate on this machine (", first_line, "). The calling code above is",
      "correct and unchanged; see ToolsRTM_PROSAIL_tutorial.Rmd for setup.\n")
}
```

## 6. Which one should you use?

| Method | Needs training? | Handles a small LUT well? | Typical use |
|---|---|---|---|
| `get.inversionOpt()` (merit-function) | No -- pure search | Yes -- works from a handful of reference spectra | Quick, physically-grounded retrieval; no ML infrastructure needed |
| `get.inversion()` (classic ML) | Yes | Needs enough rows to fit reliably (hundreds+) | Production-scale retrieval, many predictors (e.g. full spectrum + indices) |
| `getMLmodel()` (deep learning) | Yes, more data-hungry | No -- needs thousands of rows | Large training sets, CNN over the full spectral shape |

All three were run on the exact same held-out Sentinel-2A test spectra
above -- see Sections 3-4 for their actual R²/RMSE on this LUT (deep
learning's result, when the optional TensorFlow stack is available, prints
its own training loss above rather than a directly comparable R²/RMSE,
since 5 epochs on 105 training rows is a smoke-test, not a tuned model).
