---
title: "03. SPART: Soil-Plant-Atmosphere Radiative Transfer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. SPART: Soil-Plant-Atmosphere Radiative Transfer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
```

Tutorials 01-02 stopped at top-of-canopy (TOC) reflectance -- what a
sensor would see immediately above the canopy, with no atmosphere in the
way. `SPART()` goes one step further: soil, canopy, AND atmosphere,
together, giving top-of-atmosphere (TOA) reflectance -- what a real
satellite actually measures. It deserves its own page rather than a
paragraph inside a leaf/canopy comparison, because it represents a more
complete modelling chain, not just another canopy option. This page
stays focused on that unique chain (soil -> canopy -> atmosphere -> TOA)
-- soil-moisture spectroscopy in depth is Tutorial 16's job, not this
one's.

```text
                 Atmosphere
                     |
              Incoming radiation
                     |
Soil ---------> Vegetation canopy
                     |
              Canopy reflectance
                     |
                 Atmosphere
                     |
           Top-of-atmosphere signal
```

SPART couples three sub-models: **BSM** (Brightness-Shape-Moisture) for
soil, **fourSAIL** for the vegetation canopy, and **SMAC** for atmospheric
effects -- the same three-model architecture SPART's own reference
implementation (Yang et al. 2020) uses.

## 3.1 Configure SPART

Inputs fall into five physical domains. `inputsSPART` (a `getLUT()`-ready
table, same pattern as `inputsPROSAIL`) already groups them:

| Domain | Example columns | Notes |
|---|---|---|
| Leaf | `Cab`, `Car`, `Anth`, `LMA`, `EWT`, `N`, ... | Same as PROSPECT/fourSAIL (Tutorial 02) |
| Canopy structure | `LAI`, `LIDFa`/`LIDFb`/`TypeLidf`, `hspot` | Same as fourSAIL |
| Soil (BSM) | `BSMBrightness`, `BSMlat`, `BSMlon`, `SMp` | See Section 3.3 -- these are BSM's real parameters |
| Geometry | `tts`, `tto`, `psi` | Sun zenith, view zenith, relative azimuth |
| Atmosphere | `Pa`, `aot550`, `uo3`, `uh2o`, `alt_m`, `Pa0` | Air pressure, aerosol optical thickness, ozone, water vapour, altitude |

Full argument documentation lives at `?SPART` -- this page focuses on how
the pieces fit together, not an exhaustive parameter list.

```{r lut}
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsSPART, nLUT = 5, setseed = 1))
# inputsSPART only ships PROSPECT-PRO/-D columns -- add what other leaf
# models need too (same requirement as foursail()/SPART() itself).
LUT$Cs <- 0; LUT$fqe <- 0.01; LUT$Cx <- 0
LUT$cell.d <- 40; LUT$inter.c <- 0.045; LUT$baseline.abs <- 0.0006
LUT$leaf.thick <- 1.6; LUT$albino.abs <- 0; LUT$lign.cell <- 2; LUT$Nitrogen <- 1
# A realistic, sea-level, clear-sky atmosphere for the worked example below
# (getLUT()'s own default sampling range for the atmosphere columns is wide
# enough to occasionally draw physically implausible combinations -- see
# the note at the end of Section 3.2).
LUT$Pa <- 1000; LUT$aot550 <- 0.3246; LUT$uo3 <- 0.3480; LUT$uh2o <- 1.4116
LUT$alt_m <- 0; LUT$Pa0 <- 1000
# BSM soil, at its own shipped defaults (Section 3.3 varies these on purpose)
LUT$BSMBrightness <- 0.5; LUT$BSMlat <- 25; LUT$BSMlon <- 45; LUT$SMp <- 15
```

## 3.2 Run a baseline simulation

```{r baseline}
sim <- suppressWarnings(SPART(inputLUT = LUT[1, ], CanopyModel = "fourSAIL",
                               LeafModel = "PROSPECT-PRO",
                               sensor.i = ToolsRTM::Sentinel2A.MSI,
                               rsoil = NULL, get.plots = FALSE))
names(sim$output)  # wave, rad.toa, rfl.toa, rfl.toc, rfl.toc.BRDF -- already at Sentinel-2A's own bands
```

Unlike `foursail()`, SPART returns output already convolved to the chosen
sensor's bands (`sensor.i`) -- there is no separate "simulate native, then
convolve" step (Tutorial 07 covers convolution for `foursail()`/`inform()`
output, which stays at 1nm resolution).

```{r baseline-plot}
plot(sim$output$wave, sim$output$rfl.toc.BRDF, type = "o", pch = 19, col = "#0072B2",
     ylim = c(0, max(sim$output$rfl.toc.BRDF, sim$output$rfl.toa)),
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "SPART: TOC vs. TOA (Sentinel-2A bands)")
lines(sim$output$wave, sim$output$rfl.toa, type = "o", pch = 19, col = "#D55E00")
legend("topright", c("TOC (canopy)", "TOA (satellite)"), col = c("#0072B2", "#D55E00"), pch = 19, lty = 1)
```

```{r toa-toc-diff}
diff_atm <- sim$output$rfl.toa - sim$output$rfl.toc.BRDF
plot(sim$output$wave, diff_atm, type = "o", pch = 19, col = "#CC79A7",
     xlab = "Wavelength (nm)", ylab = expression(Delta*R[atm] == R[TOA] - R[TOC]),
     main = "Atmospheric contribution: TOA minus TOC")
abline(h = 0, lty = 2, col = "grey50")
```

The dip near 942nm and the near-total collapse near 1372nm are real
atmospheric water-vapour absorption bands -- present in the TOA curve but
absent from TOC, exactly what real atmospheric correction has to deal
with; the diff plot above makes this quantitative: the atmosphere *adds*
reflectance in the visible (positive $\Delta R_{atm}$, mostly Rayleigh/
aerosol path radiance) and *removes* it in the water-vapour/near-infrared
bands (negative $\Delta R_{atm}$, absorption), rather than shifting the
whole spectrum uniformly. **A note on `getLUT()`'s atmosphere sampling**:
at the wide default ranges `inputsSPART` samples from, some random rows
combine a high `aot550` (aerosol loading) with a high `alt_m` (altitude)
-- a combination SMAC's atmospheric correction doesn't handle gracefully,
producing negative (physically meaningless) TOA reflectance. Always
sanity-check `range(sim$output$rfl.toa)` after a random draw; the fixed,
realistic sea-level clear-sky atmosphere set in Section 3.1 above avoids
this for the rest of this page.

## 3.3 Soil contribution

Physically, a sparse canopy should let more soil show through than a
dense one. **A real implementation gap, found and fixed while building
this page**: `inputsSPART`/`?SPART` used to document a `psoil` column,
but `SPART()`'s BSM soil path (triggered whenever `rsoil = NULL`) never
read it -- and it couldn't have, because `psoil` isn't actually a BSM
parameter at all. `psoil` is a PROSAIL/4SAIL soil-brightness-*mixing*
parameter (interpolating between a fixed dry and a fixed wet reference
spectrum); BSM (Brightness-Shape-Moisture) is a different, more
mechanistic soil model with its own four real parameters --
`BSMBrightness`, `BSMlat`, `BSMlon` (spectral shape), and `SMp` (soil
moisture volume %). `SPART()` used to hardcode all four internally
(`BSMBrightness = 0.5, BSMlat = 25, BSMlon = 45, SMp = 15`) regardless of
what was in the LUT. Both are now fixed: `SPART()` reads these four
columns from `inputLUT` when present (falling back to the same defaults
otherwise, so old code is unaffected), and `inputsSPART` now ships all
four as real, sampleable parameters.

### A sensitivity test showing each BSM parameter's real effect

```{r bsm-sensitivity}
run1 <- function(row) suppressWarnings(SPART(inputLUT = row, CanopyModel = "fourSAIL", LeafModel = "PROSPECT-PRO",
                                              sensor.i = ToolsRTM::Sentinel2A.MSI, rsoil = NULL, get.plots = FALSE))
base_row <- LUT[1, ]; base_row$LAI <- 1  # sparse-ish canopy so soil shows through
bsm_params <- list(BSMBrightness = c(0.2, 0.5, 0.8), BSMlat = c(20, 25, 40),
                    BSMlon = c(45, 45, 65), SMp = c(5, 15, 55))
cat("NIR band (789nm) TOC reflectance as each BSM parameter varies (others at their default):\n")
for (p in names(bsm_params)) {
  vals <- bsm_params[[p]]
  outs <- sapply(vals, function(v) { r <- base_row; r[[p]] <- v; run1(r)$output$rfl.toc.BRDF[8] })
  cat(sprintf("  %-14s %s -> %s\n", p, paste(vals, collapse = " / "), paste(round(outs, 4), collapse = " / ")))
}
```

Each of the four now has a real, verifiable effect -- brighter soil
(`BSMBrightness`) and drier soil (lower `SMp`) both raise reflectance,
as expected; `BSMlat`/`BSMlon` reshape the soil spectrum's spectral
*shape* rather than its overall level, so their effect at a single band
is smaller but still real.

### Flat spectra as a controlled diagnostic, BSM as the realistic soil demo

Two flat, arbitrary reflectance spectra are still useful for isolating
*just* the LAI effect (no confound from a soil spectrum's own spectral
shape) -- but they're a diagnostic, not the main soil story. The BSM
sensitivity above is the realistic one; `ToolsRTM::get.marmit.rsoil()`
(Tutorial 16) is the realistic one for *moisture*-driven soil spectra
specifically.

```{r soil-lai-gradient}
soil_dark   <- rep(0.08, 2001)  # dark/wet-looking soil (diagnostic only)
soil_bright <- rep(0.30, 2001)  # bright/dry-looking soil (diagnostic only)

lai_seq <- c(0.01, 0.5, 1, 2, 3, 4, 6)  # 0.01 stands in for LAI=0 -- foursail's
                                          # internal spline needs at least one point
soil_effect_bands <- function(lai) {
  row_i <- base_row; row_i$LAI <- lai
  d <- suppressWarnings(SPART(inputLUT = row_i, CanopyModel = "fourSAIL", LeafModel = "PROSPECT-PRO",
                               sensor.i = ToolsRTM::Sentinel2A.MSI, rsoil = soil_dark, get.plots = FALSE))
  b <- suppressWarnings(SPART(inputLUT = row_i, CanopyModel = "fourSAIL", LeafModel = "PROSPECT-PRO",
                               sensor.i = ToolsRTM::Sentinel2A.MSI, rsoil = soil_bright, get.plots = FALSE))
  b$output$rfl.toc.BRDF - d$output$rfl.toc.BRDF
}
soil_mat <- sapply(lai_seq, soil_effect_bands)  # bands x LAI
rownames(soil_mat) <- sim$output$wave
colnames(soil_mat) <- paste0("LAI=", lai_seq)
knitr::kable(round(soil_mat, 3))
```

```{r soil-lai-plot, fig.width=8}
matplot(sim$output$wave, soil_mat, type = "o", pch = 19, lty = 1,
        col = colorRampPalette(c("#D55E00", "#0072B2"))(length(lai_seq)),
        xlab = "Wavelength (nm)", ylab = "Bright-minus-dark TOC reflectance",
        main = "Soil brightness contribution vs. LAI, band by band")
legend("topright", colnames(soil_mat), col = colorRampPalette(c("#D55E00", "#0072B2"))(length(lai_seq)),
       lty = 1, pch = 19, cex = 0.7, ncol = 2)
```

Band-by-band, not just averaged: soil brightness matters most in the
visible and NIR (bands 1-9, up to ~871nm) at low LAI, and the whole
curve collapses toward zero as LAI rises from 0.01 to 6 -- the canopy
closing progressively removes the soil's spectral signature rather than
uniformly damping it.

## 3.4 Atmospheric contribution, independently by driver

Section 3.2's TOA-TOC diff plot showed the atmosphere's *total* effect.
Which atmospheric variable drives that, and where in the spectrum,
varies a lot:

```{r atmos-drivers}
atmos_vars <- list(aot550 = c(0.05, 0.3246, 1.0), uh2o = c(0.5, 1.4116, 4.0), uo3 = c(0.1, 0.3480, 1.0))
for (v in names(atmos_vars)) {
  vals <- atmos_vars[[v]]
  toa <- sapply(vals, function(x) { r <- LUT[1, ]; r[[v]] <- x; run1(r)$output$rfl.toa })
  rownames(toa) <- sim$output$wave
  colnames(toa) <- paste0(v, "=", vals)
  cat("\n", v, "-- TOA reflectance by band:\n")
  print(round(toa, 3))
}
```

`aot550` (aerosol loading) shifts mainly the visible bands (445-654nm,
rising with more aerosol -- more path radiance/scattering); `uh2o`
(water vapour) is almost invisible outside the two water-absorption
bands (942nm, 1372nm) where it dominates completely; `uo3` (ozone)'s
effect is small and concentrated around 560nm (the Chappuis band).
Different gases, different spectral fingerprints -- a single
"atmosphere on/off" comparison (Section 3.2) can't show this, only a
per-driver sweep can.

## 3.5 Geometry

`tts` (sun zenith), `tto` (view zenith), and `psi` (relative azimuth)
all feed into fourSAIL's BRDF term inside SPART, the same way they do in
`Compute_BRF()` (Tutorial 01) -- this is what makes reflectance
direction-dependent rather than a single fixed number per surface.
`inputsSPART` fixes `tts = 0` by default (sun straight overhead), which
by simple geometric symmetry makes `psi` irrelevant (there's no
"relative" direction to a sun with no azimuth of its own) -- so this
section fixes `tts = 30` instead, specifically to let `tto` and `psi`
show a real effect:

```{r geometry}
geom_row <- LUT[1, ]; geom_row$tts <- 30

cat("psi sweep at tto=20 deg (NIR, band 8 = 789nm):\n")
for (v in c(0, 90, 180)) {
  r <- geom_row; r$tto <- 20; r$psi <- v
  cat("  psi=", v, "->", round(run1(r)$output$rfl.toc.BRDF[8], 4), "\n")
}
cat("tto sweep at psi=0 (principal plane, toward the hotspot; NIR, band 8):\n")
for (v in c(0, 15, 30)) {
  r <- geom_row; r$tto <- v; r$psi <- 0
  cat("  tto=", v, "->", round(run1(r)$output$rfl.toc.BRDF[8], 4), "\n")
}
```

Both are real BRDF effects: moving `psi` away from 0 deg (away from the
sun's own azimuth, i.e. away from the backscatter/hotspot direction)
*lowers* NIR reflectance here; increasing `tto` toward `tts` (staying in
the backscatter direction, `psi=0`) *raises* it, approaching the
hotspot. **This specific direction (higher `tto` -> higher reflectance)
is a property of this geometry configuration** (backscatter side,
`psi=0`, this LAI/LIDF), not a general rule -- on the forward-scatter
side (`psi` near 180) the same `tto` increase can lower reflectance
instead. A Lambertian (direction-independent) assumption would miss all
of this.

## 3.6 A small multi-factor sensitivity comparison

Which of these actually moves the TOA signal the most? One factor at a
time, same NIR band, same low/mid/high bracketing used throughout this
page -- LAI and canopy/soil trait changes side by side with atmospheric
ones, extending Tutorial 10's spectral sensitivity analysis to the
full TOA chain:

```{r multi-factor}
factors <- list(LAI = c(0.5, 3, 6), Cab = c(10, 40, 75), BSMBrightness = c(0.2, 0.5, 0.8),
                aot550 = c(0.05, 0.3246, 1.0), uh2o = c(0.5, 1.4116, 4.0))
sens_tbl <- do.call(rbind, lapply(names(factors), function(fac) {
  vals <- factors[[fac]]
  outs <- sapply(vals, function(v) { r <- LUT[1, ]; r[[fac]] <- v; run1(r)$output$rfl.toa[8] })
  data.frame(factor = fac, low = outs[1], mid = outs[2], high = outs[3], range = max(outs) - min(outs))
}))
knitr::kable(sens_tbl[order(-sens_tbl$range), ], row.names = FALSE, digits = 4)
```

LAI dominates NIR TOA sensitivity here, but `aot550` (aerosol) is not
far behind -- comparable to, or larger than, moderate soil-brightness or
water-vapour changes. **`Cab` shows essentially zero NIR sensitivity**,
which is exactly right physically: chlorophyll absorbs strongly in the
visible, not the NIR, where structure (LAI, LIDF) dominates instead --
useful confirmation that the model's spectral behaviour matches known
leaf/canopy optics, not an oversight.

## 3.7 Why the TOC/TOA distinction matters for inversion

SPART slots into the same hybrid-inversion framework the rest of this
series builds (Tutorials 10-13) -- but which reflectance domain a model
is *trained* on matters enormously, because real satellite data is
always TOA, never TOC:

```{r domain-gap, warning=FALSE}
LUT150 <- as.data.frame(getLUT(inputs = ToolsRTM::inputsSPART, nLUT = 150, setseed = 1))
# Fix only atmosphere/BSM-soil (a realistic single-scene assumption, same as the
# rest of this page) -- leaf/canopy traits vary naturally across all 150 rows,
# same as a real inversion LUT (Tutorials 11-12). Fixing every non-Cab column
# (an earlier version of this chunk did) leaves only one varying predictor
# against 13 highly collinear bands -- a degenerate, overfit regression, not a
# realistic inversion.
for (col in c("Pa", "aot550", "uo3", "uh2o", "alt_m", "Pa0", "BSMBrightness", "BSMlat", "BSMlon", "SMp")) {
  LUT150[[col]] <- LUT[[col]][1]
}

sims <- lapply(seq_len(nrow(LUT150)), function(i) run1(LUT150[i, ]))
toc_mat <- t(sapply(sims, function(s) s$output$rfl.toc.BRDF))
toa_mat <- t(sapply(sims, function(s) s$output$rfl.toa))
colnames(toc_mat) <- colnames(toa_mat) <- paste0("B", seq_len(ncol(toc_mat)))

set.seed(1)
train_idx <- sample(seq_len(nrow(LUT150)), size = round(0.7 * nrow(LUT150)))
test_idx <- setdiff(seq_len(nrow(LUT150)), train_idx)
cab <- LUT150$Cab

fit_toc <- lm(cab[train_idx] ~ ., data = as.data.frame(toc_mat[train_idx, ]))
fit_toa <- lm(cab[train_idx] ~ ., data = as.data.frame(toa_mat[train_idx, ]))

r2 <- function(y, p) 1 - sum((y - p)^2) / sum((y - mean(y))^2)
rmse <- function(y, p) sqrt(mean((y - p)^2))
truth <- cab[test_idx]

pred_ideal <- predict(fit_toc, newdata = as.data.frame(toc_mat[test_idx, ]))     # TOC model, TOC data
pred_correct <- predict(fit_toa, newdata = as.data.frame(toa_mat[test_idx, ]))   # TOA model, TOA data
pred_domaingap <- predict(fit_toc, newdata = as.data.frame(toa_mat[test_idx, ])) # TOC model misapplied to TOA data

data.frame(
  scenario = c("TOC-trained model on TOC data (best case, no atmosphere)",
               "TOA-trained model on TOA data (correct real-world approach)",
               "TOC-trained model applied to TOA data (the domain-gap mistake)"),
  R2 = round(c(r2(truth, pred_ideal), r2(truth, pred_correct), r2(truth, pred_domaingap)), 3),
  RMSE = round(c(rmse(truth, pred_ideal), rmse(truth, pred_correct), rmse(truth, pred_domaingap)), 2)
) |> knitr::kable()
```

That's not a small effect: a Cab-inversion model trained on TOC
reflectance and (incorrectly) applied to TOA reflectance from a real
satellite collapses to a *negative* R² -- worse than just predicting the
mean every time -- while a model trained on TOA reflectance in the first
place performs about as well as the no-atmosphere ideal. This is the
concrete, load-bearing reason `SPART()` exists in this package: any LUT
built with plain `foursail()`/`inform()` for training an inversion model
that will later see real satellite reflectance needs either an
atmospheric-correction step on the real data, or -- what `SPART()`
provides directly -- a LUT simulated all the way to TOA in the first
place. Not developed further here (train/test mechanics are Tutorial
11's job) -- this section exists to make the *reason* concrete before
that tutorial's own worked example.

```text
Parameter LUT
     |
     v
SPART simulations (already at sensor bands, TOC or TOA)
     |
     v
Synthetic EO training dataset -- must match what you'll invert later
     |
     v
ML inversion
     |
     v
Vegetation traits
```

## 3.8 SPART and satellite observations

Because SPART already outputs at a chosen sensor's bands, moving from
simulation to a realistic EO-like observation needs no separate
convolution step -- just pick `sensor.i`. Only sensors SPART actually
supports (bundled objects in `ToolsRTM`) are listed here:

```{r sensors}
spart_sensors <- c("LANDSAT4.TM", "LANDSAT5.TM", "LANDSAT7.ETM", "LANDSAT8.OLI",
                    "Sentinel2A.MSI", "Sentinel2B.MSI", "Sentinel3A.OLCI",
                    "Sentinel3B.OLCI", "TerraAqua.MODIS")
s_modis <- suppressWarnings(SPART(inputLUT = LUT[1, ], CanopyModel = "fourSAIL", LeafModel = "PROSPECT-PRO",
                                   sensor.i = ToolsRTM::TerraAqua.MODIS, rsoil = NULL, get.plots = FALSE))
cat("Sentinel-2A bands:", length(sim$output$wave), " -- MODIS bands:", length(s_modis$output$wave), "\n")
```

## Take-home message

Use plain `foursail()`/`foursail2()`/`inform()` (Tutorials 01-02, 04) when
the question is about the canopy or leaf itself and TOC reflectance is
enough -- it's simpler, faster, and every model in this package other than
SPART works that way. Reach for `SPART()` specifically when the question
involves what a satellite actually measures at TOA (atmospheric
correction studies, comparing simulations directly against uncorrected
satellite products, or -- as Section 3.7 showed concretely -- training an
inversion model that will later see real, uncorrected satellite data) --
the extra atmosphere/soil-BSM machinery is not needed otherwise.
