---
title: "05. Building Look-Up Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{05. Building Look-Up Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
```

Every tutorial so far hand-wrote one row of trait values. A LUT (Look-Up
Table) is one row per simulation, many rows at once -- the input every
sensitivity analysis, LUT-matching inversion, and ML-training workflow in
this package starts from. Four functions build one, trading off control
against simplicity differently:

| Function | Distribution per trait | Trait correlation | Typical use |
|---|---|---|---|
| `getLUT()` | Fixed set (see `?getLUT`) | Independent | Quick default LUT, no customization needed |
| `get.LUTfromRanges()` | One distribution (`"uniform"`/`"gauss"`) for every trait, your own min/max | Independent | Custom ranges, same distribution shape for all traits |
| `get_distributionLUT()` | Per-trait `"Uniform"`/`"Gaussian"` choice | Car~Cab, optional (`DepCab`) | Different traits need different distribution shapes in the same LUT |
| `getCor()` | Normal or Uniform | Any two traits, any target correlation (`rho`) | General-purpose correlated sampling beyond Car~Cab |

## 1. `getLUT()`: the quick default

```{r getlut}
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 200, setseed = 1234))
# inputsPROSAIL's own LMA row is held fixed at 0 by design -- getLUT()
# samples Prot/CBC instead (the dry-matter inputs PROSPECT-PRO uses). This
# page later simulates with LeafModel="PROSPECT-D" (Section 4), which needs
# LMA directly, so sample it from a typical real-leaf range (0.005-0.02 g/cm2).
set.seed(1235)
LUT$LMA <- runif(200, 0.005, 0.02)
dim(LUT)
knitr::kable(head(LUT, 5))
```

## 2. `get_distributionLUT()`: a different distribution per trait

Real leaf/canopy traits don't all follow the same distribution shape.
Chlorophyll content (`Cab`) tends to cluster around a typical value
(Gaussian-like); the leaf-angle parameter `LIDFa` has no such tendency
across a mixed canopy (closer to Uniform). `get_distributionLUT()` lets
each trait pick its own shape instead of forcing one distribution on all
of them.

```{r ranges}
minval <- data.frame('N' = 1.5, 'Cab' = 5, 'Car' = 0, 'Anth' = 0, 'Cbrown' = 0.0,
                      'EWT' = 0.001, 'Prot' = 0.00001, 'CBC' = 0.00001,
                      'LIDFa' = 0, 'LAI' = 0.5,
                      LAIu = 0, sd = 200, cd = 0.2, h = 5,
                      fraction_brown = 0, diss = 0.1, Cv = 0.3, Zeta = 0)
maxval <- data.frame('N' = 3, 'Cab' = 70, 'Car' = 25, 'Anth' = 7, 'Cbrown' = 0.2,
                      'EWT' = 0.035, 'Prot' = 0.03, 'CBC' = 0.03,
                      'LIDFa' = 70, 'LAI' = 7,
                      LAIu = 0.8, sd = 1000, cd = 7, h = 20,
                      fraction_brown = 1, diss = 1, Cv = 1, Zeta = 0.2)
TypeDistrib <- data.frame('N' = 'Gaussian', 'Cab' = 'Gaussian', 'Car' = 'Gaussian',
                           'Anth' = 'Uniform', 'Cbrown' = 'Uniform',
                           'EWT' = 'Uniform', 'Prot' = 'Uniform', 'CBC' = 'Uniform',
                           'LIDFa' = 'Uniform', 'LAI' = 'Gaussian',
                           LAIu = 'Uniform', sd = 'Uniform', cd = 'Uniform', h = 'Uniform',
                           fraction_brown = 'Uniform', diss = 'Uniform', Cv = 'Uniform', Zeta = 'Uniform')
Mean_gauss <- data.frame('N' = 2.2, 'Cab' = 45, 'Car' = 8, 'LAI' = 2.25)
std_gauss <- Mean_gauss / 2.0

data.LUT <- get_distributionLUT(minval = minval, maxval = maxval, nSamples = 500,
                                 TypeDistrib = TypeDistrib, Mean_gauss = Mean_gauss,
                                 Std_gauss = std_gauss, DepCab = FALSE, setseed = 246)
```

```{r distrib-hist, fig.width=9}
op <- par(mfrow = c(1, 2))
hist(data.LUT$Cab, breaks = 30, col = "#2E8B57", border = "white",
     main = "Cab (Gaussian, mean=45)", xlab = "Cab")
hist(data.LUT$LIDFa, breaks = 30, col = "#B2182B", border = "white",
     main = "LIDFa (Uniform)", xlab = "LIDFa")
par(op)
```

`Cab` (Gaussian) and `LIDFa` (Uniform) came from the exact same call above
-- the histograms show why picking the right distribution per trait
matters.

## 3. Trait correlation

Independently-sampled traits are the default -- but real traits co-vary
(e.g. `Car` tracks `Cab`; plants with more of one pigment tend to have
more of the other). Two ways to get correlated traits into a LUT:

### `DepCab`: the built-in Car~Cab correlation

```{r depcab}
data.LUT.dep <- get_distributionLUT(minval = minval, maxval = maxval, nSamples = 500,
                                     TypeDistrib = TypeDistrib, Mean_gauss = Mean_gauss,
                                     Std_gauss = std_gauss, DepCab = TRUE, setseed = 246)
cat("Car~Cab correlation, DepCab=FALSE:", round(cor(data.LUT$Cab, data.LUT$Car), 2), "\n")
cat("Car~Cab correlation, DepCab=TRUE: ", round(cor(data.LUT.dep$Cab, data.LUT.dep$Car), 2), "\n")
```

```{r depcab-plot, fig.width=9}
op <- par(mfrow = c(1, 2))
plot(data.LUT$Cab, data.LUT$Car, pch = 19, col = "#2166AC", cex = 0.6,
     xlab = "Cab", ylab = "Car", main = "DepCab = FALSE (independent)")
plot(data.LUT.dep$Cab, data.LUT.dep$Car, pch = 19, col = "#B2182B", cex = 0.6,
     xlab = "Cab", ylab = "Car", main = "DepCab = TRUE (correlated)")
par(op)
```

### `getCor()`: correlate any two traits, any target strength

```{r getcor}
cor_lut <- getCor(n_inputs = 2, nLUT = 300, distribution = "Normal", setseed = 1,
                   rho = 0.7, Varnames = c("Cab", "LAI"),
                   MinRange = c(10, 0.5), MaxRange = c(80, 7))
cat("Target rho: 0.7. Achieved correlation:",
    round(cor(cor_lut$LUT$Cab, cor_lut$LUT$LAI), 2), "\n")
```

```{r getcor-plot}
plot(cor_lut$LUT$Cab, cor_lut$LUT$LAI, pch = 19, col = "#2E8B57", cex = 0.6,
     xlab = "Cab", ylab = "LAI", main = "getCor(): Cab~LAI at rho=0.7")
```

`getCor()` returns a list: `$LUT` (the correlated trait table to actually
use) and `$Covarianza` (the covariance matrix behind it, for reference).
`distribution` accepts `"Normal"` or `"Uniform"` here -- note this is
capitalized differently from `get.LUTfromRanges()`'s own `"uniform"`/
`"gauss"`, a real inconsistency in the package worth knowing about rather
than guessing past.

## 4. From LUT rows to spectra

A LUT alone is just trait values -- combine it with the models from
Tutorials 01-04 to actually simulate:

```{r sim-from-lut}
rsoil <- rep(0.15, 2101)
sim1 <- foursail(inputLUT = LUT[1, ], rsoil = rsoil, LeafModel = "PROSPECT-D")
plot(400:2500, sim1$rsot, type = "l", col = "#0072B2",
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "One LUT row, simulated")
```

## What's next

- **Tutorial 06** -- simulating every row of a LUT efficiently with
  parallel processing, instead of a single row or a small loop.
- **Tutorial 11** -- using a LUT as the training data for hybrid trait
  inversion.
