---
title: "06. Large-Scale and Parallel RTM Simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06. Large-Scale and Parallel RTM Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(doParallel)
library(foreach)
```

Every simulation so far used a plain `sapply()`/loop over LUT rows -- fine
for hundreds of rows, slow for thousands. This page shows the
`doParallel`/`foreach` pattern this package's own course scripts use
(`Scripts/R/ForPROSAIL/1-GetSimulationsLUTs.R`) to distribute simulations
across CPU cores.

**This whole page is forward simulation, not inversion.** Every LUT row
below is a set of *known* leaf/canopy traits fed *into* `foursail()` to
produce a spectrum; nothing here goes the other way (spectrum -> trait).
Parallelizing that direction, at scale, is the only subject of this
tutorial -- inversion (traits recovered *from* spectra) is Tutorials
11-13's subject, using LUTs built the same way as here.

## 1. Build a LUT

```{r lut}
n.samples <- 300
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n.samples, setseed = 1234))
# inputsPROSAIL's own LMA row is held fixed at 0 by design -- getLUT()
# samples Prot/CBC instead (the dry-matter inputs PROSPECT-PRO uses). This
# page simulates with LeafModel="PROSPECT-D" below, which needs LMA
# directly, so sample it from a typical real-leaf range (0.005-0.02 g/cm2).
set.seed(1235)
LUT$LMA <- runif(n.samples, 0.005, 0.02)
rsoil <- rep(0.15, 2101)
```

### What's actually varying across these 300 rows

```{r lut-traits, fig.width=9, fig.height=6}
op <- par(mfrow = c(2, 2))
hist(LUT$Cab, breaks = 20, col = "#0072B2", main = "Cab (chlorophyll)", xlab = "ug/cm2")
hist(LUT$LAI, breaks = 20, col = "#009E73", main = "LAI", xlab = "m2/m2")
hist(LUT$EWT, breaks = 20, col = "#D55E00", main = "EWT (leaf water)", xlab = "cm")
hist(LUT$LMA, breaks = 20, col = "#CC79A7", main = "LMA (dry matter)", xlab = "g/cm2")
par(op)
```

This is the actual sampled input space `foursail()` runs over below --
not just a row count. `getLUT()` draws each trait independently from
`inputsPROSAIL`'s own documented ranges/distributions (Tutorial 01).

## 2. Sequential baseline

```{r sequential}
t_seq <- system.time({
  sims_seq <- lapply(seq_len(n.samples), function(i) {
    sail_i <- foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D")
    sail_i$rsot
  })
})
cat("Sequential:", round(t_seq[["elapsed"]], 2), "s for", n.samples, "simulations\n")
```

## 3. Parallel with `doParallel`/`foreach`

```{r parallel}
no_cores <- max(1, parallel::detectCores() - 2)
# CRAN check machines cap parallel workers at 2; respect that limit so this
# vignette rebuilds cleanly under R CMD check --as-cran.
if (nzchar(Sys.getenv("_R_CHECK_LIMIT_CORES_"))) no_cores <- min(no_cores, 2L)
cl <- makeCluster(no_cores)
registerDoParallel(cl)

t_par <- system.time({
  sims_par <- foreach(i = 1:n.samples, .packages = "ToolsRTM") %dopar% {
    sail_i <- foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D")
    sail_i$rsot
  }
})
stopCluster(cl)

cat("Parallel (", no_cores, "cores):", round(t_par[["elapsed"]], 2), "s for", n.samples, "simulations\n")
cat("Speedup:", round(t_seq[["elapsed"]] / t_par[["elapsed"]], 2), "x\n")
```

Three things matter for this pattern to actually work, not just look
parallel:

- **`makeCluster()`/`registerDoParallel()`/`stopCluster()`** always paired
  -- an unclosed cluster leaks worker processes across the whole R
  session, compounding if this chunk runs more than once (e.g. re-knitting
  interactively).
- **`.packages = "ToolsRTM"`** inside `foreach()` -- each worker process
  starts fresh, with no packages attached; without this, `foursail()`
  isn't found on the workers and every iteration errors.
- **`no_cores <- detectCores() - 2`**, not `detectCores()` itself --
  leaves headroom for the OS and this R session's own main process, a
  convention this package's own course scripts (`ForPROSAIL`,
  `Sensibility`) use throughout.

```{r verify}
identical(sims_seq[[1]], sims_par[[1]])  # same result, sequential or parallel
```

### The simulations themselves

```{r sim-spectra-plot, fig.width=8}
wave <- 400:2500
spectra_mat <- do.call(rbind, sims_par)
matplot(wave, t(spectra_mat[sample(seq_len(n.samples), 40), ]), type = "l", lty = 1,
        col = adjustcolor("#0072B2", alpha.f = 0.3),
        xlab = "Wavelength (nm)", ylab = "TOC reflectance (rsot)",
        main = "40 of the 300 parallel-simulated spectra")
lines(wave, colMeans(spectra_mat), col = "#D55E00", lwd = 2)
legend("topleft", c("individual simulations", "mean across all 300"), col = c("#0072B2", "#D55E00"), lty = 1, lwd = c(1, 2))
```

This is what the parallel run in Section 3 actually produced -- the
spread across the 300 curves is a direct consequence of the trait
histograms above, not noise: every curve is one specific, known
`(Cab, LAI, EWT, LMA, ...)` combination run forward through `foursail()`.

## 4. When parallel is (and isn't) worth it

```{r speedup-note}
cat("Per-simulation cost: sequential", round(1000 * t_seq[["elapsed"]] / n.samples, 2),
    "ms, parallel", round(1000 * t_par[["elapsed"]] / n.samples, 2), "ms\n")
```

`foursail()` itself is fast (milliseconds per call) -- the parallel
speedup above has to overcome the fixed overhead of starting a cluster and
shipping `ToolsRTM`/data to every worker, so it only pays off once
`n.samples` is large enough (thousands, not hundreds) or each simulation
is individually expensive (e.g. `SPART()`'s atmosphere calculations,
Tutorial 03; the full SCOPE energy balance, covered in the SCOPEinR
package's own tutorials). For a few hundred rows, the sequential
`lapply()`/`sapply()` used throughout Tutorials 01-05 is often simpler and
not meaningfully slower once cluster startup is accounted for.

## 5. Filtering non-finite rows

Large parallel runs sometimes include a few unstable simulations (e.g.
Liberty's leaf solver, Tutorial 02, near the edges of its own parameter
range) that come back as `NaN` rather than erroring outright. Always
check before using the results downstream:

```{r filter}
spectra_mat <- do.call(rbind, sims_par)
ok <- apply(spectra_mat, 1, function(r) all(is.finite(r)))
cat(sum(!ok), "of", n.samples, "simulations were non-finite and would be dropped.\n")
spectra_mat <- spectra_mat[ok, , drop = FALSE]
LUT_ok <- LUT[ok, ]
```

## What's next

- **Tutorial 07** -- convolving these simulated spectra onto real sensor
  bands.
- **Tutorial 14** -- the same simulate-in-parallel pattern as one step in
  a full end-to-end pipeline.
