---
title: "11. From Physics to Vegetation Traits: Hybrid Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{11. From Physics to Vegetation Traits: Hybrid Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

Every tutorial so far ran the forward model: traits in, spectrum out.
**Inversion** goes the other way -- given ONLY the sensor-band reflectance
a satellite or field spectrometer would measure, retrieve a biophysical
trait (chlorophyll content, `Cab`, throughout this page) without knowing
the ground truth. This is the **hybrid** approach: instead of solving the
radiative transfer equations analytically (rarely possible) or field-
calibrating an empirical index-to-trait relationship (needs field data
per site/crop), simulate a LUT with the physics you already have, and
learn the reflectance-to-trait relationship from the simulation itself.

```text
Parameter LUT
     |
     v
RTM simulations (Tutorials 01-06)
     |
     v
Sensor convolution (Tutorial 07)
     |
     v
Synthetic training dataset (reflectance, trait) pairs
     |
     v
Inversion method
     |
     v
Vegetation traits
```

Three inversion methods share this exact framework, differing only in the
last step:

- **`get.inversionOpt()`** (this page): no model is fit at all. It ranks
  every spectrum in a reference LUT by how closely it matches an observed
  spectrum under a chosen merit function, then averages the trait values
  of the closest matches. Pure library search.
- **`get.inversion()`** (Tutorial 12): fits a statistical/ML model
  (Random Forest, PLSR, SVM, ... 12 algorithms via `caret`) on a training
  LUT, then predicts on new spectra.
- **`getMLmodel()`** (Tutorial 13): fits a deep-learning model on the
  same kind of data.

## 1. Simulate a LUT and convolve to Sentinel-2A

```{r simulate}
n_samples <- 700
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))

wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))

refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
wl_bands <- as.numeric(names(se2a)[-1])
band_names <- paste0("B", seq_along(wl_bands))
names(se2a) <- c("id", band_names)
```

700 simulated spectra, convolved to Sentinel-2A's 13 real bands -- this is
what every inversion method below actually sees, not the native 1nm
spectrum.

## 2. A train/test split, shared by every method

```{r split}
set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
test_idx  <- setdiff(seq_len(n_samples), train_idx)

LUT_train <- LUT[train_idx, ]; LUT_test <- LUT[test_idx, ]
se2a_mat <- as.matrix(se2a[, band_names])
se2a_train_mat <- se2a_mat[train_idx, ]; se2a_test_mat <- se2a_mat[test_idx, ]

r2_f   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))

cat("Train:", length(train_idx), "spectra. Test (held out):", length(test_idx), "spectra.\n")
```

## 3. `get.inversionOpt()`: LUT merit-function matching

`method` picks how "closeness" between two spectra is measured. `nOpt`
controls how many of the closest training spectra get averaged together
for the final trait estimate:

```{r inversionopt-methods, results='hide'}
opt_rmse <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-RMSE", nOpt = 5)
opt_fge  <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-FGE", nOpt = 5)
opt_dwt  <- get.inversionOpt(rfl.sensor = se2a_test_mat, rfl.rtm = se2a_train_mat,
                              LUT = LUT_train, wave = wl_bands, method = "merit-DWT", nOpt = 5)
```

Each call returns a 2-element list: `[[1]]` (`rfl.b`) is the matched
reflectance itself, `[[2]]` (`LUT.best`) is the corresponding trait table:

```{r inversionopt-check}
opt_metrics <- data.frame(
  method = c("merit-RMSE", "merit-FGE", "merit-DWT"),
  R2   = c(r2_f(LUT_test$Cab, opt_rmse[[2]]$Cab), r2_f(LUT_test$Cab, opt_fge[[2]]$Cab), r2_f(LUT_test$Cab, opt_dwt[[2]]$Cab)),
  RMSE = c(rmse_f(LUT_test$Cab, opt_rmse[[2]]$Cab), rmse_f(LUT_test$Cab, opt_fge[[2]]$Cab), rmse_f(LUT_test$Cab, opt_dwt[[2]]$Cab))
)
knitr::kable(opt_metrics, digits = 3)
```

`merit-RMSE`/`merit-FGE` compare raw band reflectance; `merit-DWT`
compares each spectrum's discrete wavelet transform coefficients instead
(sensitive to overall spectral shape rather than each band's exact value).
`merit-NRMSE`, `merit-MAE`, `merit-NMB`, and `merit-1stD`
(first-derivative matching) are the remaining built-in options;
`custom_stat` accepts your own `function(sim, obs)`.

```{r inversionopt-plot}
plot(LUT_test$Cab, opt_rmse[[2]]$Cab, pch = 19, col = "#2166AC",
     xlab = "Observed Cab", ylab = "Predicted Cab (merit-RMSE, nOpt=5)",
     main = "LUT merit-function matching")
abline(0, 1, col = "grey40", lty = 2)
```

## Why this method, and when

`get.inversionOpt()` needs no training step -- a quick, physically-
grounded retrieval with no ML infrastructure. Its limitation: accuracy
is capped by how well the reference LUT covers the true trait space,
and it re-searches the whole reference LUT for every new observation (no
learned model to reuse). Tutorial 12 builds actual fitted models
instead, trading that simplicity for better accuracy at scale.

## 4. Is 700 samples actually enough?

A LUT-search method's accuracy is capped by how densely its reference
LUT covers the trait space; PROSAIL-hybrid-inversion literature
typically recommends LUT sizes in the thousands (often tens of
thousands for operational retrieval) for reliable coverage of even a
handful of free parameters. 700 rows -- 490 of them used as the search
reference after the train/test split -- is a real step up from a
bare-minimum demo size, but still short of that. To make the effect
concrete rather than just cite a number, here is `get.inversionOpt()`
run at 100 rows, at this page's own 700, and at a 2000-row reference:

```{r sample-size-check, results='hide'}
run_experiment <- function(n_samples, seed = 1) {
  LUT_n <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = seed))
  refl_n <- t(sapply(seq_len(n_samples), function(i) {
    foursail(inputLUT = LUT_n[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
  }))
  refl_X_n <- as.data.frame(refl_n); colnames(refl_X_n) <- paste0("X", wl)
  refl_X_n <- cbind(id = seq_len(nrow(refl_X_n)), refl_X_n)
  se2a_n <- suppressMessages(get.spectra.convolved(rfl = refl_X_n, sensor = "Sentinel2a", plot.spectra = FALSE))
  names(se2a_n) <- c("id", band_names)

  set.seed(seed)
  train_idx_n <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
  test_idx_n  <- setdiff(seq_len(n_samples), train_idx_n)
  se2a_n_mat <- as.matrix(se2a_n[, band_names])

  opt_n <- get.inversionOpt(rfl.sensor = se2a_n_mat[test_idx_n, ], rfl.rtm = se2a_n_mat[train_idx_n, ],
                             LUT = LUT_n[train_idx_n, ], wave = wl_bands, method = "merit-RMSE", nOpt = 5)
  data.frame(n = n_samples, n_train = length(train_idx_n),
             R2 = r2_f(LUT_n[test_idx_n, ]$Cab, opt_n[[2]]$Cab),
             RMSE = rmse_f(LUT_n[test_idx_n, ]$Cab, opt_n[[2]]$Cab))
}
sample_size_results <- rbind(run_experiment(100), run_experiment(700), run_experiment(2000))
```

```{r sample-size-table}
knitr::kable(sample_size_results, digits = 3, row.names = FALSE)
```

At 100 samples, R² is negative -- the method has no real skill, worse
than just guessing the mean. Jumping to 700 -- this page's own LUT size
-- turns that into real, usable skill (R² = 0.573): the biggest single
gain sits in that first jump, not in going further. Continuing to 2000
still helps (R² = 0.648, lower RMSE) but by a much smaller margin than
100-to-700 did -- diminishing returns have clearly set in by 700 rows,
even though 700 remains short of the "thousands" literature
recommendation. **700 rows is a reasonable balance for a tutorial that
still needs to build in a reasonable time: enough to demonstrate a
real, usable hybrid-inversion result, not just the theory.** For actual
applications, especially with more than one free parameter to retrieve
simultaneously, still build a LUT of at least a few thousand rows
(Tutorial 06 covers parallelizing exactly that).

## What's next

- **Tutorial 12** -- `get.inversion()`: 12 ML algorithms, compared on this
  same LUT.
- **Tutorial 13** -- `getMLmodel()`: deep learning on the same data.
- **Tutorial 14** -- the full pipeline end-to-end, LUT to trait maps.
