---
title: "12. Comparing ML Algorithms for RTM Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{12. Comparing ML Algorithms for RTM Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

Tutorial 11 covered `get.inversionOpt()` -- pure LUT-search, no model
fit at all. This page introduces `get.inversion()` for the first time,
and compares several of its 12 supported algorithms (via `caret`) on the
same LUT/train-test split, so the comparison is apples-to-apples.

## 1. Same LUT, same split as Tutorial 11

```{r simulate}
n_samples <- 150
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
names(se2a) <- c("id", band_names)

set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
test_idx  <- setdiff(seq_len(n_samples), train_idx)
train_df <- cbind(LUT[train_idx, ], se2a[train_idx, band_names])
test_df  <- cbind(LUT[test_idx, ],  se2a[test_idx,  band_names])

r2_f   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))
```

```{r sim-plot, fig.width=8}
matplot(wl, t(refl[sample(seq_len(n_samples), 30), ]), type = "l", lty = 1,
        col = adjustcolor("#0072B2", alpha.f = 0.3),
        xlab = "Wavelength (nm)", ylab = "TOC reflectance (rsot)",
        main = "30 of the 150 simulated spectra behind this LUT")
```

## 2. Fit several algorithms, predict Cab on the held-out test set

`get.inversion()` supports 12 algorithms: `"PLSR"`, `"SVM"`, `"RF"`,
`"GB"`, `"NN"`, `"Bayesian"`, `"AdaBag"`, `"BRNN"`, `"xGB"`, `"RVM"`,
`"qLASSO"`, `"Ensemble"`. `"NN"` (caret's `nnet` tuning grid) is skipped
here -- impractically slow against this many predictors at only 105
training rows, the same reason this package's own course pipeline scripts
(`Scripts/R/*/2-inversion_ML.R`) skip it too. A representative subset run
for real below; the rest of the call is identical for any of the others.

**`"Ensemble"` (stacks SVM + Gradient Boosting + a neural net via
`caretEnsemble`) is left out of the comparison below on purpose**: while
verifying this page, it surfaced two real bugs in `get.inversion()`'s
source (both fixed directly in the package for this release --
`fmla.n` was referenced without being defined in that branch, and its
internal `stackControl` passed the whole training data.frame to
`caret::createFolds()` instead of the response column, misaligning fold
sizes) -- but even after both fixes, `caretEnsemble::caretStack()` still
fails with `"pred_rows == pred_rows[1L] are not all TRUE"`, a deeper
row-alignment mismatch across the three stacked sub-models' predictions
that needs further investigation into how their tuning grids interact,
not something safe to guess-fix here.

```{r fit-algorithms, results='hide'}
algorithms <- c("PLSR", "SVM", "RF", "GB")
fits <- lapply(algorithms, function(algo) {
  get.inversion(data = train_df, depVar = "Cab", inputs = band_names,
                algorithm = algo, n.samples = nrow(train_df), seed = 42)
})
names(fits) <- algorithms
```

```{r evaluate}
metrics <- do.call(rbind, lapply(algorithms, function(algo) {
  pred <- as.numeric(predict(fits[[algo]]$model, newdata = test_df[, c("Cab", band_names)]))
  data.frame(algorithm = algo, R2 = r2_f(test_df$Cab, pred), RMSE = rmse_f(test_df$Cab, pred))
}))
knitr::kable(metrics[order(-metrics$R2), ], digits = 3, row.names = FALSE)
```

```{r best-plot}
best_algo <- metrics$algorithm[which.max(metrics$R2)]
pred_best <- as.numeric(predict(fits[[best_algo]]$model, newdata = test_df[, c("Cab", band_names)]))
plot(test_df$Cab, pred_best, pch = 19, col = "#2166AC",
     xlab = "Observed Cab", ylab = paste("Predicted Cab (", best_algo, ")"),
     main = paste("Best algorithm on this LUT:", best_algo))
abline(0, 1, col = "grey40", lty = 2)
```

### All four algorithms, not just the winner

A single "best" scatter plot hides how the others actually did. Every
algorithm's predicted-vs-observed, side by side, plus the R2/RMSE bars
from Section 2's table:

```{r all-algos-plot, fig.width=9, fig.height=7}
op <- par(mfrow = c(2, 2))
for (algo in algorithms) {
  pred <- as.numeric(predict(fits[[algo]]$model, newdata = test_df[, c("Cab", band_names)]))
  r2_here <- round(r2_f(test_df$Cab, pred), 3)
  plot(test_df$Cab, pred, pch = 19, col = "#2166AC",
       xlab = "Observed Cab", ylab = "Predicted Cab",
       main = sprintf("%s (R2=%.3f)", algo, r2_here))
  abline(0, 1, col = "grey40", lty = 2)
}
par(op)
```

```{r r2-rmse-bars, fig.width=8}
op <- par(mfrow = c(1, 2))
barplot(setNames(metrics$R2, metrics$algorithm), col = "#0072B2", ylab = "R2", main = "R2 by algorithm")
barplot(setNames(metrics$RMSE, metrics$algorithm), col = "#D55E00", ylab = "RMSE", main = "RMSE by algorithm")
par(op)
```

## 3. Comparing against Tutorial 11's merit-function matching

```{r vs-opt, results='hide'}
se2a_mat <- as.matrix(se2a[, band_names])
opt_rmse <- get.inversionOpt(rfl.sensor = se2a_mat[test_idx, ], rfl.rtm = se2a_mat[train_idx, ],
                              LUT = LUT[train_idx, ], wave = as.numeric(sub("B", "", band_names)),
                              method = "merit-RMSE", nOpt = 5)
```

```{r vs-opt-table}
comparison <- rbind(metrics[, c("algorithm", "R2", "RMSE")],
                     data.frame(algorithm = "inversionOpt (merit-RMSE)",
                                R2 = r2_f(LUT[test_idx, ]$Cab, opt_rmse[[2]]$Cab),
                                RMSE = rmse_f(LUT[test_idx, ]$Cab, opt_rmse[[2]]$Cab)))
knitr::kable(comparison[order(-comparison$R2), ], digits = 3, row.names = FALSE)
```

At this LUT size (150 rows, 105 training), a fitted ML model and pure
LUT-search inversion are often close -- the gap widens in the ML model's
favour with more training data, and in `get.inversionOpt()`'s favour when
too little data is available to fit a model reliably (Tutorial 11's
discussion of when to use which).

## 4. Multiple traits: do spectral indices actually help?

Every comparison so far predicted one trait (Cab) from raw bands
alone. Real applications need several traits, and raw reflectance
bands are not the only useful input -- vegetation indices (Tutorial
09) can carry trait-specific signal that a generic ML algorithm has to
work harder to reconstruct from bands on its own. This section checks
that directly, on the same LUT and train/test split, across six
traits: Random Forest fit twice per trait -- bands only, and bands
plus a small, *per-trait* selection of the indices most correlated
with that specific trait (not one fixed index set reused everywhere).

```{r multi-trait-setup}
# Real Sentinel-2A band names (not the generic B1..B10 from Section 1) --
# getIndicesSE2.ML()'s index formulas need to find specific bands like
# B8A/B11/B12 by name.
se2a_full <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_full) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
bands_real <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")  # skip the atmospheric-only bands (B1, B9, B10)

idx_ml <- suppressMessages(getIndicesSE2.ML(df = se2a_full[, -1], sensor = "Sentinel-2a", df.data = NULL, fast.process = TRUE))
cat("Candidate indices:", ncol(idx_ml), "\n")
```

```{r multi-trait-select}
traits <- c("Cab", "Car", "Anth", "LAI", "EWT", "Cbrown")
n_top <- 5  # indices kept per trait

# Correlation against training rows only, per trait -- an index useful for
# EWT is not necessarily useful for Anth, so this is computed separately
# for each trait rather than picking one index set for all of them.
select_indices <- function(trait) {
  cors <- sapply(names(idx_ml), function(nm) suppressWarnings(cor(idx_ml[train_idx, nm], LUT[train_idx, trait])))
  cors <- cors[is.finite(cors)]
  names(sort(abs(cors), decreasing = TRUE))[seq_len(min(n_top, length(cors)))]
}
selected <- setNames(lapply(traits, select_indices), traits)
knitr::kable(data.frame(trait = traits, top_indices = sapply(selected, paste, collapse = ", ")), row.names = FALSE)
```

```{r multi-trait-fit, results='hide', fig.show='hide'}
run_pair <- function(trait) {
  idx_cols <- selected[[trait]]
  df_bands <- cbind(LUT[trait], se2a_full[bands_real])
  df_full  <- cbind(df_bands, idx_ml[idx_cols])

  fit_bands <- get.inversion(data = df_bands[train_idx, ], depVar = trait, inputs = bands_real,
                              algorithm = "RF", n.samples = length(train_idx), seed = 42)
  fit_full  <- get.inversion(data = df_full[train_idx, ], depVar = trait, inputs = c(bands_real, idx_cols),
                              algorithm = "RF", n.samples = length(train_idx), seed = 42)

  pred_bands <- as.numeric(predict(fit_bands$model, newdata = df_bands[test_idx, c(trait, bands_real)]))
  pred_full  <- as.numeric(predict(fit_full$model,  newdata = df_full[test_idx,  c(trait, bands_real, idx_cols)]))

  data.frame(trait = trait,
             R2_bands_only = r2_f(LUT[test_idx, trait], pred_bands),
             R2_bands_plus_indices = r2_f(LUT[test_idx, trait], pred_full))
}
trait_results <- do.call(rbind, lapply(traits, run_pair))
trait_results$delta <- trait_results$R2_bands_plus_indices - trait_results$R2_bands_only
```

```{r multi-trait-table}
knitr::kable(trait_results[order(-trait_results$delta), ], digits = 3, row.names = FALSE)
```

```{r multi-trait-plot, fig.width=8}
ord <- order(trait_results$delta)
barplot(rbind(trait_results$R2_bands_only[ord], trait_results$R2_bands_plus_indices[ord]),
        beside = TRUE, names.arg = trait_results$trait[ord],
        col = c("#999999", "#0072B2"), ylab = "R2 (independent test set)",
        main = "Bands only vs. bands + top-5 correlated indices, per trait")
legend("topleft", c("Bands only", "+ indices"), fill = c("#999999", "#0072B2"), bty = "n")
```

**Adding indices helps every single trait here, but by very different
amounts.** `EWT` and `Cbrown` gain the most -- unsurprising, since
water- and senescence-sensitive indices (built from SWIR bands the raw
per-band RF split has to rediscover on its own) target exactly that
signal directly. `Cab`/`Car`/`LAI` gain a real but more modest amount --
RF was already extracting a fair fraction of the available structure
from the bands alone. `Anth` stays the weakest predictor of the six
even with indices added, consistent with anthocyanin's famously subtle
spectral footprint (Tutorial 09) -- indices help, but they don't turn a
physically hard retrieval into an easy one. The lesson generalises
past this specific LUT: *which* indices to add, computed and selected
per trait, matters more than adding indices in general.

## What's next

- **Tutorial 13** -- deep learning (`getMLmodel()`) on the same kind of
  data, and when it's worth the extra complexity over the algorithms here.
- **Tutorial 14** -- this whole simulate-convolve-invert chain as one
  coherent pipeline, across every canopy model this package supports.
