---
title: "Getting LUTs for SCOPE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting LUTs for SCOPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

SCOPE needs more input columns than a plain PROSAIL LUT -- leaf
biochemistry, canopy structure, soil/meteorology, and viewing geometry
all in one row (see `SCOPEinR_tutorial.Rmd`'s "Input structure" section
for the full field list). `getLUT.SCOPE()` builds that from a single
ranges table, and this page covers the two things you'll want beyond the
default call: which distribution each variable samples from, and how to
correlate two of them.

## 1. `inputs_SCOPE.csv`: one distribution per variable, already built in

Unlike `ToolsRTM::get_distributionLUT()` (a separate function taking
distribution as an explicit argument), SCOPE's ranges table has the
distribution choice baked into the CSV itself -- the `Distribution`
column, one of `"Uniform"`, `"Gaussian"`, or `"Fixed"`:

```{r inputs-table}
path_input <- system.file("input", package = "SCOPEinR")
inputLUT <- read.table(file.path(path_input, "inputs_SCOPE.csv"), header = TRUE, sep = ",")

inputLUT[inputLUT$variable %in% c("Cab", "LAI", "Vcmax25", "tts"),
         c("variable", "lower", "upper", "Distribution", "Mean_D", "Std_D", "default")]
```

`"Fixed"` rows (most of the meteorology -- `Ta`, `Rin`, `Rli`, ... --
and constants like `BallBerry0`) are held at their `default` value every
time, regardless of `lower`/`upper`. `"Uniform"` rows are sampled evenly
across `[lower, upper]`. `"Gaussian"` rows (`Cab` here) are sampled around
`Mean_D` with spread `Std_D`, truncated to stay within `[lower, upper]`.

```{r build-lut}
n_samples <- 200
LUT <- getLUT.SCOPE(inputLUT = inputLUT, nLUT = n_samples, setseed = 1)
```

```{r distrib-hist, fig.width=9}
op <- par(mfrow = c(1, 2))
hist(LUT$Cab, breaks = 25, col = "#2E8B57", border = "white",
     main = "Cab (Gaussian, mean=50)", xlab = "Cab")
hist(LUT$LAI, breaks = 25, col = "#B2182B", border = "white",
     main = "LAI (Uniform)", xlab = "LAI")
par(op)
```

## 2. Correlating two SCOPE traits

Real leaf traits co-vary -- e.g. plants with higher photosynthetic
capacity (`Vcmax25`) often also carry more chlorophyll (`Cab`).
`getLUT.SCOPE()` samples every variable independently; `ToolsRTM::getCor()`
(covered in depth in the `ToolsRTM` package's own `Getting-LUTs` article)
generates a correlated pair instead, which can then be spliced into the
LUT in place of its independent columns:

```{r correlate}
pigments <- ToolsRTM::getCor(n_inputs = 2, setseed = 7, distribution = "Uniform",
                              nLUT = n_samples, rho = 0.6, Varnames = c("Cab", "Vcmax25"),
                              MinRange = c(5, 0.75), MaxRange = c(90, 250))

LUT$Cab     <- pigments$LUT$Cab
LUT$Vcmax25 <- pigments$LUT$Vcmax25

cat("Target rho: 0.6. Achieved correlation:", round(cor(LUT$Cab, LUT$Vcmax25), 2), "\n")
```

```{r correlate-plot}
plot(LUT$Cab, LUT$Vcmax25, pch = 19, col = "#2166AC", cex = 0.6,
     xlab = "Cab", ylab = "Vcmax25", main = "Correlated Cab~Vcmax25 for a SCOPE LUT")
```

This `LUT` is now ready for `get.SCOPE()`/`get.SCOPE.parallel()` exactly
as built by the plain `getLUT.SCOPE()` call -- only the sampling of these
two columns changed. See `How-in-R-SCOPEinR.Rmd` (Section 4) for the same
`getCor()` pattern applied to `LIDFa`/`LIDFb` instead, and
`SCOPEinR_tutorial.Rmd` (Section 6) for running the resulting LUT through
`get.SCOPE()` at scale.
