---
title: "Trait inversion from SCOPE reflectance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trait inversion from SCOPE reflectance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
library(randomForest)
```

The `sensitivity-scope` article on this site found that `Vcmax25`
(photosynthetic capacity) has no direct radiative-transfer signature --
Fluspect/4SAIL don't take it as an input at all, so any reflectance
response is indirect, through the leaf temperature the energy balance
solves for. This page tests the practical consequence: **can you actually
retrieve `Vcmax25` from sensor-band reflectance, the way you can retrieve
`Cab` or `LAI`?**

## 1. Simulate a LUT and convolve to Sentinel-2A

```{r simulate, results='hide'}
path_input <- system.file("input", package = "SCOPEinR")
scope_options <- read.table(file.path(path_input, "setoptions.csv"), header = TRUE, sep = ",")
inputLUT <- read.table(file.path(path_input, "inputs_SCOPE.csv"), header = TRUE, sep = ",")

n_samples <- 500
set.seed(11)
LUT <- getLUT.SCOPE(inputLUT = inputLUT, nLUT = n_samples)

db_sims <- get.SCOPE(LUT = LUT, n.LUT = n_samples, options.SCOPE = scope_options,
                      optipar = SCOPEinR::optipar2021.Pro.CX, leaf.model = "fluspect-CX",
                      canopy.model = "fourSAIL", get.outputs = "ALL", get.plots = FALSE)

wl_optical <- 400:2400
n <- length(wl_optical)

band_refl <- t(sapply(db_sims, function(r) {
  rfl_i <- r$data.rad$reflapp[1:n]
  # reflapp is a radiance ratio -- genuinely near-zero incident irradiance
  # at a handful of water-vapor/O2 absorption wavelengths makes it
  # numerically unstable there in every simulation (see How-in-R-SCOPEinR.Rmd,
  # Section 5). Linearly interpolate over those narrow gaps before convolving.
  bad <- !is.finite(rfl_i)
  if (any(bad)) rfl_i[bad] <- approx(wl_optical[!bad], rfl_i[!bad], xout = wl_optical[bad])$y
  df_i <- data.frame(wave = wl_optical, rfl = rfl_i)
  get.spectral.convolution.srf(df_i, ToolsRTM::srf.sentinel2a)$RFL
}))
colnames(band_refl) <- paste0("B", seq_len(ncol(band_refl)))
```

## 2. Train/test split, shared across all three traits

```{r split}
set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
test_idx  <- setdiff(seq_len(n_samples), train_idx)

r2_f   <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))

invert_trait <- function(trait) {
  ml_data <- data.frame(band_refl, y = LUT[[trait]])
  rf <- randomForest(y ~ ., data = ml_data[train_idx, ], ntree = 300)
  pred <- predict(rf, ml_data[-train_idx, ])
  obs  <- ml_data$y[-train_idx]
  list(obs = obs, pred = pred, R2 = r2_f(obs, pred), RMSE = rmse_f(obs, pred))
}
```

## 3. Retrieve `Cab`, `LAI`, and `Vcmax25` -- same method, same data

```{r invert-all, results='hide'}
inv_cab    <- invert_trait("Cab")
inv_lai    <- invert_trait("LAI")
inv_vcmax  <- invert_trait("Vcmax25")
```

```{r invert-summary}
summary_df <- data.frame(
  trait = c("Cab", "LAI", "Vcmax25"),
  direct_RT_input = c("Yes (Fluspect)", "Yes (4SAIL)", "No (biochemistry only)"),
  R2   = c(inv_cab$R2, inv_lai$R2, inv_vcmax$R2),
  RMSE = c(inv_cab$RMSE, inv_lai$RMSE, inv_vcmax$RMSE)
)
knitr::kable(summary_df, digits = 3)
```

```{r invert-plots, fig.width=10, fig.height=3.5}
op <- par(mfrow = c(1, 3))
plot(inv_cab$obs, inv_cab$pred, pch = 19, col = "#2E8B57",
     xlab = "Observed Cab", ylab = "Predicted", main = sprintf("Cab (R2=%.2f)", inv_cab$R2))
abline(0, 1, col = "grey40", lty = 2)
plot(inv_lai$obs, inv_lai$pred, pch = 19, col = "#2166AC",
     xlab = "Observed LAI", ylab = "Predicted", main = sprintf("LAI (R2=%.2f)", inv_lai$R2))
abline(0, 1, col = "grey40", lty = 2)
plot(inv_vcmax$obs, inv_vcmax$pred, pch = 19, col = "#B2182B",
     xlab = "Observed Vcmax25", ylab = "Predicted", main = sprintf("Vcmax25 (R2=%.2f)", inv_vcmax$R2))
abline(0, 1, col = "grey40", lty = 2)
par(op)
```

`Vcmax25` comes back essentially unretrievable (R² below zero -- worse
than just predicting the mean every time) **even at 500 simulations**:
its only path into reflectance is the small, indirect leaf-temperature
effect the `sensitivity-scope` article isolated, not a direct optical
signature a 13-band sensor can separate from everything else varying in
this LUT at once. More training data doesn't fix this, because the
problem isn't sample size -- it's that the signal genuinely isn't there
in reflectance to begin with. That's the expected result, not a failure
of the method: SCOPE's own coupling of radiative transfer and
biochemistry doesn't imply every biochemical trait becomes remotely
retrievable just because it can be simulated.

`LAI` retrieves well (R² ~0.7) -- its NIR-plateau signature is strong
enough to come through even with every other trait varying at once.
`Cab` retrieves moderately (R² ~0.4) -- weaker than its strong, direct
role in Fluspect alone would suggest, because `getLUT.SCOPE()` varies
every free parameter simultaneously (leaf biochemistry, LAI, leaf-angle
distribution, geometry -- see `getting-luts-scope`), adding real
confounding noise a smaller, more targeted LUT wouldn't have. Sample size
does matter for `Cab` specifically: this page's 500-row LUT retrieves it
noticeably better than an earlier 80-row version of this same experiment
did (R²  0.23 there vs. 0.40 here) -- unlike `Vcmax25`, `Cab` has a real
signal to find, just one more training data helps Random Forest separate
from the LUT's other simultaneous nuisance variation.

See `How-in-R-SCOPEinR.Rmd` (Section 5) for the same LAI-retrieval
pattern at N=100 walked through step by step, and the `ToolsRTM` package's
`InversionOpt` article for the LUT merit-function-matching alternative to
Random Forest (applied there to plain PROSAIL rather than full SCOPE).
