---
title: "Sensitivity analysis for SCOPE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis for SCOPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

SCOPE couples radiative transfer, energy balance, and photosynthesis --
so a trait that has no direct radiative-transfer role (`Vcmax25`,
photosynthetic capacity) can still leave a signature in reflectance
*indirectly*, through the leaf temperature the energy balance solves for.
This page sweeps one trait at a time, holding everything else at the
bundled example row, and looks at three different kinds of SCOPE output:
reflectance (`refl`), canopy temperature (`Tcave`), and photosynthesis
(`Actot`).

```{r baseline}
path_input <- system.file("input", package = "SCOPEinR")
scope_options <- read.table(file.path(path_input, "setoptions.csv"), header = TRUE, sep = ",")
LUT_default <- read.table(file.path(path_input, "LUT_input.csv"), header = TRUE, sep = ",")

run_scope <- function(row) {
  get.SCOPE(LUT = row, options.SCOPE = scope_options, optipar = SCOPEinR::optipar2021.Pro.CX,
            leaf.model = "fluspect-CX", canopy.model = "fourSAIL",
            get.outputs = "ALL", get.plots = FALSE)[[1]]
}
```

## 1. Sweeping `Cab` (chlorophyll content)

```{r sweep-cab}
cab_values <- seq(15, 65, length.out = 6)
res_cab <- lapply(cab_values, function(v) {
  row_i <- LUT_default[1, ]; row_i$Cab <- v
  run_scope(row_i)
})

wl_optical <- 400:2400
n <- length(wl_optical)
cols <- colorRampPalette(c("gold", "darkgreen"))(length(cab_values))
matplot(wl_optical, sapply(res_cab, function(r) r$data.rad$refl[1:n]), type = "l", lty = 1, col = cols,
        xlab = "Wavelength (nm)", ylab = "Reflectance", main = "TOC reflectance sensitivity to Cab")
legend("topright", paste("Cab =", round(cab_values)), col = cols, lty = 1, cex = 0.7)
```

Chlorophyll's own signature (red-edge/visible absorption) dominates here
-- the expected, direct radiative-transfer effect.

## 2. Sweeping `Vcmax25` (photosynthetic capacity)

`Vcmax25` has no role in Fluspect/4SAIL's radiative transfer at all --
any reflectance change it causes is entirely indirect, through the
energy balance:

```{r sweep-vcmax}
vcmax_values <- seq(20, 200, length.out = 6)
res_vcmax <- lapply(vcmax_values, function(v) {
  row_i <- LUT_default[1, ]; row_i$Vcmax25 <- v
  run_scope(row_i)
})

Tcave_vals <- sapply(res_vcmax, function(r) r$data.fluxes$Tcave)
Actot_vals <- sapply(res_vcmax, function(r) r$data.fluxes$Actot)

op <- par(mfrow = c(1, 2))
plot(vcmax_values, Tcave_vals, type = "o", pch = 19, col = "#2166AC",
     xlab = "Vcmax25", ylab = "Canopy-average leaf temperature (degC)", main = "Tcave vs Vcmax25")
plot(vcmax_values, Actot_vals, type = "o", pch = 19, col = "#B2182B",
     xlab = "Vcmax25", ylab = "Actot (umol m-2 s-1)", main = "Actot vs Vcmax25")
par(op)
```

```{r sweep-vcmax-refl}
cols2 <- colorRampPalette(c("gold", "darkblue"))(length(vcmax_values))
matplot(wl_optical, sapply(res_vcmax, function(r) r$data.rad$refl[1:n]), type = "l", lty = 1, col = cols2,
        xlab = "Wavelength (nm)", ylab = "Reflectance",
        main = "TOC reflectance sensitivity to Vcmax25 (indirect, via energy balance)")
legend("topright", paste("Vcmax25 =", round(vcmax_values)), col = cols2, lty = 1, cex = 0.7)
```

`Actot` (canopy photosynthesis) responds strongly and directly to
`Vcmax25`, as expected from the Farquhar-type model. The reflectance
panel is the real point of this section: any spread visible there is
*not* Fluspect/4SAIL reacting to `Vcmax25` (it has no such input) -- it's
the small, indirect effect of a different leaf temperature feeding back
into the thermal part of the spectrum. Compare its magnitude to the
`Cab` sweep above to judge how much (or little) that indirect pathway
actually matters for the optical range.

## 3. Sweeping `LAI`

```{r sweep-lai}
lai_values <- seq(0.5, 6, length.out = 6)
res_lai <- lapply(lai_values, function(v) {
  row_i <- LUT_default[1, ]; row_i$LAI <- v
  run_scope(row_i)
})

cols3 <- colorRampPalette(c("gold", "darkred"))(length(lai_values))
matplot(wl_optical, sapply(res_lai, function(r) r$data.rad$refl[1:n]), type = "l", lty = 1, col = cols3,
        xlab = "Wavelength (nm)", ylab = "Reflectance", main = "TOC reflectance sensitivity to LAI")
legend("topright", paste("LAI =", round(lai_values, 1)), col = cols3, lty = 1, cex = 0.7)
```

NIR reflectance rises with LAI up to a plateau (more leaf layers
scattering, same as plain PROSAIL/4SAIL) -- see the `marmit-soil-in-canopy`
article (ToolsRTM package) for how this same LAI-driven canopy closure
also controls how much a *soil* trait can influence the signal.

## Summary

| Trait swept | Direct RT effect? | Where it shows up |
|---|---|---|
| `Cab` | Yes -- Fluspect input | Visible/red-edge reflectance, strong |
| `LAI` | Yes -- 4SAIL input | NIR plateau height, strong |
| `Vcmax25` | No -- biochemistry only | `Actot` strongly; reflectance only indirectly, via leaf temperature |

See `Scripts/R/Sensibility/2-Sobol_perband_sensitivity.R` (ToolsRTM
package's own scripts) for a formal, per-band Sobol sensitivity analysis
across many traits at once, rather than this page's one-at-a-time sweeps.
