---
title: "Trait & LUT Glossary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trait & LUT Glossary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

`SCOPEinR::get.SCOPE()` reads one LUT row with ~65 columns -- leaf
biochemistry, photosynthesis, fluorescence/NPQ, canopy structure, soil,
aerodynamics, and meteorology, all at once. This page is the field guide
to that row: what every variable physically means, its unit, its
realistic range, and its default -- taken directly from the package's
own `inputs_SCOPE.csv` (not re-typed, so it can't drift out of sync).

For *how* `getLUT.SCOPE()` turns this table into random samples
(`Uniform`/`Gaussian`/`Fixed` sampling, correlating two variables), see
[Getting LUTs for SCOPE](getting-luts-scope.html) -- this page is the
*meaning* of each variable, that page is the *mechanics* of sampling it.
For SCOPE's own run-time numerical/solver options (iteration limits,
convergence tolerances -- not plant/soil/meteo traits at all), see
[Tutorial 00](../t00-scope-options.html).

```{r load-csv}
path_input <- system.file("input", package = "SCOPEinR")
inputLUT <- read.table(file.path(path_input, "inputs_SCOPE.csv"), header = TRUE, sep = ",")
```

## 1. SCOPE's models at a glance

`get.SCOPE()` isn't one model -- it's five distinct components chained
together, each solving a different piece of physics, in this order:

| Component | What it does | Depends on |
|---|---|---|
| **Fluspect-Cx (SCOPE variant)** (`getFluspect.Cx.SCOPE()`) | Leaf optics: reflectance/transmittance + fluorescence excitation-emission matrices, per canopy layer (mSCOPE multi-layer wrapper: `get.fluspect_mSCOPE()`). Same PROSPECT/Fluspect physics as ToolsRTM's leaf models (see its [Parameter & Trait Glossary](../../toolsrtm/articles/parameter-glossary.html#models-at-a-glance)) -- just the SCOPE-specific wrapper. | Leaf biochemistry traits (Section 2) |
| **RTMo** (`get.RTMo()`) | Optical top-of-canopy BRDF -- physically the same turbid-medium idea as fourSAIL, re-implemented to plug into the layers below. Stops at reflectance; assumes no temperature yet. | Fluspect-Cx output + canopy structure (Section 5) + BSM soil |
| **ebal** (`get.ebal()`) | The energy balance: **iteratively solves** leaf and soil temperature so absorbed radiation balances sensible + latent heat, calling the biochemistry model at every candidate temperature. This is what makes SCOPE fundamentally different from PROSAIL/SPART -- temperature is solved for, not assumed. | RTMo output + aerodynamic resistances (Section 7) + meteorology (Section 8) |
| **biochemical** (`get.biochemical()`) | Leaf-level photosynthesis (Farquhar/Collatz) and fluorescence yield, given a leaf micro-environment -- called repeatedly by `ebal` at each candidate temperature, not run standalone. | Photosynthesis traits (Section 3) + NPQ traits (Section 4) |
| **RTMf** (`get.RTMf()`) / **RTMz** (`get.RTMz()`, optional) | Canopy-level fluorescence radiance/flux, and a small zeaxanthin (photoprotection) correction to the TOC spectrum -- both derived from the already-solved energy balance, not computed independently. | ebal output + Fluspect-Cx's fluorescence matrices |

`get.SCOPE()` runs all five for one LUT row and returns everything
together (`data.rad` reflectance/fluorescence, `data.fluxes` energy
balance/photosynthesis). The rest of this page documents the ~65 input
variables these five components read, grouped by which one consumes them.

## 2. Leaf biochemistry (PROSPECT-PRO / Fluspect-Cx traits)

These are the *same* PROSPECT/Fluspect leaf traits ToolsRTM's own
[Parameter & Trait Glossary](../../toolsrtm/articles/parameter-glossary.html)
covers in full detail (meaning, typical range, which leaf model reads
each one) -- SCOPE's leaf-optics step is literally
`getFluspect.Cx.SCOPE()` under the hood. Only the SCOPE-specific
range/default/units, straight from `inputs_SCOPE.csv`, are repeated here:

```{r leaf-table, echo = FALSE}
leaf_vars <- c("N", "Cab", "Car", "Anth", "LMA", "EWT", "alpha", "Cbrown",
                "Cs", "Prot", "CBC", "Cx", "fqe")
knitr::kable(inputLUT[inputLUT$variable %in% leaf_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

`rho_thermal`/`tau_thermal` (both fixed at 0.01) are the leaf's
thermal-infrared (~8-14 micron) reflectance/transmittance -- leaves are
close to blackbody in the TIR, so these stay near zero and are rarely
varied.

## 3. Photosynthesis (Farquhar/Collatz + Ball-Berry)

The traits `scopeinpython.biochemical`/`SCOPEinR::get.biochemical()`
actually consume, at every candidate leaf temperature during the
energy-balance iteration:

```{r photo-table, echo = FALSE}
photo_vars <- c("Vcmax25", "BallBerrySlope", "BallBerry0", "Type", "kV", "Rdparam")
knitr::kable(inputLUT[inputLUT$variable %in% photo_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

- **`Vcmax25`** -- maximum carboxylation capacity of Rubisco at 25degC,
  the single biggest driver of photosynthetic capacity (and therefore
  fluorescence yield) in the whole model. Low values (<20) are stressed/
  senescent canopies; 40-120 is typical healthy crop/forest; the CSV's
  full 0.75-250 range covers everything from near-dead tissue to the
  most productive C4 crops.
- **`BallBerrySlope`**/**`BallBerry0`** -- the Ball-Berry stomatal
  conductance model: `gs = BallBerrySlope * A * RH / Cs + BallBerry0`.
  A steeper slope means stomata track photosynthesis more tightly;
  `BallBerry0` is the residual (cuticular) conductance when `A` -> 0.
- **`Type`** -- photosynthetic pathway, `"C3"` or `"C4"` (stored here as
  `default = 3` -> `"C3"`). Changes which Farquhar sub-model
  (Rubisco-limited vs. PEPcase-limited) is used.
- **`kV`** -- canopy-depth extinction coefficient for `Vcmax` -- real
  canopies down-regulate photosynthetic capacity in shaded lower leaves,
  and `kV` sets how fast.
- **`Rdparam`** -- dark (mitochondrial) respiration as a fraction of
  `Vcmax25` (`Rd = Rdparam * Vcmax25`).

## 4. Fluorescence / non-photochemical quenching (NPQ)

The van der Tol et al. (2014) fluorescence-yield model traits, on top of
the leaf's own `fqe` (Section 2):

```{r fluo-table, echo = FALSE}
fluo_vars <- c("Kn0", "Knalpha", "Knbeta", "Tyear", "beta", "kNPQs", "qLs", "stressfactor", "spectrum")
knitr::kable(inputLUT[inputLUT$variable %in% fluo_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

- **`Kn0`/`Knalpha`/`Knbeta`** -- empirical constants shaping how the NPQ
  rate constant `Kn` responds to light history (the reversible,
  photoprotective quenching most leaves show within minutes).
- **`beta`** -- fraction of absorbed PAR partitioned to Photosystem II
  (the rest goes to PSI); default 0.51 matches the published SCOPE
  default.
- **`kNPQs`**/**`qLs`**/**`stressfactor`** -- *sustained* (slow-recovering,
  stress-related) quenching terms, all defaulted "off"
  (`kNPQs = 0`, `qLs = 1` = fully open PSII, `stressfactor = 1` = no
  down-regulation) -- set these away from default to simulate a
  chronically stressed canopy rather than one just responding to
  instantaneous light.
- **`Tyear`** -- a growth/acclimation temperature (degC) feeding the
  biochemical model's temperature-correction functions.
- **`spectrum`** -- internal switch selecting which bundled
  fluorescence/optical parameter set to use; leave at the default `1`
  unless told otherwise.

## 5. Canopy structure

```{r canopy-table, echo = FALSE}
canopy_vars <- c("LAI", "hc", "LIDFa", "LIDFb", "TypeLidf", "leafwidth", "hspot", "Cv", "crowndiameter")
knitr::kable(inputLUT[inputLUT$variable %in% canopy_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

`LAI`, `LIDFa`/`LIDFb`/`TypeLidf`, and `hspot` are exactly the fourSAIL
traits ToolsRTM's glossary covers (Sections 2-3 there, including the
named-LIDF-shape table). SCOPE-specific additions:

- **`hc`** -- canopy height, needed for the aerodynamic-resistance chain
  (Section 7) that plain optical-only fourSAIL never needs.
- **`leafwidth`** -- characteristic leaf width, used in both the
  aerodynamic boundary-layer resistance and (implicitly) the hot-spot
  geometry.
- **`Cv`** -- vertical foliage clumping/coverage fraction.
- **`crowndiameter`** -- used together with `Cv` for a clumping
  correction, the SCOPE analogue of INFORM's `cd`.

## 6. Soil

```{r soil-table, echo = FALSE}
soil_vars <- c("rss", "rs_thermal", "cs", "rhos", "lambdas", "SMC", "BSMBrightness", "BSMlat", "BSMlon")
knitr::kable(inputLUT[inputLUT$variable %in% soil_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

`SMC`, `BSMBrightness`, `BSMlat`, `BSMlon` are BSM's soil-reflectance
traits -- see ToolsRTM's glossary Section 6 (shared with SPART) for what
each one physically means. SCOPE-specific soil-thermal additions:

- **`rss`** -- soil (sub-surface) resistance to evaporation.
- **`rs_thermal`** -- a thermal-emission surface-resistance analogue used
  in the TIR flux calculation.
- **`cs`**, **`rhos`**, **`lambdas`** -- soil specific heat capacity, bulk
  density, and thermal conductivity, feeding the soil heat-flux/
  temperature part of the energy balance (`ebal`).

## 7. Aerodynamics / energy-balance resistances

Feeds `scopeinpython.thermal`/`SCOPEinR::get.resistances()`; not needed
at all for optical-only runs (`run_rtmo()` alone), only once `ebal()`
closes the energy balance:

```{r aero-table, echo = FALSE}
aero_vars <- c("zo", "d", "Cd", "rb", "CR", "CD1", "Psicor", "CSSOIL", "rbs", "rwc", "z")
knitr::kable(inputLUT[inputLUT$variable %in% aero_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

`zo` (roughness length) and `d` (zero-plane displacement height) are
normally *derived* from canopy height via `get_zo_and_d()` rather than
hand-set -- the fixed defaults here are placeholders, not values meant
to be sampled independently of `hc`.

## 8. Meteorology, geometry, and site/time

```{r meteo-table, echo = FALSE}
meteo_vars <- c("Rin", "Rli", "Ta", "p", "ea", "RH", "u", "Ca", "Oa",
                  "startDate", "endDate", "LAT", "LON", "timezn", "tts", "tto", "psi")
knitr::kable(inputLUT[inputLUT$variable %in% meteo_vars,
                       c("variable", "units", "lower", "upper", "Distribution", "default")],
             row.names = FALSE)
```

- **`Rin`**/**`Rli`** -- incoming shortwave (solar) and longwave
  (thermal) radiation, W/m2.
- **`Ta`**, **`p`**, **`ea`**, **`u`** -- air temperature (degC), pressure
  (hPa), actual vapour pressure (hPa), wind speed (m/s).
- **`RH`** -- a real, documented quirk: the CSV's own `units` column
  says `"hPa"`, but the `0-1` range and `0.64` default show it's
  actually sampled as a **fraction** (0-1), not a pressure -- a
  mislabeled units cell in the CSV itself, reproduced as-is here rather
  than silently "fixed", since it doesn't affect what value is actually
  sampled.
- **`Ca`**, **`Oa`** -- atmospheric CO2 (ppm) and O2 concentration, both
  feeding the Farquhar Ci-solver directly.
- **`tts`**/**`tto`**/**`psi`** -- sun zenith, view zenith, relative
  azimuth (degrees) -- identical geometry traits to fourSAIL/SPART.
- **`LAT`**/**`LON`**/**`timezn`**/**`startDate`**/**`endDate`** -- site
  location and simulation period, used when driving SCOPE from a real
  time series (Tutorial 06) rather than one fixed LUT row.

## 9. A real LUT: sampling and inspecting a few traits

`getLUT.SCOPE()` (see [Getting LUTs for
SCOPE](getting-luts-scope.html) for the full mechanics) reads each
row's `Distribution` and samples accordingly -- `Uniform` traits spread
evenly across `[lower, upper]`, `Gaussian` traits (only `Cab`, by
default) cluster around `Mean_D`/`Std_D` while still respecting
`[lower, upper]`, and `Fixed` traits never vary:

```{r lut-sample, fig.height = 3.5}
set.seed(1)
lut <- getLUT.SCOPE(inputLUT = inputLUT, nLUT = 500, setseed = 1)

op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(lut$Cab, breaks = 20, col = "#009E73", main = "Cab (Gaussian)",
     xlab = "ug/cm2")
hist(lut$LAI, breaks = 20, col = "#0072B2", main = "LAI (Uniform)",
     xlab = "m2/m2")
hist(lut$Vcmax25, breaks = 20, col = "#D55E00", main = "Vcmax25 (Uniform)",
     xlab = "umol/m2/s")
par(op)
```

`Cab`'s histogram is visibly bell-shaped (its `Distribution` is
`"Gaussian"`, `Mean_D = 50`, `Std_D = 20`, still clipped to `[5, 90]`),
while `LAI` and `Vcmax25` (both `"Uniform"`) are flat across their own
ranges -- exactly the sampling behaviour Section 1 of [Getting LUTs for
SCOPE](getting-luts-scope.html) describes, now visible on three
concrete traits from this page's own glossary.

## What's next

- **Tutorial 05** -- building a full SCOPE LUT and running it through
  `get.SCOPE()` end to end.
- **[Getting LUTs for SCOPE](getting-luts-scope.html)** -- the sampling
  mechanics (`Distribution` handling, correlating two traits) this page
  only summarizes.
- **ToolsRTM's [Parameter & Trait
  Glossary](../../toolsrtm/articles/parameter-glossary.html)** -- the
  full leaf/canopy/soil/atmosphere reference for everything SCOPE's
  leaf-optics and canopy-BRDF steps share with plain PROSAIL/SPART.
