---
title: "13. Deep Learning for RTM Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{13. Deep Learning for RTM Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

`getMLmodel()` is `get.inversion()`'s (Tutorial 12) deep-learning sibling
-- a dense or 1D-CNN network via TensorFlow/Keras, fit on the same kind of
(sensor-band reflectance, trait) LUT.

## 1. Same simulate-and-split pattern as Tutorials 10-11

```{r simulate}
n_samples <- 150
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
names(se2a) <- c("id", band_names)

set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
train_df <- cbind(LUT[train_idx, ], se2a[train_idx, band_names])
```

## 2. `getMLmodel()`: deep learning (TensorFlow/Keras)

Needs a working Python/TensorFlow/tf-keras stack reachable through
`reticulate` -- genuinely machine-specific setup. Wrapped in `tryCatch()`
here so this page stays buildable on a machine without that stack
configured, while still showing the real, correct calling code (the same
pattern used throughout this package's own course scripts,
`Scripts/R/*/3-inversion_DL.R`):

```{r dl-inversion}
dl_result <- tryCatch({
  local_venv_root <- file.path(Sys.getenv("LOCALAPPDATA"), "r-reticulate-venvs")
  Sys.setenv(RETICULATE_VIRTUALENV_ROOT = local_venv_root)
  Sys.setenv(WORKON_HOME = local_venv_root)
  Sys.setenv(TF_USE_LEGACY_KERAS = "1")
  library(reticulate); library(keras)
  dl_model <- getMLmodel(dataset = train_df, depVar = "Cab", model = "Hidden-layers",
                          optimizer = "adam", n.epochs = 5)
  list(ok = TRUE, model = dl_model)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))

if (dl_result$ok) {
  cat("DL model trained. Final training loss:",
      round(tail(dl_result$model$history$metrics$loss, 1), 3), "\n")
} else {
  first_line <- strsplit(dl_result$msg, "\n", fixed = TRUE)[[1]][1]
  cat("Skipped -- no working Python/TensorFlow/tf-keras stack reachable through",
      "reticulate on this machine (", first_line, "). The calling code above is",
      "correct and unchanged.\n")
}
```

`model = "Hidden-layers"` can be swapped for `"CNN"` (1D convolutional,
treating the band sequence as a spatial dimension) -- same call otherwise.
`n.epochs = 5` above is a smoke test, not a tuned model; a real run uses
far more (`Scripts/R/*/3-inversion_DL.R` trains both architectures per
trait to convergence).

**Important, machine-specific setup note**: `getMLmodel()` calls
`callback_early_stopping()` without a package prefix internally, so
`library(keras)` must actually be *attached* (not just installed) before
calling it, or the fit fails with
`could not find function "callback_early_stopping"` -- every DL script in
this package's `Scripts/R/` starts with the exact `Sys.setenv()`/
`library()` block shown above for that reason.

## 3. Which one should you use? (Tutorials 10-12, together)

| Method | Needs training? | Handles a small LUT well? | Typical use |
|---|---|---|---|
| `get.inversionOpt()` (Tutorial 11) | No -- pure search | Yes -- works from a handful of reference spectra | Quick, physically-grounded retrieval; no ML infrastructure needed |
| `get.inversion()` (Tutorial 12) | Yes | Needs enough rows to fit reliably (hundreds+) | Production-scale retrieval, many predictors (full spectrum + indices) |
| `getMLmodel()` (this page) | Yes, more data-hungry | No -- needs thousands of rows | Large training sets, CNN over the full spectral shape |

## What's next

- **Tutorial 14** -- simulate, convolve, index, and invert as one
  end-to-end pipeline, across every canopy model this package supports.
