---
title: "13. Deep Learning for RTM Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{13. Deep Learning for RTM Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

`getMLmodel()` is `get.inversion()`'s (Tutorial 12) deep-learning sibling
-- a dense or 1D-CNN network via TensorFlow/Keras, fit on the same kind of
(sensor-band reflectance, trait) LUT. Needs a working Python/TensorFlow/
tf-keras stack reachable through `reticulate` -- genuinely machine-
specific setup, wrapped in `tryCatch()` throughout this page so it stays
buildable on a machine without that stack configured, while showing the
real, correct calling code either way.

```text
RTM LUT (getLUT)
     |
     v
Spectral simulation (foursail)
     |
     v
Sensor / hyperspectral representation (Sentinel-2A bands, or PRISMA bands for the CNN)
     |
     v
Train / validation / test split
     |
     v
Normalization (fit on train only)
     |
     v
DL training, with early stopping on validation loss
     |
     v
Prediction on the INDEPENDENT test set
     |
     v
R2 / RMSE / bias, true-vs-predicted plot
     |
     v
Comparison against Tutorial 12's Random Forest, same split
```

## 1. Three real bugs, found and fixed while building this page

**Bug 1 -- zero-variance columns break the internal normalization.** An
earlier version of this page passed `getMLmodel()` a `dataset`
containing the *entire* training LUT (every trait column) concatenated
with the reflectance bands, and training failed with
`infinite or missing values in 'x'`. Traced to the real cause:
`getMLmodel()` has no `inputs =` argument -- unlike `get.inversion()`
(Tutorial 12), it treats **every column of `dataset` other than
`depVar`** as a predictor. `inputsPROSAIL`'s LUT fixes several trait
columns at a single constant value by design (`LMA`, `alpha`, `LIDFb`,
`TypeLidf`, `hspot`, `tts`, `psi` all have zero variance in a plain
`getLUT()` draw) -- feeding those into `getMLmodel()`'s internal
`Normalize` preprocessing (range-scaling each column by its own
min/max) divides by zero for every one of them, producing the `Inf`/`NA`
values the error message reports. **Fix**: build `dataset` from only
`depVar` plus the actual reflectance-band columns, exactly as
`get.inversion(..., inputs = band_names)` already does explicitly for
Tutorial 12 -- `getMLmodel()` needs the same discipline applied
manually, since it has no `inputs =` parameter to enforce it.

**Bug 2 -- a `relu` output layer can die permanently.** Both the
`'Hidden-layers'` and `'CNN'` branches inside `getMLmodel()` ended their
last, single-unit regression layer with `activation = 'relu'`. A relu
unit's gradient is exactly zero whenever its pre-activation is negative
-- for a single output unit on an unbounded-scale regression target,
one unlucky batch of negative pre-activations early in training can
permanently zero out that unit's gradient, with no way back regardless
of epoch budget. **Fix**: changed to `activation = 'linear'`, the
standard choice for an unbounded regression output (fixed directly in
`getMLmodel.R`/`getMLmodel_withRetrain.R`, not worked around here).

**Bug 3 -- predicting in the wrong input space (the big one).**
`getMLmodel()`'s training step range-scales the input bands to \[0,1\]
internally (via `getSplitData()`), fits the network on that scaled
data, and returns the raw Keras model *plus* the fitted scaler
(`$Scalar.train`) as two separate pieces -- it does **not** wrap them
together into a single self-scaling `predict()`. Every earlier version
of this page called `predict(model$model, as.matrix(test_bands))`
directly on the **raw, unscaled** test reflectance -- training and
prediction were silently happening in two different input spaces. This
is precisely why Sections 5-7's dense/CNN results used to look
identical to a badly overfit or broken model (near-zero-to-catastrophic
test R2) despite a completely normal-looking internal training curve:
the network was well-trained *for scaled inputs* and then evaluated on
inputs it had never actually seen the distribution of. **Fix**: apply
the returned `Scalar.train` to test data with
`predict(model$Scalar.train, test_data)` before passing it to the
network, exactly the pattern `ToolsRTM::getPredicts()` (the package's
own dedicated prediction helper, which also handles reversing
`depVar.trans`'s target scaling via `getReverse.trans()`) already
assumes callers will follow. This single fix is what turns Sections
5-7 below from "deep learning doesn't really work here" into deep
learning genuinely competing with Random Forest -- Bugs 1-2 were real,
but this one was the actual story.

```{r setup-check}
local_venv_root <- file.path(Sys.getenv("LOCALAPPDATA"), "r-reticulate-venvs")
Sys.setenv(RETICULATE_VIRTUALENV_ROOT = local_venv_root)
Sys.setenv(WORKON_HOME = local_venv_root)
Sys.setenv(TF_USE_LEGACY_KERAS = "1")
keras_ok <- tryCatch({ library(reticulate); library(keras); TRUE }, error = function(e) FALSE)
cat("Working Keras/TensorFlow stack available:", keras_ok, "\n")
```

**Important, machine-specific setup note**: `getMLmodel()` calls
`callback_early_stopping()` without a package prefix internally, so
`library(keras)` must actually be *attached* (not just installed)
before calling it, or the fit fails with
`could not find function "callback_early_stopping"`.

## 2. A properly sized LUT, not a 150-row smoke test

`n_samples = 150` (an earlier version of this page) is far too small to
demonstrate what deep learning is actually for -- 6000 rows here, large
enough to show real training dynamics without an impractical pkgdown
build time:

```{r simulate-large}
r2_f <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))
bias_f <- function(obs, pred) mean(pred - obs)

n_samples <- 6000
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
names(se2a) <- c("id", band_names)
```

## 3. A real train / validation / test split -- evaluated only on the held-out test set

RTM inversion specifically needs an *independent* test set, not just a
train/test split: the LUT is synthetic, so nothing stops the model from
memorizing quirks of the exact rows it trained on unless a third,
completely untouched slice is reserved for the final number reported.
60% train / 20% validation (used *during* training, for early stopping)
/ 20% test (touched only once, at the end):

```{r split}
set.seed(1)
idx <- sample(seq_len(n_samples))
n_train <- round(0.6 * n_samples); n_val <- round(0.2 * n_samples)
train_idx <- idx[1:n_train]
val_idx   <- idx[(n_train + 1):(n_train + n_val)]
test_idx  <- idx[(n_train + n_val + 1):n_samples]
cat("train:", length(train_idx), " val:", length(val_idx), " test:", length(test_idx), "\n")

dataset_dense <- cbind(Cab = LUT$Cab, se2a[, band_names])  # depVar + bands only -- Section 1's fix
train_df <- dataset_dense[train_idx, ]
```

**Normalization, stated explicitly**: `getMLmodel()`'s internal
`getSplitData()` step fits its range/PCA scaler on the training split
only, then applies that *same* fitted transform to whatever data is
passed to `predict()` later -- no information from validation or test
rows leaks into the scaling parameters. This matters specifically for
RTM inversion because the LUT's own trait ranges (not real-world
variability) determine what "normal" looks like to the model; fitting
the scaler on the full dataset would let test-set values quietly
influence how training data gets scaled.

## 4. Dense network, with early stopping, on Sentinel-2 bands

```{r dl-dense, results = 'hide'}
dl_result <- tryCatch({
  dl_dense <- getMLmodel(dataset = train_df, depVar = "Cab", model = "Hidden-layers",
                          optimizer = "adam", n.epochs = 300, batch.size = 64, save.model = FALSE)
  list(ok = TRUE, model = dl_dense)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))
```

```{r dl-dense-check}
if (dl_result$ok) {
  h <- dl_result$model$history$metrics
  cat("Epochs actually run:", length(h$loss), "of 300 requested\n")
  cat("Epoch with lowest validation loss:", which.min(h$val_loss), "\n")
} else {
  cat("Skipped -- no working Python/TensorFlow/tf-keras stack on this machine (",
      strsplit(dl_result$msg, "\n", fixed = TRUE)[[1]][1], ").\n")
}
```

```{r learning-curve, fig.width=7}
if (dl_result$ok) {
  h <- dl_result$model$history$metrics
  plot(seq_along(h$loss), h$loss, type = "l", col = "#2166AC",
       ylim = range(c(h$loss, h$val_loss)), xlab = "Epoch", ylab = "Loss (MSE)",
       main = "Learning curve: training vs. validation loss")
  lines(seq_along(h$val_loss), h$val_loss, col = "#D55E00")
  legend("topright", c("Training loss", "Validation loss"), col = c("#2166AC", "#D55E00"), lty = 1)
}
```

**Read the learning curve before trusting the model.** In this run,
`getMLmodel()`'s built-in `callback_early_stopping()` (`patience = 5`,
monitoring validation loss) never actually triggered -- the lowest
validation loss is at the very last epoch of the 300-epoch budget, so
the network was still (slowly) improving when training stopped, not
demonstrably plateaued. That's a real, honest limitation of this demo
(a production run would keep training further, or raise the learning
rate) -- but it is not the same claim the pre-Bug-3 version of this
page made: this model already reaches R2=0.84 on the independent test
set (Section 5) *despite* not having fully converged, which is a
meaningfully different, more encouraging situation than the R2-near-
zero, still-not-converged result an unfixed `predict()` call used to
produce. There is real headroom left on the table here (more epochs,
a lower learning rate, or both would likely improve on 0.84 further),
not a broken model to explain away.

## 5. Independent-test-set result -- and a direct comparison against Tutorial 12's Random Forest

```{r dl-test-eval}
if (dl_result$ok) {
  # A real bug, found by re-reading this exact chunk: getMLmodel() trains
  # on X range-scaled to [0,1] internally (via getSplitData()) but returns
  # the raw keras model plus a *separate* scaler -- predict() does not
  # re-apply that scaler for you. Feeding raw, unscaled reflectance here
  # would train and predict in two different input spaces, which is
  # consistent with everything Sections 5-7 originally showed: clean
  # internal train/val convergence, but a badly broken held-out test R2.
  X_test_scaled <- predict(dl_result$model$Scalar.train, dataset_dense[test_idx, band_names])
  pred_test_dl <- as.numeric(predict(dl_result$model$model, as.matrix(X_test_scaled)))
  obs_test <- dataset_dense$Cab[test_idx]
  cat("Dense DL, independent test set: R2=", round(r2_f(obs_test, pred_test_dl), 3),
      " RMSE=", round(rmse_f(obs_test, pred_test_dl), 2),
      " bias=", round(bias_f(obs_test, pred_test_dl), 2), "\n")
}
```

```{r rf-comparison}
if (dl_result$ok) {
  fit_rf <- get.inversion(data = dataset_dense[train_idx, ], depVar = "Cab", inputs = band_names,
                           algorithm = "RF", n.samples = length(train_idx), seed = 42)
  pred_test_rf <- as.numeric(predict(fit_rf$model, newdata = dataset_dense[test_idx, c("Cab", band_names)]))
  comparison <- data.frame(
    method = c("Dense DL (getMLmodel)", "Random Forest (get.inversion, Tutorial 12)"),
    R2 = round(c(r2_f(obs_test, pred_test_dl), r2_f(obs_test, pred_test_rf)), 3),
    RMSE = round(c(rmse_f(obs_test, pred_test_dl), rmse_f(obs_test, pred_test_rf)), 2)
  )
  knitr::kable(comparison, row.names = FALSE)
}
```

```{r true-vs-predicted, fig.width=9}
if (dl_result$ok) {
  op <- par(mfrow = c(1, 2))
  plot(obs_test, pred_test_dl, pch = 19, col = "#D55E00",
       xlab = "Observed Cab", ylab = "Predicted Cab", main = "Dense DL")
  abline(0, 1, col = "grey40", lty = 2)
  plot(obs_test, pred_test_rf, pch = 19, col = "#2166AC",
       xlab = "Observed Cab", ylab = "Predicted Cab", main = "Random Forest")
  abline(0, 1, col = "grey40", lty = 2)
  par(op)
}
```

**On this LUT (6000 rows, 10 Sentinel-2 bands), Random Forest still has
a real edge, but the dense network is now a genuine competitor, not a
broken demo.** R2=0.822/RMSE=6.53 for the dense network against
R2=0.858/RMSE=5.84 for RF -- a real gap, but a small, ordinary one
between two properly-fit models, nothing like the catastrophic
mismatch Bug 3 above used to produce. Ten multispectral bands and a
6000-row LUT are close to RF's best case, so RF keeping a modest lead
here is a reasonable, unsurprising result -- not evidence that deep
learning "doesn't work" for this problem.

## 6. The 1D-CNN, on hyperspectral (PRISMA) bands

A dense network on 10 multispectral bands is a weak demonstration of
what a 1D CNN specifically buys you -- convolution exploits
*spectral-neighborhood* structure (adjacent, correlated bands), which
10 widely-spaced multispectral bands barely have. PRISMA's ~230
contiguous hyperspectral bands are a much fairer test of that idea:

```{r cnn-prisma, results = 'hide'}
n_hyp <- 4000
LUT_h <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_hyp, setseed = 3))
refl_h <- t(sapply(seq_len(n_hyp), function(i) {
  foursail(inputLUT = LUT_h[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X_h <- as.data.frame(refl_h); colnames(refl_X_h) <- paste0("X", wl); refl_X_h <- cbind(id = seq_len(n_hyp), refl_X_h)
prisma <- suppressMessages(get.spectra.convolved(rfl = refl_X_h, sensor = "PRISMA", plot.spectra = FALSE))
prisma_bands <- paste0("P", seq_len(ncol(prisma) - 1))
names(prisma) <- c("id", prisma_bands)
cat("PRISMA bands:", length(prisma_bands), "\n")

idx_h <- sample(seq_len(n_hyp))
train_h <- idx_h[1:round(0.6 * n_hyp)]
test_h  <- idx_h[(round(0.8 * n_hyp) + 1):n_hyp]
dataset_cnn <- cbind(Cab = LUT_h$Cab, prisma[, prisma_bands])

cnn_result <- tryCatch({
  m <- getMLmodel(dataset = dataset_cnn[train_h, ], depVar = "Cab", model = "CNN",
                   optimizer = "adam", n.epochs = 300, batch.size = 64, save.model = FALSE)
  list(ok = TRUE, model = m)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))
```

```{r cnn-eval}
if (cnn_result$ok) {
  # Same train/predict scaling mismatch fixed in Section 5 -- apply the
  # returned Scalar.train to the test bands before feeding the network.
  X_test_h_scaled <- predict(cnn_result$model$Scalar.train, dataset_cnn[test_h, prisma_bands])
  pred_test_cnn <- as.numeric(predict(cnn_result$model$model,
    array_reshape(as.matrix(X_test_h_scaled), c(length(test_h), length(prisma_bands), 1))))
  obs_test_h <- dataset_cnn$Cab[test_h]
  cat("CNN (PRISMA, ", length(prisma_bands), "bands), independent test set: R2=",
      round(r2_f(obs_test_h, pred_test_cnn), 3), " RMSE=", round(rmse_f(obs_test_h, pred_test_cnn), 2), "\n")
}
```

```{r cnn-true-vs-predicted, fig.width=5, fig.height=5}
if (cnn_result$ok) {
  plot(obs_test_h, pred_test_cnn, pch = 19, col = "#009E73",
       xlab = "Observed Cab", ylab = "Predicted Cab",
       main = sprintf("1D-CNN, PRISMA (%d bands)", length(prisma_bands)))
  abline(0, 1, col = "grey40", lty = 2)
}
```

The dense-vs-1D-CNN architectural difference itself: `"Hidden-layers"`
is a stack of plain fully-connected layers, treating each band as an
independent input with no notion of spectral order; `"CNN"` instead
runs 1D convolution + pooling over the bands *in their spectral
sequence*, so it can learn local absorption-feature shapes (a dip
spanning several adjacent bands) as a reusable pattern -- something a
dense network has to learn independently, band-combination by band-
combination, with no structural help.

**R2=0.832/RMSE=6.64 on 234 raw PRISMA bands -- in the same range as
Section 5's 10-band comparison (dense 0.822, RF 0.858), not a
degradation from the much higher input dimensionality.** That itself
is worth noting: a plain dense network on 10 bands and a 1D-CNN on 234
land in a similar place here, which says the CNN's structural
advantage (exploiting spectral neighborhoods) isn't yet the dominant
factor on this particular synthetic Cab-from-reflectance mapping --
not because the CNN is broken, but because there isn't much local
spectral structure in this problem that a well-tuned dense network or
RF weren't already capturing from the bands directly.

## 7. Fewer, better features: does band+index selection help more than raw bands?

Sections 4-6 threw every band at the network -- all 10 Sentinel-2
bands, all 234 PRISMA bands -- and let the model find what matters on
its own. Tutorial 12 showed a different, more targeted approach for
Random Forest: a compact core of physically relevant bands, plus a
small, *per-trait* selection of the most correlated spectral indices
(Tutorial 09). Does that also help a dense network, and does it hold
for a second trait (LAI), not just Cab?

```{r feature-select-setup}
# Real Sentinel-2A band names (getIndicesSE2.ML()'s formulas look up
# specific bands like B8A/B11/B12 by name) -- a second, separate naming of
# the same se2a convolution from Section 2; band_names/dataset_dense above
# are untouched.
se2a_real <- se2a
names(se2a_real) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
core_bands <- c("B4", "B5", "B7", "B8A", "B11")  # a compact, physically-motivated core -- not all 10

idx_ml <- suppressMessages(getIndicesSE2.ML(df = se2a_real[, -1], sensor = "Sentinel-2a", df.data = NULL, fast.process = TRUE))

n_top <- 4  # indices kept per trait
select_indices <- function(trait) {
  cors <- sapply(names(idx_ml), function(nm) suppressWarnings(cor(idx_ml[train_idx, nm], LUT[train_idx, trait])))
  cors <- cors[is.finite(cors)]
  names(sort(abs(cors), decreasing = TRUE))[seq_len(min(n_top, length(cors)))]
}
selected <- setNames(lapply(c("Cab", "LAI"), select_indices), c("Cab", "LAI"))
knitr::kable(data.frame(trait = names(selected),
                         features = sapply(selected, function(x) paste(c(core_bands, x), collapse = ", "))),
             row.names = FALSE)
```

```{r feature-select-fit, results='hide', fig.show='hide'}
run_trait <- function(trait) {
  feat <- c(core_bands, selected[[trait]])
  df <- cbind(LUT[trait], se2a_real[core_bands], idx_ml[selected[[trait]]])

  # depVar.trans=TRUE: also range-scale the *target*, not just the inputs.
  # With only a handful of highly informative features the network fits
  # training data almost immediately; on LAI's narrow raw range (~0.5-7)
  # that made unscaled-target MSE gradients explode early in training
  # (verified: R2 as bad as -5000 with depVar.trans's default of FALSE).
  # predict() then returns range-scaled [0,1] predictions, so invert them
  # by hand using the returned scaler before computing R2 in real units.
  dl <- getMLmodel(dataset = df[train_idx, ], depVar = trait, model = "Hidden-layers",
                    optimizer = "adam", n.epochs = 300, batch.size = 64, save.model = FALSE,
                    depVar.trans = TRUE)
  # Same X-scaling mismatch fixed in Sections 5-6, found and fixed *after*
  # this section's original depVar.trans experiment -- explains why scaling
  # Y alone made things worse, not better: X was still being fed raw either
  # way, and now two mismatched spaces (X raw-vs-scaled, Y real-vs-scaled)
  # were compounding instead of one.
  X_test_scaled <- predict(dl$Scalar.train, df[test_idx, feat])
  pred_dl_scaled <- as.numeric(predict(dl$model, as.matrix(X_test_scaled)))
  y_range <- dl$Scalar.Ytrain$ranges[, trait]
  pred_dl <- pred_dl_scaled * (y_range[2] - y_range[1]) + y_range[1]

  fit_rf <- get.inversion(data = df[train_idx, ], depVar = trait, inputs = feat,
                           algorithm = "RF", n.samples = length(train_idx), seed = 42)
  pred_rf <- as.numeric(predict(fit_rf$model, newdata = df[test_idx, c(trait, feat)]))

  obs <- LUT[test_idx, trait]
  list(summary = data.frame(trait = trait, n_features = length(feat),
                             R2_dense = r2_f(obs, pred_dl), R2_RF = r2_f(obs, pred_rf)),
       obs = obs, pred_dl = pred_dl, pred_rf = pred_rf)
}
trait_runs <- setNames(lapply(c("Cab", "LAI"), run_trait), c("Cab", "LAI"))
feature_results <- do.call(rbind, lapply(trait_runs, `[[`, "summary"))
```

```{r feature-select-table}
knitr::kable(feature_results, digits = 3, row.names = FALSE)
```

```{r feature-select-plot, fig.width=9, fig.height=9}
op <- par(mfrow = c(2, 2))
for (trait in c("Cab", "LAI")) {
  r <- trait_runs[[trait]]
  lims <- range(c(r$obs, r$pred_dl, r$pred_rf))
  plot(r$obs, r$pred_dl, pch = 19, col = "#D55E00", xlim = lims, ylim = lims,
       xlab = paste("Observed", trait), ylab = paste("Predicted", trait),
       main = sprintf("%s -- Dense, 9 features (R2=%.3f)", trait, r2_f(r$obs, r$pred_dl)))
  abline(0, 1, col = "grey40", lty = 2)
  plot(r$obs, r$pred_rf, pch = 19, col = "#2166AC", xlim = lims, ylim = lims,
       xlab = paste("Observed", trait), ylab = paste("Predicted", trait),
       main = sprintf("%s -- Random Forest, 9 features (R2=%.3f)", trait, r2_f(r$obs, r$pred_rf)))
  abline(0, 1, col = "grey40", lty = 2)
}
par(op)
```

**Fewer, better-chosen features help both models, and now dense
genuinely competes with -- on LAI, even edges out -- Random Forest.**
On Cab, dense (0.840) and RF (0.853) are essentially tied, both
matching or slightly beating Section 5's all-10-band RF baseline
(0.858) from a fraction of the inputs. On LAI, never even attempted
with the generic 10-band set above, dense (0.689) *narrowly beats* RF
(0.685) -- a real, if small, reversal of the pattern in Sections 5-6.

This result looked completely different -- both `R2_dense` values
catastrophically negative -- before Bug 3 in Section 1 was found and
fixed. That's worth sitting with: the earlier, broken version of this
exact section would have "shown" that giving a dense network fewer,
better features makes it *worse*, actively contradicting Tutorial 12's
finding for RF and inviting exactly the wrong lesson (dense networks
can't handle small, informative feature sets). The real lesson is the
opposite -- once dense and RF are compared on equal footing, feature
selection helps both, and a properly trained dense network on a
handful of informative inputs is perfectly competitive, not fragile.

## 8. Which one should you use? (Tutorials 10-13, together) -- corrected

An earlier version of this table implied `get.inversionOpt()` (pure
LUT search, no training) works fine from a small reference LUT. Tutorial
11's own real test contradicts that: at ~100 reference spectra,
`get.inversionOpt()` scored R² close to zero (no real skill); it only
became useful around 2000. **All three methods need substantially more
data than a quick demo LUT provides for genuinely reliable results** --
the difference between them is *how* they degrade and recover, not
whether one of them is exempt from needing data:

| Method | Needs training? | Small-LUT behavior (verified) | Typical use |
|---|---|---|---|
| `get.inversionOpt()` (Tutorial 11) | No -- pure search | Degrades smoothly but really does need real coverage -- ~100 reference spectra gave R2 near zero in Tutorial 11's own test; ~2000 gave real skill | Quick, physically-grounded retrieval when a genuinely well-populated reference LUT already exists |
| `get.inversion()` (Tutorial 12) | Yes | The most forgiving of the three in practice -- RF stayed usable even on Tutorial 12's 150-row LUT, and keeps a small, consistent edge over dense/CNN on this page's 6000-row one (Sections 5-6) | General-purpose retrieval; often the right default |
| `getMLmodel()` (this page) | Yes, more so, and *easy to silently misuse* | Once trained and evaluated correctly (Section 1's three bugs fixed), competitive with RF -- 0.822-0.840 R2 on Cab, actually edges out RF on LAI with a compact feature set (Section 7). Still the most data-hungry of the three, and the least forgiving of a scaling mistake: a wrong `predict()` call produces a *plausible-looking* broken number, not an error | Large LUTs, hyperspectral input, and/or when the extra setup care is worth it |

## 9. When should I use deep learning?

**Once implemented correctly, more readily than this page's earlier,
broken version of itself would have suggested.** The real lesson of
this page isn't "DL underperforms here" -- Bug 3 in Section 1 made it
look that way, and that conclusion would have been wrong. Once
training and prediction actually operate in the same input space, a
plain dense network is a genuine, competitive option, not a fallback
for when you have no other choice. What still tips the decision toward
`get.inversion()` (Tutorial 12) as the default, or toward `getMLmodel()`
specifically, is more mundane:

- **Setup discipline, honestly the main cost now**: `get.inversion()`'s
  `predict()` (an ordinary caret model object) handles its own
  preprocessing internally -- you cannot forget to rescale test data,
  because there's no separate scaler to remember. `getMLmodel()`
  returns the raw network and its scaler as two pieces on purpose (for
  flexibility), but that means the caller is responsible for
  `predict(model$Scalar.train, newdata)` before every prediction, and
  for reversing `depVar.trans` with `getReverse.trans()` if it was
  used. Bug 3 above is exactly what forgetting that step looks like --
  worth internalizing before reaching for `getMLmodel()` in your own
  code, not just reading about it here.
- **LUT size**: still real -- thousands of rows, not hundreds, gave
  both dense and CNN room to actually learn (Section 2). Nothing here
  suggests DL degrades *more gracefully* than RF on a small LUT; if
  anything, assume the opposite until shown otherwise.
- **Input dimensionality**: hyperspectral (dozens to hundreds of
  correlated bands) is where a CNN's spectral-neighborhood structure
  has real potential to exploit local absorption-feature shape (Section
  6) -- though on this page's own synthetic Cab mapping, that potential
  didn't yet translate into a clear edge over the plain dense network.
- **Overfitting risk, watched, not assumed**: the learning-curve check
  in Section 4 is not optional -- a validation curve that's still
  falling means the reported test number isn't from a converged model;
  one that's risen back up after a minimum means the model has started
  memorizing the (synthetic) training LUT specifically. A clean-looking
  learning curve is necessary but, as this page found out the hard way,
  nowhere near sufficient -- it says nothing about whether the
  *prediction* pipeline is internally consistent with training.

For this package's own bundled models, `get.inversion()` (Tutorial 12)
remains the lower-friction default -- not because deep learning
performs worse, but because it is much harder to silently misuse.
Reach for `getMLmodel()` when the LUT is genuinely large or the sensor
is hyperspectral, and budget real attention for Section 1's setup
discipline, not just training epochs.

## What's next

- **Tutorial 14** -- simulate, convolve, index, and invert as one
  end-to-end pipeline, across every canopy model this package supports.
