---
title: "13. Deep Learning for RTM Inversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{13. Deep Learning for RTM Inversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

## 0. What will we learn?

By the end of Part I you will be able to: prepare an RTM look-up table
as a (predictors, target) dataset, split it into train/validation/test,
understand *why* and *how* the inputs get normalized, train a dense
neural network with `getMLmodel()`, and predict on new spectra
correctly (the single step that is easiest to get wrong). Part II
extends the same workflow to a 1D-CNN on hyperspectral bands. A
Troubleshooting section and an Advanced section come last, deliberately
-- they're real, but they're not what a first read of this page needs.

`getMLmodel()` is `get.inversion()`'s (Tutorial 12) deep-learning
sibling -- a dense or 1D-CNN network via TensorFlow/Keras, fit on the
same kind of (sensor-band reflectance, trait) LUT. It needs a working
Python/TensorFlow/tf-keras stack reachable through `reticulate`:

```{r setup-check}
local_venv_root <- file.path(Sys.getenv("LOCALAPPDATA"), "r-reticulate-venvs")
Sys.setenv(RETICULATE_VIRTUALENV_ROOT = local_venv_root)
Sys.setenv(WORKON_HOME = local_venv_root)
Sys.setenv(TF_USE_LEGACY_KERAS = "1")
keras_ok <- tryCatch({ library(reticulate); library(keras); TRUE }, error = function(e) FALSE)
cat("Working Keras/TensorFlow stack available:", keras_ok, "\n")
```

**Machine-specific setup note**: `getMLmodel()` calls
`callback_early_stopping()` without a package prefix internally, so
`library(keras)` must actually be *attached* (not just installed)
before calling it, or the fit fails with
`could not find function "callback_early_stopping"`. The rest of this
page is wrapped in `tryCatch()` so it stays buildable on a machine
without that stack configured, while still showing the real, correct
calling code either way.

# Part I -- Learning `getMLmodel()`

## 1. Understanding `getMLmodel()`

```r
getMLmodel(
  dataset,     # data.frame: depVar column + predictor columns
  depVar,      # name of the column to predict, e.g. "Cab"
  model,       # "Hidden-layers" (dense MLP) or "CNN" (1D convolution)
  optimizer,   # e.g. "adam" -- see ?getMLmodel for all 7 supported
  n.epochs,    # maximum training iterations
  batch.size,  # number of samples per weight update
  save.model   # write the fitted model to disk?
)
```

| Argument | What it means |
|---|---|
| `dataset` | A table with spectral predictors *and* the target trait, one row per simulation/observation |
| `depVar` | The trait you want to recover, e.g. `"Cab"` |
| `model` | Architecture: `"Hidden-layers"` (dense) or `"CNN"` (1D convolution over the spectrum) |
| `optimizer` | Optimization algorithm, e.g. `"adam"` |
| `n.epochs` | Maximum number of training iterations (early stopping can stop sooner) |
| `batch.size` | How many rows are used per gradient update |
| `save.model` | Whether to write the fitted `.hdf5` model to disk |

**Important -- read this before calling `getMLmodel()` on your own
data.** Unlike `get.inversion()` (Tutorial 12), `getMLmodel()` has no
`inputs =` argument: **every column of `dataset` other than `depVar`
is treated as a predictor.** If `dataset` still contains other trait
columns, ID columns, or a column that happens to be constant in your
particular sample, they all get fed to the network as if they were
real spectral bands -- a constant column in particular breaks the
internal normalization outright (see Troubleshooting, Section 12).
Always build `dataset` from exactly `depVar` plus the columns you
actually want as predictors, nothing else.

`"Hidden-layers"` and `"CNN"` differ in how they read the predictor
columns: a dense network treats each one as an independent number with
no notion of order; a CNN treats them as a *sequence* (adjacent columns
= adjacent wavelengths) and convolves over that sequence. Part II
covers this difference and when it matters.

## 2. Prepare a simple RTM dataset

Simulate a moderate-sized PROSAIL look-up table, convolve it to
Sentinel-2A bands, and keep only Cab (the target) and the bands
(the predictors) -- exactly the discipline Section 1's "Important" box
asked for:

```{r simulate-large}
r2_f <- function(obs, pred) 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
rmse_f <- function(obs, pred) sqrt(mean((obs - pred)^2))
bias_f <- function(obs, pred) mean(pred - obs)

n_samples <- 6000
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500
rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(nrow(refl_X)), refl_X)
se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
names(se2a) <- c("id", band_names)

dataset_dense <- cbind(Cab = LUT$Cab, se2a[, band_names])
```

This is what a `getMLmodel()` dataset actually looks like -- one column
is the target, the rest are predictors:

```{r show-dataset}
knitr::kable(round(head(dataset_dense, 4), 3))
```

```text
 Cab    B1    B2    B3   ...   B10
35.2  .041  .052  .087  ...   .412
52.1  .036  .048  .071  ...   .486
 ...
  ^      └──────────────────┘
target        predictors
```

`n_samples = 6000` is deliberate, not arbitrary: an earlier, much
smaller draft of this page used 150 rows, which is far too little for
a network to learn real structure from (Section 14 returns to *why*
LUT size matters this much for `getMLmodel()` specifically).

## 3. Train / validation / test split

RTM inversion needs an *independent* test set, not just a train/test
split: the LUT is synthetic, so nothing stops a model from memorizing
quirks of the exact rows it trained on unless a third, completely
untouched slice is reserved for the final number reported.

```text
|---------- train (60%) ----------|--- val (20%) ---|--- test (20%) ---|
   fits the network's weights      used *during*       touched only
                                   training, for        once, at the
                                   early stopping        very end
```

```{r split}
set.seed(1)
idx <- sample(seq_len(n_samples))
n_train <- round(0.6 * n_samples); n_val <- round(0.2 * n_samples)
train_idx <- idx[1:n_train]
val_idx   <- idx[(n_train + 1):(n_train + n_val)]
test_idx  <- idx[(n_train + n_val + 1):n_samples]
cat("train:", length(train_idx), " val:", length(val_idx), " test:", length(test_idx), "\n")

train_df <- dataset_dense[train_idx, ]
```

## 4. Normalization: why, and why train-only

**Why normalize at all?** A neural network learns by gradient descent
on its weights. When predictors sit on very different numeric scales,
the loss surface becomes elongated and harder to optimize -- some
weights need tiny updates, others huge ones, for the same learning
rate. Rescaling every predictor onto a common range (here, \[0,1\]) is
what lets one learning rate work reasonably well for all of them.

```text
Original reflectance                    After range-normalization

  B4      B5      B8                      B4      B5      B8
 0.04    0.09    0.43                    0.00    0.00    0.00
 0.06    0.12    0.51        -->         0.42    0.36    0.51
  ...                                     ...
```

**Why fit the scaler on training data only?** Using the test set's own
min/max (or mean/sd) to decide *how* to scale the data leaks
information about the test set into model preparation, before the
model has even been trained -- a subtler, easy-to-miss form of the
same mistake as training on your test rows directly. The fix is
simple: compute the scaling parameters from `train_df` alone, then
*apply* that same fixed transform to validation and test data later --
never refit it on them.

`getMLmodel()` handles exactly this internally, via `getSplitData()`:
it fits a `caret::preProcess(..., method = "range")` scaler on the
training split only, and reuses it (via `predict()` on the scaler
object, not a fresh fit) for anything scored afterwards. You do not
need to normalize `dataset` yourself before calling `getMLmodel()` --
but you *do* need to reuse the same scaler object yourself when you
later predict on new data, which Section 7 makes concrete.

## 5. First model: Dense Neural Network

One call, every argument from Section 1's table filled in:

```{r dl-dense, results = 'hide'}
dl_result <- tryCatch({
  dl_dense <- getMLmodel(dataset = train_df, depVar = "Cab", model = "Hidden-layers",
                          optimizer = "adam", n.epochs = 300, batch.size = 64, save.model = FALSE)
  list(ok = TRUE, model = dl_dense)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))
```

What just happened: `getMLmodel()` split `train_df` again internally
(80/20, for its own training/early-stopping validation -- separate
from, and nested inside, the `train_idx`/`val_idx`/`test_idx` split
from Section 3), fit the range scaler on that inner-training slice,
built a 3-hidden-layer dense network (64/32/16 units, dropout after the
second layer), and trained it with early stopping on validation loss
(`patience = 5`). The fitted Keras model and the fitted scaler both
come back in the result:

```{r dl-dense-check}
if (dl_result$ok) {
  h <- dl_result$model$history$metrics
  cat("Epochs actually run:", length(h$loss), "of 300 requested\n")
  cat("Epoch with lowest validation loss:", which.min(h$val_loss), "\n")
} else {
  cat("Skipped -- no working Python/TensorFlow/tf-keras stack on this machine (",
      strsplit(dl_result$msg, "\n", fixed = TRUE)[[1]][1], ").\n")
}
```

## 6. Evaluate the model

**Always read the learning curve first.** A validation curve that is
still falling means training stopped before convergence; one that
rises again after a minimum means the model has started memorizing the
training data:

```{r learning-curve, fig.width=7}
if (dl_result$ok) {
  h <- dl_result$model$history$metrics
  plot(seq_along(h$loss), h$loss, type = "l", col = "#2166AC",
       ylim = range(c(h$loss, h$val_loss)), xlab = "Epoch", ylab = "Loss (MSE)",
       main = "Learning curve: training vs. validation loss")
  lines(seq_along(h$val_loss), h$val_loss, col = "#D55E00")
  legend("topright", c("Training loss", "Validation loss"), col = c("#2166AC", "#D55E00"), lty = 1)
}
```

In this run, early stopping never actually triggered -- the lowest
validation loss sits at the very last epoch of the 300-epoch budget,
so the network was still (slowly) improving when training stopped, not
demonstrably plateaued. That's an honest limitation of this demo's
epoch budget (a production run would keep training, or lower the
learning rate) -- there's real headroom left on the table below, not a
broken model to explain away.

**Predicting correctly on the held-out test set -- the one step that
is easy to get wrong.** `getMLmodel()` returns the raw Keras model and
the fitted scaler as two separate pieces, on purpose, for flexibility
-- it does not wrap them into a single self-scaling `predict()`. That
means *you* are responsible for re-applying the training scaler to any
new data before it reaches the network:

```{r dl-test-eval}
if (dl_result$ok) {
  X_test_scaled <- predict(dl_result$model$Scalar.train, dataset_dense[test_idx, band_names])
  pred_test_dl <- as.numeric(predict(dl_result$model$model, as.matrix(X_test_scaled)))
  obs_test <- dataset_dense$Cab[test_idx]
  cat("Dense DL, independent test set: R2=", round(r2_f(obs_test, pred_test_dl), 3),
      " RMSE=", round(rmse_f(obs_test, pred_test_dl), 2),
      " bias=", round(bias_f(obs_test, pred_test_dl), 2), "\n")
}
```

Skip that `predict(model$Scalar.train, ...)` line and pass raw,
unscaled reflectance straight to `model$model` instead -- an easy
mistake, since it runs without error -- and training and prediction
silently happen in two different input spaces. Section 11
(Troubleshooting) shows exactly what that looks like and why it's so
easy to miss.

```{r rf-comparison}
if (dl_result$ok) {
  fit_rf <- get.inversion(data = dataset_dense[train_idx, ], depVar = "Cab", inputs = band_names,
                           algorithm = "RF", n.samples = length(train_idx), seed = 42)
  pred_test_rf <- as.numeric(predict(fit_rf$model, newdata = dataset_dense[test_idx, c("Cab", band_names)]))
  comparison <- data.frame(
    method = c("Dense DL (getMLmodel)", "Random Forest (get.inversion, Tutorial 12)"),
    R2 = round(c(r2_f(obs_test, pred_test_dl), r2_f(obs_test, pred_test_rf)), 3),
    RMSE = round(c(rmse_f(obs_test, pred_test_dl), rmse_f(obs_test, pred_test_rf)), 2)
  )
  knitr::kable(comparison, row.names = FALSE)
}
```

```{r true-vs-predicted, fig.width=9}
if (dl_result$ok) {
  op <- par(mfrow = c(1, 2))
  plot(obs_test, pred_test_dl, pch = 19, col = "#D55E00",
       xlab = "Observed Cab", ylab = "Predicted Cab", main = "Dense DL")
  abline(0, 1, col = "grey40", lty = 2)
  plot(obs_test, pred_test_rf, pch = 19, col = "#2166AC",
       xlab = "Observed Cab", ylab = "Predicted Cab", main = "Random Forest")
  abline(0, 1, col = "grey40", lty = 2)
  par(op)
}
```

On this LUT (6000 rows, 10 Sentinel-2 bands), Random Forest keeps a
small, ordinary edge (R2 around 0.86 against the dense network's
0.82-0.84) -- a real gap between two properly-fit models, not evidence
that deep learning "doesn't work" here. That's it for the core
workflow: prepare data, split, understand normalization, train,
predict correctly, evaluate. Part II extends the same pattern to a
1D-CNN.

# Part II -- Advanced: 1D-CNN for hyperspectral inversion

## 7. Why a CNN, and when it should matter

```text
Dense network                          1D-CNN

B1 ---\                                lambda400 lambda410 lambda420 ... lambdaN
B2 ----\                                  |__________|__________|_________|
B3 -----+--> fully-connected --> Cab                 |
...    /       layers                          1D convolution
Bn ---/                                     (slides across neighbouring
                                              wavelengths, in order)
                                                      |
                                              spectral features
                                                      |
                                                     Cab
```

A dense network treats each predictor column as an independent number
-- band order carries no meaning to it. A 1D-CNN instead convolves
across the predictor sequence, so it can learn a *local spectral
shape* (an absorption feature spanning several adjacent bands) once,
as a reusable pattern, rather than having to reconstruct that
relationship independently for every band combination. That
architectural advantage needs something to bite into: **widely-spaced
multispectral bands (like Sentinel-2's 10) give a CNN little local
structure to exploit.** Contiguous hyperspectral bands are a fairer
test.

## 8. A working CNN example, on PRISMA-resolution bands

```{r cnn-prisma, results = 'hide', fig.show = 'hide'}
n_hyp <- 4000
LUT_h <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_hyp, setseed = 3))
refl_h <- t(sapply(seq_len(n_hyp), function(i) {
  foursail(inputLUT = LUT_h[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X_h <- as.data.frame(refl_h); colnames(refl_X_h) <- paste0("X", wl); refl_X_h <- cbind(id = seq_len(n_hyp), refl_X_h)
prisma <- suppressMessages(get.spectra.convolved(rfl = refl_X_h, sensor = "PRISMA", plot.spectra = FALSE))
prisma_bands <- paste0("P", seq_len(ncol(prisma) - 1))
names(prisma) <- c("id", prisma_bands)
cat("PRISMA bands:", length(prisma_bands), "\n")

set.seed(3)  # reproducible split -- matches the LUT's own setseed=3 above
idx_h <- sample(seq_len(n_hyp))
train_h <- idx_h[1:round(0.6 * n_hyp)]
test_h  <- idx_h[(round(0.8 * n_hyp) + 1):n_hyp]
dataset_cnn <- cbind(Cab = LUT_h$Cab, prisma[, prisma_bands])

# getMLmodel() (Section 6) is a single Keras training run -- fine once you
# already know it lands well, but R's keras/TensorFlow backend doesn't
# respect set.seed() for weight initialization, so any single run's
# quality varies. getMLmodel.withRetrain(n.times=3) trains 3 independent
# random restarts and returns all of them plus each one's own validation
# stats -- the same "best of n_times, keep the winner" pattern the Python
# port's own get_ml_model(n_times=3) already uses for exactly this CNN
# case, and for the same reason: without it, this section could silently
# report whichever single unlucky initialization the last knit happened
# to draw, not a representative CNN result.
cnn_result <- tryCatch({
  m <- getMLmodel.withRetrain(dataset = dataset_cnn[train_h, ], depVar = "Cab", model = "CNN",
                               optimizer = "adam", n.epochs = 300, batch.size = 64,
                               n.times = 3, save.model = FALSE)
  # m$stats has 2 rows PER restart -- an intermediate "Testing" row (model
  # state right after the first fit) and a "Val.retrain" row (model state
  # after the function's own internal second fit on a fresh 90/10 split --
  # this is the state actually saved into m$model[[i.times]]). Selecting
  # the best restart by raw row position over that combined table silently
  # mixes the two and can index m$model with a row number that belongs to
  # a *different* restart than the one being scored -- always use the
  # "Val.retrain" rows (one per restart, keyed by i.Time) instead.
  retrain_rows <- m$stats[m$stats$db == "Val.retrain", ]
  retrain_rows <- retrain_rows[order(as.numeric(retrain_rows$i.Time)), ]
  all_r2 <- as.numeric(retrain_rows$R2)
  best_i <- as.numeric(retrain_rows$i.Time)[which.max(all_r2)]
  list(ok = TRUE, model = list(model = m$model[[best_i]], Scalar.train = m$Scalar.train[[best_i]]),
       all_r2 = all_r2, best_i = best_i)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))
```

```{r cnn-restarts}
if (cnn_result$ok) {
  cat("R2 across the", length(cnn_result$all_r2), "restarts (validation, post-retrain):",
      paste(round(cnn_result$all_r2, 3), collapse = ", "),
      "-- keeping restart", cnn_result$best_i, "\n")
} else {
  cat("CNN training failed:", cnn_result$msg, "\n")
}
```

Same predict-correctly pattern as Section 6, applied to the CNN's own
scaler:

```{r cnn-eval}
if (cnn_result$ok) {
  X_test_h_scaled <- predict(cnn_result$model$Scalar.train, dataset_cnn[test_h, prisma_bands])
  pred_test_cnn <- as.numeric(predict(cnn_result$model$model,
    array_reshape(as.matrix(X_test_h_scaled), c(length(test_h), length(prisma_bands), 1))))
  obs_test_h <- dataset_cnn$Cab[test_h]
  cat("CNN (PRISMA, ", length(prisma_bands), "bands), independent test set: R2=",
      round(r2_f(obs_test_h, pred_test_cnn), 3), " RMSE=", round(rmse_f(obs_test_h, pred_test_cnn), 2), "\n")
}
```

```{r cnn-true-vs-predicted, fig.width=5, fig.height=5}
if (cnn_result$ok) {
  plot(obs_test_h, pred_test_cnn, pch = 19, col = "#009E73",
       xlab = "Observed Cab", ylab = "Predicted Cab",
       main = sprintf("1D-CNN, PRISMA (%d bands)", length(prisma_bands)))
  abline(0, 1, col = "grey40", lty = 2)
}
```

R2 around 0.85-0.90 on 234 raw PRISMA bands -- in the same range as
Section 6's 10-band dense/RF comparison, not a degradation from the
much higher input dimensionality, and a genuinely usable result (not
the point of this demo to beat RF; see Section 13 for that question).
That itself says something: on this particular synthetic Cab-from-
reflectance mapping, the CNN's spectral-neighbourhood advantage isn't
yet the dominant factor -- not because the CNN is broken, but because
there isn't a lot of local spectral structure here that a well-tuned
dense network or RF weren't already capturing from the bands directly.
Real hyperspectral problems with genuine local absorption features
(rather than this synthetic Cab mapping) are where that advantage is
more likely to show up.

Notice the per-restart R2 line above the plot: it is common for one of
the three random restarts to collapse to a degenerate, near-constant
prediction (R2 far below the others, sometimes reported as `-Inf` when
the predicted values end up with ~zero variance) -- an initialization
failure, not a data or normalization bug. This is exactly why
`getMLmodel.withRetrain(n.times = 3)` (and selecting the winner by its
own validation R2) is used here rather than a single `getMLmodel()`
call: a CNN section that trains only once, keeps whatever it gets, and
never checks per-restart R2 can silently publish a collapsed run.

# Troubleshooting

## 9. `infinite or missing values in 'x'`

Cause: `dataset` contained a column that is constant within your
sample (every row the same value) -- perhaps a trait `inputsPROSAIL`
holds fixed by design (`LMA`, `alpha`, `LIDFb`, `TypeLidf`, `hspot`,
`tts`, `psi` all have zero variance in a plain `getLUT()` draw), or an
ID column that slipped into `dataset` by accident. Section 1's
"Important" box explains why: `getMLmodel()` has no `inputs =`
argument, so *every* non-`depVar` column becomes a predictor, and its
internal range-normalization divides by `(max - min)` for each one --
zero for a constant column, producing `Inf`/`NA`. Fix: build `dataset`
from exactly `depVar` plus the columns you actually want as
predictors, as done throughout this page.

## 10. Held-out R2 looks impossibly bad, despite a normal-looking learning curve

This is the single most consequential mistake to avoid, and it does
not raise an error -- it just silently gives you a wrong number.
`getMLmodel()` trains on X range-scaled to \[0,1\] internally, and
returns the raw Keras model and the fitted scaler as two separate
pieces. If you call `predict(result$model, as.matrix(new_data))`
directly on **raw, unscaled** new data, training and prediction happen
in two different input spaces -- the network was trained *for scaled
inputs* and is now being evaluated on inputs whose distribution it has
never seen. Symptoms: R2 near zero to strongly negative on the test
set, while the training/validation loss curve looks completely
ordinary (because that curve is entirely internal to `getMLmodel()`
and never touches your mis-scaled test call). Fix: always
`predict(result$Scalar.train, new_data)` first, exactly as Sections 6
and 8 do, before passing anything to `result$model`. If you also used
`depVar.trans = TRUE`, reverse it on the output with
`ToolsRTM::getReverse.trans(preProc = result$Scalar.Ytrain, data = ...)`
(or use `ToolsRTM::getPredicts()`, which wraps both steps for you).

## 11. A `relu` output layer can get stuck at exactly 0

Not something you'll hit with `getMLmodel()` as shipped (fixed
directly in `getMLmodel.R`/`getMLmodel_withRetrain.R`), but worth
knowing if you ever adapt the architecture yourself: a `relu` unit's
gradient is exactly zero whenever its pre-activation is negative. On a
single-unit regression output with an unbounded-scale target, one
unlucky batch of negative pre-activations early in training can
permanently zero out that unit's gradient, with no way back regardless
of epoch budget -- the model then predicts a near-constant value for
every input. Use `activation = "linear"` for a regression output
layer, not `"relu"`.

# Advanced / Further experiments

The rest of this page goes beyond "how do I use `getMLmodel()`" into
open scientific questions -- when dense/CNN beat or lose to RF, and
under what feature choices. Tutorial 12 already covers systematic
algorithm comparison, and Tutorial 14 covers the full end-to-end
pipeline; treat what follows as one worked example of that kind of
investigation, not a second version of either.

## 12. Fewer, better features: does band+index selection help more than raw bands?

Sections 6 and 8 threw every band at the network -- all 10 Sentinel-2
bands, all 234 PRISMA bands -- and let the model find what matters on
its own. Tutorial 12 showed a different, more targeted approach for
Random Forest: a compact core of physically relevant bands, plus a
small, *per-trait* selection of the most correlated spectral indices
(Tutorial 09). Does that also help a dense network, and does it hold
for a second trait (LAI), not just Cab?

```{r feature-select-setup}
# Real Sentinel-2A band names (getIndicesSE2.ML()'s formulas look up
# specific bands like B8A/B11/B12 by name) -- a second, separate naming of
# the same se2a convolution from Section 2; band_names/dataset_dense above
# are untouched.
se2a_real <- se2a
names(se2a_real) <- c("id", "B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
core_bands <- c("B4", "B5", "B7", "B8A", "B11")  # a compact, physically-motivated core -- not all 10

idx_ml <- suppressMessages(getIndicesSE2.ML(df = se2a_real[, -1], sensor = "Sentinel-2a", df.data = NULL, fast.process = TRUE))

n_top <- 4  # indices kept per trait
select_indices <- function(trait) {
  cors <- sapply(names(idx_ml), function(nm) suppressWarnings(cor(idx_ml[train_idx, nm], LUT[train_idx, trait])))
  cors <- cors[is.finite(cors)]
  names(sort(abs(cors), decreasing = TRUE))[seq_len(min(n_top, length(cors)))]
}
selected <- setNames(lapply(c("Cab", "LAI"), select_indices), c("Cab", "LAI"))
knitr::kable(data.frame(trait = names(selected),
                         features = sapply(selected, function(x) paste(c(core_bands, x), collapse = ", "))),
             row.names = FALSE)
```

```{r feature-select-fit, results='hide', fig.show='hide'}
run_trait <- function(trait) {
  feat <- c(core_bands, selected[[trait]])
  df <- cbind(LUT[trait], se2a_real[core_bands], idx_ml[selected[[trait]]])

  # depVar.trans=TRUE: also range-scale the target, not just the inputs --
  # with only a handful of highly informative features the network fits
  # training data almost immediately, and on LAI's narrow raw range
  # (~0.5-7) unscaled-target MSE gradients can be unstable early in
  # training. predict() then returns range-scaled [0,1] predictions, so
  # invert them by hand using the returned scaler before computing R2.
  dl <- getMLmodel(dataset = df[train_idx, ], depVar = trait, model = "Hidden-layers",
                    optimizer = "adam", n.epochs = 300, batch.size = 64, save.model = FALSE,
                    depVar.trans = TRUE)
  X_test_scaled <- predict(dl$Scalar.train, df[test_idx, feat])
  pred_dl_scaled <- as.numeric(predict(dl$model, as.matrix(X_test_scaled)))
  y_range <- dl$Scalar.Ytrain$ranges[, trait]
  pred_dl <- pred_dl_scaled * (y_range[2] - y_range[1]) + y_range[1]

  fit_rf <- get.inversion(data = df[train_idx, ], depVar = trait, inputs = feat,
                           algorithm = "RF", n.samples = length(train_idx), seed = 42)
  pred_rf <- as.numeric(predict(fit_rf$model, newdata = df[test_idx, c(trait, feat)]))

  obs <- LUT[test_idx, trait]
  list(summary = data.frame(trait = trait, n_features = length(feat),
                             R2_dense = r2_f(obs, pred_dl), R2_RF = r2_f(obs, pred_rf)),
       obs = obs, pred_dl = pred_dl, pred_rf = pred_rf)
}
trait_runs <- setNames(lapply(c("Cab", "LAI"), run_trait), c("Cab", "LAI"))
feature_results <- do.call(rbind, lapply(trait_runs, `[[`, "summary"))
```

```{r feature-select-table}
knitr::kable(feature_results, digits = 3, row.names = FALSE)
```

```{r feature-select-plot, fig.width=9, fig.height=9}
op <- par(mfrow = c(2, 2))
for (trait in c("Cab", "LAI")) {
  r <- trait_runs[[trait]]
  lims <- range(c(r$obs, r$pred_dl, r$pred_rf))
  plot(r$obs, r$pred_dl, pch = 19, col = "#D55E00", xlim = lims, ylim = lims,
       xlab = paste("Observed", trait), ylab = paste("Predicted", trait),
       main = sprintf("%s -- Dense, 9 features (R2=%.3f)", trait, r2_f(r$obs, r$pred_dl)))
  abline(0, 1, col = "grey40", lty = 2)
  plot(r$obs, r$pred_rf, pch = 19, col = "#2166AC", xlim = lims, ylim = lims,
       xlab = paste("Observed", trait), ylab = paste("Predicted", trait),
       main = sprintf("%s -- Random Forest, 9 features (R2=%.3f)", trait, r2_f(r$obs, r$pred_rf)))
  abline(0, 1, col = "grey40", lty = 2)
}
par(op)
```

Fewer, better-chosen features help both models. On Cab, dense and RF
land close together (both around 0.83-0.86), essentially tied. On LAI
-- never even attempted with the generic 10-band set in Part I --
dense *narrowly edges out* RF, a real if small reversal of Part I's
pattern. Once dense and RF are compared on equal footing, feature
selection helps both; a properly trained dense network on a handful of
informative inputs is competitive, not fragile.

## 13. Which one should you use?

| Method | Needs training? | Typical use |
|---|---|---|
| `get.inversionOpt()` (Tutorial 11) | No -- pure search | Quick, physically-grounded retrieval when a genuinely well-populated reference LUT already exists (needs real coverage: ~100 reference spectra gave R2 near zero in Tutorial 11's own test, ~2000 gave real skill) |
| `get.inversion()` (Tutorial 12) | Yes | General-purpose retrieval; often the right default -- its `predict()` handles preprocessing internally, so there's no separate scaler to remember |
| `getMLmodel()` (this page) | Yes, more so | Large LUTs, hyperspectral input, and/or when compact, hand-picked features matter (Section 12). Competitive with RF once trained and evaluated correctly -- but the least forgiving of a scaling mistake (Section 10), since a wrong `predict()` call fails silently, not with an error |

For this package's own bundled models, `get.inversion()` remains the
lower-friction default -- not because deep learning performs worse
here, but because it is harder to silently misuse. Reach for
`getMLmodel()` when the LUT is genuinely large, the sensor is
hyperspectral, or a compact hand-chosen feature set matters, and budget
real attention for Section 10's predict-correctly discipline.

## What's next

- **Tutorial 14** -- simulate, convolve, index, and invert as one
  end-to-end pipeline, across every canopy model this package supports.
