---
title: "14. End-to-End RTM Inversion Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{14. End-to-End RTM Inversion Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(ggplot2)
library(doParallel)
library(foreach)
```

Tutorials 05-12 built the pipeline one stage at a time: LUT, parallel
simulation, sensor convolution, indices, sensitivity, and three flavours
of inversion. This page runs the **whole chain as one coherent pipeline**
-- the same shape as this package's own course scripts
(`Scripts/R/ForPROSAIL/`, `ForFoursail2/`, `ForINFORM/`, `ForSPART/`,
`ForMARMIT/`) -- and, more importantly, shows how the *same* pipeline
looks across different canopy models with only a couple of lines changed.

```text
getLUT()
   |
   v
simulate_RTM()  (fourSAIL / foursail2 / INFORM)
   |
   v
get.spectra.convolved()  (Sentinel-2A / 2B / PRISMA)
   |
   v
getIndices() / getIndicesSE2()
   |
   v
get.inversion()
   |
   v
Vegetation traits
```

## 1. One function, three canopy models

`simulate_RTM()` dispatches to `foursail()`/`foursail2()`/`inform()` based
on `canopy.model` -- `canopy.model` is a single variable, not a separate
script per model:

```{r pipeline-function}
run_pipeline <- function(canopy.model, leaf.model = "PROSPECT-D", n.samples = 500, seed = 1) {
  LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n.samples, setseed = seed))
  LUT$Cs <- 0; LUT$fqe <- 0.01; LUT$Cx <- 0
  LUT$cell.d <- 40; LUT$inter.c <- 0.045; LUT$baseline.abs <- 0.0006
  LUT$leaf.thick <- 1.6; LUT$albino.abs <- 0; LUT$lign.cell <- 2; LUT$Nitrogen <- 1
  LUT$fraction_brown <- 0.1; LUT$diss <- 0.5; LUT$Cv <- 1; LUT$Zeta <- 0
  LUT$LAIu <- 0.5; LUT$sd <- 650; LUT$cd <- 4.5; LUT$h <- 20; LUT$skyl <- 0.1

  rsoil <- rep(0.15, 2101)
  # 500 simulations per canopy model, x3 models -- parallelized the same way
  # as Tutorial 06, rather than a plain sapply(), to keep this vignette's
  # build time reasonable at this sample size.
  no_cores <- max(1, parallel::detectCores() - 2)
  cl <- makeCluster(no_cores)
  registerDoParallel(cl)
  refl_list <- foreach(i = seq_len(n.samples), .packages = "ToolsRTM") %dopar% {
    suppressMessages(simulate_RTM(inputLUT = LUT[i, ], rsoil = rsoil,
                                   leaf.model = leaf.model, canopy.model = canopy.model))$rsot
  }
  stopCluster(cl)
  refl <- do.call(rbind, refl_list)
  colnames(refl) <- paste0("R.", 400:2500)

  refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", 400:2500); refl_X <- cbind(id = seq_len(n.samples), refl_X)
  se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
  band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
  names(se2a) <- c("id", band_names)

  train_idx <- sample(seq_len(n.samples), size = round(0.7 * n.samples))
  train_df <- cbind(LUT[train_idx, ], se2a[train_idx, band_names])
  test_df  <- cbind(LUT[-train_idx, ], se2a[-train_idx, band_names])

  fit <- get.inversion(data = train_df, depVar = "LAI", inputs = band_names,
                        algorithm = "RF", n.samples = nrow(train_df), seed = 42)
  pred <- as.numeric(predict(fit$model, newdata = test_df[, c("LAI", band_names)]))
  r2 <- 1 - sum((test_df$LAI - pred)^2) / sum((test_df$LAI - mean(test_df$LAI))^2)

  list(LUT = LUT, refl = refl, obs = test_df$LAI, pred = pred, r2 = r2)
}
```

## 2. Before inverting: is INFORM's own forward model even LAI-sensitive here?

Before running any inversion, check the physics first -- same rigor as
Tutorial 10's spectral sensitivity analysis. Vary only `LAI`, hold every
crown-geometry parameter (`sd`/`cd`/`h`/`LAIu` -- stem density, crown
diameter, tree height, understory LAI) at this package's documented
typical stand values, and look at reflectance at four representative
bands:

```{r inform-forward-sensitivity, fig.width=9}
base_row <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 1, setseed = 1))
base_row$Cs <- 0; base_row$fqe <- 0.01; base_row$Cx <- 0
base_row$cell.d <- 40; base_row$inter.c <- 0.045; base_row$baseline.abs <- 0.0006
base_row$leaf.thick <- 1.6; base_row$albino.abs <- 0; base_row$lign.cell <- 2; base_row$Nitrogen <- 1
base_row$fraction_brown <- 0.1; base_row$diss <- 0.5; base_row$Cv <- 1; base_row$Zeta <- 0
base_row$LAIu <- 0.5; base_row$sd <- 650; base_row$cd <- 4.5; base_row$h <- 20; base_row$skyl <- 0.1
rsoil <- rep(0.15, 2101)

lai_seq <- c(0.5, 1, 2, 3, 4, 5, 6, 7)
bands <- c(red = 665, rededge = 705, nir = 800, swir = 1650)
sens_inform <- sapply(lai_seq, function(lai) {
  r <- base_row; r$LAI <- lai
  suppressMessages(inform(inputLUT = r, rsoil = rsoil, LeafModel = "PROSPECT-D"))[bands - 400 + 1]
})
sens_foursail <- sapply(lai_seq, function(lai) {
  r <- base_row; r$LAI <- lai
  foursail(inputLUT = r, rsoil = rsoil, LeafModel = "PROSPECT-D")$rsot[bands - 400 + 1]
})
rownames(sens_inform) <- rownames(sens_foursail) <- names(bands)

op <- par(mfrow = c(1, 2))
matplot(lai_seq, t(sens_inform), type = "o", pch = 19, lty = 1, col = 1:4,
        xlab = "LAI", ylab = "Reflectance", main = "INFORM: reflectance vs. LAI\n(crown geometry fixed)",
        ylim = range(sens_inform, sens_foursail))
legend("right", rownames(sens_inform), col = 1:4, pch = 19, lty = 1, cex = 0.7)
matplot(lai_seq, t(sens_foursail), type = "o", pch = 19, lty = 1, col = 1:4,
        xlab = "LAI", ylab = "Reflectance", main = "fourSAIL: reflectance vs. LAI",
        ylim = range(sens_inform, sens_foursail))
par(op)
```

```{r inform-forward-sensitivity-table}
cat("INFORM range (max-min) across LAI 0.5-7:\n")
print(round(apply(sens_inform, 1, function(x) max(x) - min(x)), 4))
cat("fourSAIL range (max-min) across LAI 0.5-7:\n")
print(round(apply(sens_foursail, 1, function(x) max(x) - min(x)), 4))
```

That's the real, verified starting point: **under this fixed-geometry
parameterization, INFORM's own forward-simulated reflectance is far
less sensitive to LAI than fourSAIL's is** -- `red` and `rededge` are
nearly flat across the whole LAI range (INFORM's total swing is roughly
1/20th to 1/40th of fourSAIL's at those bands), and even `nir`/`swir`
move noticeably less. This isn't an inversion-algorithm artifact; it's
visible directly in the forward simulations, before any ML model is
involved. Forest canopy reflectance in INFORM is a function of *both*
LAI and crown geometry together (via its gap-fraction/hotspot terms,
`get.SCOPE`-style stand-structure formulation); with geometry pinned at
one fixed value, LAI alone has comparatively little room left to move
the signal.

## 3. Does letting crown geometry vary help, or hurt?

The natural next question: if fixed geometry caps how much LAI alone
can influence the signal, does allowing `sd`/`cd`/`h`/`LAIu` to vary
across the LUT (closer to how a real forest actually varies) improve
LAI retrieval -- or does it just add more ways for different
parameter combinations to produce the same reflectance (equifinality)?
Three progressively more complex LUTs, 300 rows each, same RF
inversion setup as Section 2 above:

```{r progressive-complexity, results='hide'}
n.samples <- 300
run_case <- function(vary_geom = FALSE, vary_extra = FALSE, seed = 1) {
  LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n.samples, setseed = seed))
  LUT$Cs <- 0; LUT$fqe <- 0.01; LUT$Cx <- 0
  LUT$cell.d <- 40; LUT$inter.c <- 0.045; LUT$baseline.abs <- 0.0006
  LUT$leaf.thick <- 1.6; LUT$albino.abs <- 0; LUT$lign.cell <- 2; LUT$Nitrogen <- 1
  LUT$fraction_brown <- 0.1; LUT$diss <- 0.5; LUT$Cv <- 1; LUT$Zeta <- 0
  LUT$LAIu <- 0.5; LUT$sd <- 650; LUT$cd <- 4.5; LUT$h <- 20; LUT$skyl <- 0.1

  set.seed(seed + 100)
  if (vary_geom) {
    LUT$sd <- runif(n.samples, 200, 1000); LUT$cd <- runif(n.samples, 2, 7)
    LUT$h  <- runif(n.samples, 8, 25); LUT$LAIu <- runif(n.samples, 0, 0.8)
  }
  if (vary_extra) {
    LUT$Car <- runif(n.samples, 4, 15); LUT$Anth <- runif(n.samples, 0, 2); LUT$Cbrown <- runif(n.samples, 0, 0.3)
  }

  refl <- t(sapply(seq_len(n.samples), function(i) {
    suppressMessages(inform(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-D"))
  }))
  refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", 400:2500); refl_X <- cbind(id = seq_len(n.samples), refl_X)
  se2a <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
  band_names <- paste0("B", seq_along(as.numeric(names(se2a)[-1])))
  names(se2a) <- c("id", band_names)

  set.seed(seed)
  train_idx <- sample(seq_len(n.samples), size = round(0.7 * n.samples))
  train_df <- cbind(LUT[train_idx, ], se2a[train_idx, band_names])
  test_df  <- cbind(LUT[-train_idx, ], se2a[-train_idx, band_names])
  fit <- get.inversion(data = train_df, depVar = "LAI", inputs = band_names,
                        algorithm = "RF", n.samples = nrow(train_df), seed = 42)
  pred <- as.numeric(predict(fit$model, newdata = test_df[, c("LAI", band_names)]))
  1 - sum((test_df$LAI - pred)^2) / sum((test_df$LAI - mean(test_df$LAI))^2)
}
r2_a <- run_case(vary_geom = FALSE, vary_extra = FALSE)  # A: LAI varies, structure fixed
r2_b <- run_case(vary_geom = TRUE,  vary_extra = FALSE)  # B: LAI + crown geometry vary
r2_c <- run_case(vary_geom = TRUE,  vary_extra = TRUE)   # C: B + understory/leaf params vary
```

```{r progressive-table}
data.frame(
  case = c("A: LAI varies, geometry fixed", "B: LAI + crown geometry vary", "C: B + understory/leaf vary"),
  LAI_R2 = round(c(r2_a, r2_b, r2_c), 3)
) |> knitr::kable()
```

The answer is unambiguous: **letting crown geometry vary makes LAI
retrieval *worse*, not better** (R² drops from A to B), and adding
further understory/leaf variability on top (C) doesn't recover it
either. This is equifinality in action: once `sd`/`cd`/`h`/`LAIu` are
free to vary too, many different (LAI, geometry) combinations produce
similar reflectance, and the RF model has no way to disentangle which
one actually happened. **Fixing crown geometry isn't the cause of
INFORM's weaker LAI retrieval** -- if anything, it's the setting where
retrieval is *least* bad, by isolating the one real effect (Section 2)
that INFORM's own physics does show. The full explanation has two real
parts: INFORM's forward reflectance is objectively less LAI-sensitive
than fourSAIL/foursail2's under this parameterization (Section 2), and
whatever weak signal LAI does have gets further diluted the moment
realistic structural variability is allowed (this section) -- more
structural representation, more parameter interaction, not "INFORM is
worse."

## 4. Why is LAI retrieval weaker with INFORM in this experiment?

```{r run-all, results='hide'}
set.seed(1)
res_foursail  <- run_pipeline("fourSAIL")
set.seed(1)
res_foursail2 <- run_pipeline("foursail2")
set.seed(1)
res_inform    <- run_pipeline("INFORM")
```

```{r results-table}
results <- data.frame(
  canopy_model = c("fourSAIL", "foursail2", "INFORM"),
  LAI_R2 = c(res_foursail$r2, res_foursail2$r2, res_inform$r2)
)
knitr::kable(results, digits = 3)
```

```{r inform-vs-foursail2, fig.width=9}
op <- par(mfrow = c(1, 2))
plot(res_foursail2$obs, res_foursail2$pred, pch = 19, col = "#009E73",
     xlab = "Observed LAI", ylab = "Predicted LAI", main = paste("foursail2, R2 =", round(res_foursail2$r2, 2)))
abline(0, 1, lty = 2, col = "grey40")
plot(res_inform$obs, res_inform$pred, pch = 19, col = "#D55E00",
     xlab = "Observed LAI", ylab = "Predicted LAI", main = paste("INFORM, R2 =", round(res_inform$r2, 2)))
abline(0, 1, lty = 2, col = "grey40")
par(op)
```

**Under the LUT configuration used in this example, LAI retrieval with
INFORM is substantially weaker than with fourSAIL and foursail2. This
result is specific to the parameterization and inversion experiment
used here and should not be interpreted as a general limitation of
INFORM.** Sections 2-3's sensitivity experiments show how LAI interacts
with crown geometry, understory/background, and other structural
parameters, illustrating the importance of parameter identifiability
and equifinality when moving from homogeneous-canopy models
(fourSAIL/foursail2) to forest RTMs with explicit crown structure
(INFORM). A real forest-inversion study would need either a much larger
LUT specifically designed to break these parameter correlations, prior
constraints on crown geometry from independent data (e.g. LiDAR-derived
stand structure), or accepting wider uncertainty on LAI specifically
when using a structurally richer model like INFORM.

## 3. SPART: structurally different, same downstream pipeline

`SPART()` (Tutorial 03) already outputs at sensor bands directly -- there
is no separate "simulate native, then convolve" step, so its own version
of this pipeline runs `SPART()` once per LUT row instead of
`simulate_RTM()` + `get.spectra.convolved()`, then feeds the result into
the exact same `get.inversion()` call:

```{r spart-note, eval = FALSE}
# ForSPART/1-simulate_LUT.R's version of this pipeline, conceptually:
sim_i <- SPART(inputLUT = LUT[i, ], CanopyModel = "fourSAIL", LeafModel = "PROSPECT-PRO",
                sensor.i = ToolsRTM::Sentinel2A.MSI, rsoil = NULL, get.plots = FALSE)
toc_i <- sim_i$output$rfl.toc.BRDF  # already at Sentinel-2A bands -- no convolution step needed
```

## 4. MARMIT: the odd one out, on purpose

`ForMARMIT` in this package's course scripts has no leaf or canopy model
at all -- no `Cab`/`LAI`/`EWT` to invert. Its target trait is `SMC`
(gravimetric soil moisture, via `get.marmit.rsoil()`'s own physics), and
it inverts almost perfectly (R² typically 0.9+) since soil reflectance's
response to moisture is close to deterministic physics, unlike the
noisier proxy relationship trait retrieval from canopy reflectance usually
involves. Not run here (see Tutorial 03's soil-contribution section and
the `marmit-soil-in-canopy` article for the underlying MARMIT physics).

## What's next

- **Tutorial 15** -- the same inversion framework applied to real
  Sentinel-2 imagery retrieved via STAC, including a genuine spatial map.
