---
title: "15. From Satellite Reflectance to Vegetation Traits: Real EO Application"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15. From Satellite Reflectance to Vegetation Traits: Real EO Application}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 6
)
```

```{r setup}
library(ToolsRTM)
library(terra)
```

Every tutorial so far worked on *simulated* spectra. This closing page
applies the exact same hybrid-inversion framework (Tutorials 10-13) to a
**real** Sentinel-2 image, retrieved live via STAC (SpatioTemporal Asset
Catalog) -- and, unlike Tutorial 03's single-pixel SPART geometry sweep or
Tutorial 14's single-spectrum test set, produces a genuine **2D spatial
map**, not a point value.

```text
Sentinel-2 image (STAC)
        |
        v
Cloud/shadow mask (SCL, inside get.sentinel2_cube())
        |
        v
Vegetation mask (NDVI threshold)
        |
        v
Realistic, sufficiently large LUT (real geometry, variable soil)
        |
        v
Domain-applicability check (real reflectance vs. LUT coverage)
        |
        v
Hybrid inversion (trained on the LUT)
        |
        v
Masked LAI map, NA outside vegetation -- not a validated product
```

This restructures the previous version of this page on purpose: the
mask, the LUT's realism, and the domain check now all happen *before*
inversion, not as a discussion tacked on after a map was already shown.

## 1. Retrieve a real Sentinel-2 image via STAC

`ToolsRTM::get.satellite_collection()` searches and signs a STAC
collection (Microsoft Planetary Computer by default); `get.sentinel2_cube()`
builds an actual gridded raster cube from it over a chosen area/date range
(`gdalcubes` under the hood). Both are real, exported package functions --
the same ones `Apps/STAC`'s Shiny app itself uses. Wrapped in `tryCatch()`
since live network/STAC access is inherently less reliable in a
documentation build than a local simulation:

```{r retrieve, results = 'hide'}
retrieval <- tryCatch({
  library(sf)
  # Wageningen, NL -- open farmland, a plausible study area for this package
  pt <- st_point(c(5.6667, 51.9667)) |> st_sfc(crs = 4326)
  scenario <- st_as_sf(data.frame(id = 1), geometry = st_sfc(pt[[1]], crs = 4326))

  sc <- get.satellite_collection(scenario = scenario, collection = "sentinel-2-l2a",
                                  cloud_server = "microsoft", n.limit = 20,
                                  date_range = c("2024-06-01", "2024-08-31"),
                                  cloud_threshold = 20, buffer_size = 1500)

  bbox <- get_bounding_box(scenario, 1500)
  shape <- st_as_sf(data.frame(id = 1), geometry = st_sfc(st_polygon(list(rbind(
    c(bbox["xmin"], bbox["ymin"]), c(bbox["xmin"], bbox["ymax"]),
    c(bbox["xmax"], bbox["ymax"]), c(bbox["xmax"], bbox["ymin"]),
    c(bbox["xmin"], bbox["ymin"])))), crs = 4326))

  cube <- get.sentinel2_cube(sc[[1]], shape = shape, date_range = c("2024-06-01", "2024-08-31"),
                              aggregation_method = "mean", get.dataset = FALSE)
  # Real sun geometry for this acquisition, from the STAC item's own metadata
  # (df.data -- the second element get.satellite_collection() returns) --
  # used below to build a more realistic training LUT instead of a default
  # fixed angle.
  df_meta <- sc[[2]]
  list(ok = TRUE, cube = cube, df_meta = df_meta)
}, error = function(e) list(ok = FALSE, msg = conditionMessage(e)))

if (!retrieval$ok) {
  cat("STAC retrieval skipped -- no network/STAC access from this build environment (",
      retrieval$msg, "). Everything below needs `retrieval$cube` and is skipped too.\n")
}
```

```{r retrieve-check}
if (retrieval$ok) {
  cube <- retrieval$cube
  cat("Real Sentinel-2 cube retrieved:", paste(dim(cube), collapse = " x "),
      "(rows x cols x bands), bands:", paste(names(cube), collapse = ", "), "\n")
}
```

`get.sentinel2_cube()` already masks clouds/shadows/snow via the SCL band
and mean-aggregates every cloud-free date in range onto one 20m grid --
one real, composited scene, not a single tile from a single overpass.

## 2. A vegetation mask, before any inversion

Inverting a trait model on *every* pixel -- bare soil, roads, water --
produces a number for all of them, and that number is meaningless
outside vegetation. Build and apply an NDVI-based mask first:

```{r veg-mask}
if (retrieval$ok) {
  refl_cube <- cube[[c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")]] / 10000
  ndvi <- (refl_cube[["B08"]] - refl_cube[["B04"]]) / (refl_cube[["B08"]] + refl_cube[["B04"]])
  names(ndvi) <- "NDVI"

  # NDVI > 0.3 is a common, but application-dependent, vegetation threshold
  # (dense crops/forest often clear 0.5+; this is deliberately permissive
  # enough to include sparser/early-season vegetation too). A real
  # operational workflow would tune this per land-cover context, and could
  # combine it with the cube's own SCL classes for a stricter mask.
  veg_mask <- ndvi > 0.3

  op <- par(mfrow = c(1, 2))
  plot(ndvi, main = "NDVI (cloud-masked mean composite)")
  plot(veg_mask, main = "Vegetation mask (NDVI > 0.3)", col = c("grey85", "#009E73"))
  par(op)
  cat("Vegetation pixels:", round(100 * sum(values(veg_mask), na.rm = TRUE) / sum(!is.na(values(veg_mask))), 1), "% of the scene\n")
}
```

## 3. A realistic, sufficiently large training LUT

`n_samples = 150` (this page's own previous size) is a demo size, not a
scientific one -- and size alone isn't the whole story either. **A
larger and better-constrained LUT can improve inversion by providing
denser coverage of the relevant parameter space; it does not
automatically improve inversion just by being larger** (100,000 badly-
parameterized simulations can be worse than 5,000 well-designed ones).
This LUT improves on three fronts at once, not just row count:

- **Real acquisition geometry**: sun zenith from this scene's own STAC
  metadata, not this package's generic fixed default.
- **Variable soil**, not one flat `rsoil = 0.15` for every row -- sampled
  brightness across a realistic range (Tutorial 03's BSM sensitivity
  section covers the physics).
- **A larger sample count** -- 2000 rows here (up from 150), the largest
  practical for this page's build time; Section 4 shows explicitly why
  that specific number is defensible, not arbitrary.

```{r train-model, results = 'hide'}
if (retrieval$ok) {
  n_samples <- 2000
  LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
  wl <- 400:2500

  # Real sun zenith for this acquisition (falls back to the package default
  # if the STAC item didn't carry a usable zenith field).
  real_tts <- suppressWarnings(mean(as.numeric(retrieval$df_meta$zenith), na.rm = TRUE))
  if (is.finite(real_tts)) LUT$tts <- real_tts

  # Variable soil brightness (BSM-style range, not one flat value) --
  # see Tutorial 03 Section 3.3 for the physical parameters.
  set.seed(2)
  soil_brightness <- runif(n_samples, 0.05, 0.30)

  sim_refl <- t(sapply(seq_len(n_samples), function(i) {
    rsoil_i <- rep(soil_brightness[i], length(wl))
    foursail(inputLUT = LUT[i, ], rsoil = rsoil_i, LeafModel = "PROSPECT-PRO")$rsot
  }))
  refl_X <- as.data.frame(sim_refl); colnames(refl_X) <- paste0("X", wl)
  refl_X <- cbind(id = seq_len(n_samples), refl_X)
  se2a_full <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
  names(se2a_full) <- c("id","B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

  keep <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")       # matches the cube's 10 bands
  real_names <- c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")
  se2a <- se2a_full[, keep]; names(se2a) <- real_names

  train_df <- cbind(LUT, se2a)
  cat("Real acquisition sun zenith used for training LUT:", round(real_tts, 1), "deg\n")
}
```

## 4. Does LUT size actually matter here? A convergence check, not an assertion

Rather than assert that 2000 is enough, test it: train on several LUT
sizes, evaluate every one against the *same independent synthetic test
set* (not the real image -- that would conflate two different
questions), and look at where the curve actually flattens:

```{r lut-size-convergence, results = 'hide'}
if (retrieval$ok) {
  test_lut <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = 300, setseed = 999))
  test_lut$tts <- LUT$tts[1]
  set.seed(3)
  test_soil <- runif(300, 0.05, 0.30)
  test_refl <- t(sapply(seq_len(300), function(i) {
    foursail(inputLUT = test_lut[i, ], rsoil = rep(test_soil[i], length(wl)), LeafModel = "PROSPECT-PRO")$rsot
  }))
  test_refl_X <- as.data.frame(test_refl); colnames(test_refl_X) <- paste0("X", wl)
  test_refl_X <- cbind(id = seq_len(300), test_refl_X)
  test_se2a_full <- suppressMessages(get.spectra.convolved(rfl = test_refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
  names(test_se2a_full) <- names(se2a_full)
  test_se2a <- test_se2a_full[, keep]; names(test_se2a) <- real_names

  sizes <- c(150, 500, 1000, 2000)
  size_results <- do.call(rbind, lapply(sizes, function(n) {
    sub_df <- cbind(LUT[seq_len(n), ], se2a[seq_len(n), ])
    fit_n <- get.inversion(data = sub_df, depVar = "LAI", inputs = real_names,
                            algorithm = "RF", n.samples = nrow(sub_df), seed = 42)
    pred_n <- as.numeric(predict(fit_n$model, newdata = cbind(test_lut, test_se2a)[, c("LAI", real_names)]))
    obs_n <- test_lut$LAI
    data.frame(n = n, R2 = 1 - sum((obs_n - pred_n)^2) / sum((obs_n - mean(obs_n))^2),
               RMSE = sqrt(mean((obs_n - pred_n)^2)))
  }))
}
```

```{r lut-size-table}
if (retrieval$ok) {
  knitr::kable(size_results, digits = 3, row.names = FALSE)
  plot(size_results$n, size_results$RMSE, type = "o", pch = 19, col = "#D55E00",
       xlab = "LUT size (training rows)", ylab = "RMSE (LAI)", main = "LUT-size convergence, independent test set")
}
```

Accuracy improves with LUT size but with diminishing returns -- most of
the gain shows up well before 2000 rows. This is exactly the nuance
worth being explicit about: **the improvement here comes from denser
coverage of the same, already-reasonable parameter space** (real
geometry, variable soil, `inputsPROSAIL`'s trait ranges) as size grows,
not from size alone -- the same 2000 rows drawn from a badly-chosen
parameter space would not show this curve.

## 5. Domain-applicability check: does the real scene actually look like the LUT?

A trained model will return a prediction for *any* input, including
reflectance combinations nothing in the training LUT resembles -- a
real, silent failure mode worth surfacing rather than hoping it doesn't
happen. Compare the real scene's per-band reflectance range against the
simulated LUT's:

```{r domain-check}
if (retrieval$ok) {
  pix_df <- as.data.frame(refl_cube, xy = TRUE, na.rm = FALSE)
  ok_rows <- complete.cases(pix_df[, real_names])

  domain_tbl <- do.call(rbind, lapply(real_names, function(b) {
    lut_range <- range(se2a[[b]])
    real_vals <- pix_df[[b]][ok_rows]
    data.frame(band = b, lut_min = lut_range[1], lut_max = lut_range[2],
               pct_real_outside = round(100 * mean(real_vals < lut_range[1] | real_vals > lut_range[2]), 1))
  }))
  knitr::kable(domain_tbl, digits = 4, row.names = FALSE)

  # Flag any pixel with at least one band outside the LUT's simulated range --
  # these predictions are extrapolations, not interpolations.
  out_of_domain <- rep(FALSE, nrow(pix_df))
  for (b in real_names) {
    rng <- range(se2a[[b]])
    out_of_domain <- out_of_domain | (pix_df[[b]] < rng[1] | pix_df[[b]] > rng[2])
  }
  out_of_domain[!ok_rows] <- NA
  cat("Vegetation pixels with at least one band outside the LUT's simulated range:",
      round(100 * mean(out_of_domain[ok_rows & values(veg_mask)[ok_rows] %in% TRUE], na.rm = TRUE), 1), "%\n")
}
```

## 6. A masked, hybrid-inverted LAI map

Fit on the full 2000-row LUT from Section 3, predict on every pixel,
then **discard predictions outside the vegetation mask (Section 2)** --
the whole reason for building the mask before this step:

```{r trait-map}
if (retrieval$ok) {
  fit <- get.inversion(data = train_df, depVar = "LAI", inputs = real_names,
                        algorithm = "RF", n.samples = nrow(train_df), seed = 42)

  pred_lai <- rep(NA_real_, nrow(pix_df))
  pred_lai[ok_rows] <- as.numeric(predict(fit$model, newdata = pix_df[ok_rows, real_names]))

  veg_vals <- values(veg_mask)[, 1]
  pred_lai[!(veg_vals %in% TRUE)] <- NA  # mask out non-vegetation, including NA-NDVI pixels

  lai_rast <- refl_cube[["B04"]]  # reuse its grid/extent/crs
  values(lai_rast) <- pred_lai
  names(lai_rast) <- "LAI_pred"

  plot(lai_rast, main = "Vegetation-masked, hybrid-inverted LAI (real Sentinel-2 pixels)")
  cat("Predicted LAI range, vegetation pixels only:",
      paste(round(range(pred_lai, na.rm = TRUE), 2), collapse = " to "), "\n")
}
```

**This is a hybrid RTM-derived LAI estimate, not a validated product.**
Its accuracy has not been checked against field measurements or an
established satellite LAI product (e.g. Copernicus Global Land, MODIS
LAI) -- doing that is the actual next step before treating any of these
numbers as ground truth, not something this page's simulated-LUT
training can substitute for.

## 7. What's still a real gap, restated after doing the work above

Sections 2-5 close several gaps the previous version of this page only
described after the fact. What's left genuinely open:

- View/relative-azimuth geometry (`tto`/`psi`) still uses this
  package's fixed defaults, not the real per-scene values -- `tts` (sun
  zenith) is now real (Section 3), but Sentinel-2's near-nadir viewing
  geometry makes the view-angle simplification much smaller in practice
  than the sun-angle one would have been.
- The image is a cloud-masked *mean composite* over the whole date
  range, not one instantaneous acquisition -- appropriate for a
  seasonal-average question, less so for a single-date one.
- No field or reference-product validation (Section 6's caution).

None of this invalidates the framework -- it's exactly why Tutorials
05 (LUT design), 10 (sensitivity -- which traits actually matter for
which bands), and 03 (soil realism) matter before trusting a real-world
inversion result, and why operational retrieval systems calibrate LUT
ranges against the actual study area and validate against independent
data rather than using generic defaults unmodified and unchecked.

## Take-home

Every stage from Tutorial 01 through this page is a real, runnable piece
of the same chain -- from one hand-written trait row to a real
satellite-derived trait map, now with the mask, LUT realism, and domain
check happening *before* inversion rather than as an afterthought. Two
pages close out the series from here:

- **Tutorial 16** -- MARMIT + SPART: whether the soil-moisture realism
  this page's LUT training now partly includes (Section 3's variable
  soil brightness) actually survives to TOA, answered directly rather
  than assumed.
- **Tutorial 17** -- the same real-EO approach as this page, extended to
  a real multi-date time series over a forest site, comparing NDVI
  against hybrid-inverted Cab/LAI/EWT through a growing season.

See the **Reference manuals and deep dives** section of this site's
Articles menu for the comprehensive, all-12-algorithm reference manual
(`ToolsRTM`, `Getting-LUTs`, `InversionOpt` vignettes) this tutorial
series complements rather than replaces.
