---
title: "17. Monitoring a Forest Site Through Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{17. Monitoring a Forest Site Through Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
library(ggplot2)
```

Tutorial 15 retrieved ONE real Sentinel-2 scene and produced a spatial
trait map. This closing page retrieves a real, **monthly** time series
across a growing season, over a real forest site, and compares plain
spectral indices (NDVI, NDRE, NDMI) against physically-retrieved traits
(Cab, LAI, EWT) from the hybrid-inversion framework -- using two
structurally different canopy models (fourSAIL, INFORM) trained on the
same real pixels, specifically to check whether the retrieved trajectory
is trustworthy or an artifact of model choice.

```text
Sentinel-2 -> cloud/SCL mask (inside get.sentinel2_cube) -> forest mask (NDVI)
        |
        v
Robust per-date stats (median + IQR) over masked forest pixels
        |
        v
   +----+----+----+
   |    |    |    |
   v    v    v    v
NDVI/NDRE/NDMI   Cab/LAI/EWT, fourSAIL-trained AND INFORM-trained
   |    |    |    |
   +----+----+----+
        |
        v
Domain check + compare trajectories -- including where the two models disagree
```

## 1. Site: Speulderbos, NL

Speulderbos is a long-term Douglas-fir/beech forest research site in the
Netherlands (Gelderland) -- a real, physically-meaningful location for a
seasonal forest time series.

```{r site}
library(sf); library(terra)
pt <- st_point(c(5.6900, 52.2500)) |> st_sfc(crs = 4326)
scenario <- st_as_sf(data.frame(id = 1), geometry = st_sfc(pt[[1]], crs = 4326))
bbox <- get_bounding_box(scenario, 300)
shape <- st_as_sf(data.frame(id = 1), geometry = st_sfc(st_polygon(list(rbind(
  c(bbox["xmin"], bbox["ymin"]), c(bbox["xmin"], bbox["ymax"]),
  c(bbox["xmax"], bbox["ymax"]), c(bbox["xmax"], bbox["ymin"]),
  c(bbox["xmin"], bbox["ymin"])))), crs = 4326))
real_names <- c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")
```

## 2. Train fourSAIL AND INFORM models on a larger, forest-realistic LUT

Two real improvements over an earlier version of this page: `n_samples`
is 1500 (up from 200), and soil is sampled across a realistic brightness
range rather than one fixed `rsoil = 0.15` -- both matching Tutorial
15's LUT-realism argument. `LAI` is constrained to 2-7 (plausible for a
closed forest canopy, not the package's generic default range). A
second model, trained on `inform()` simulations instead of `foursail()`
(forest crown geometry -- `sd`/`cd`/`h`/`LAIu` -- varying too, per
Tutorial 14's equifinality finding), lets Section 5 check whether the
seasonal trajectory below actually depends on which canopy model
produced it:

```{r train-both, results = 'hide'}
n_samples <- 1500
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
LUT$LAI <- runif(n_samples, 2, 7)
wl <- 400:2500
set.seed(11)
soil_b <- runif(n_samples, 0.05, 0.30)
sim_refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rep(soil_b[i], length(wl)), LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(sim_refl); colnames(refl_X) <- paste0("X", wl)
refl_X <- cbind(id = seq_len(n_samples), refl_X)
se2a_full <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_full) <- c("id","B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")
keep <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")
se2a <- se2a_full[, keep]; names(se2a) <- real_names
train_df <- cbind(LUT, se2a)
fits_sail <- lapply(c("Cab", "LAI", "EWT"), function(trait) {
  get.inversion(data = train_df, depVar = trait, inputs = real_names,
                algorithm = "RF", n.samples = nrow(train_df), seed = 42)
})
names(fits_sail) <- c("Cab", "LAI", "EWT")

n_inform <- 800
LUT_i <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_inform, setseed = 2))
LUT_i$Cs <- 0; LUT_i$fqe <- 0.01; LUT_i$Cx <- 0
LUT_i$cell.d <- 40; LUT_i$inter.c <- 0.045; LUT_i$baseline.abs <- 0.0006
LUT_i$leaf.thick <- 1.6; LUT_i$albino.abs <- 0; LUT_i$lign.cell <- 2; LUT_i$Nitrogen <- 1
LUT_i$fraction_brown <- 0.1; LUT_i$diss <- 0.5; LUT_i$Cv <- 1; LUT_i$Zeta <- 0
LUT_i$LAI <- runif(n_inform, 2, 7)
set.seed(12)
LUT_i$LAIu <- runif(n_inform, 0, 0.8); LUT_i$sd <- runif(n_inform, 400, 900)
LUT_i$cd <- runif(n_inform, 3, 6); LUT_i$h <- runif(n_inform, 15, 30); LUT_i$skyl <- 0.1
soil_bi <- runif(n_inform, 0.05, 0.30)
sim_refl_i <- t(sapply(seq_len(n_inform), function(i) {
  suppressMessages(inform(inputLUT = LUT_i[i, ], rsoil = rep(soil_bi[i], length(wl)), LeafModel = "PROSPECT-D"))
}))
refl_X_i <- as.data.frame(sim_refl_i); colnames(refl_X_i) <- paste0("X", wl)
refl_X_i <- cbind(id = seq_len(n_inform), refl_X_i)
se2a_full_i <- suppressMessages(get.spectra.convolved(rfl = refl_X_i, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_full_i) <- names(se2a_full)
se2a_i <- se2a_full_i[, keep]; names(se2a_i) <- real_names
train_df_i <- cbind(LUT_i, se2a_i)
fits_inform <- lapply(c("Cab", "LAI", "EWT"), function(trait) {
  get.inversion(data = train_df_i, depVar = trait, inputs = real_names,
                algorithm = "RF", n.samples = nrow(train_df_i), seed = 42)
})
names(fits_inform) <- c("Cab", "LAI", "EWT")
```

## 3. Retrieve a real, monthly time series -- forest-masked, not a raw scene mean

Monthly windows March-November (9 dates, up from the previous 5),
wrapped in `tryCatch()` per date since any individual window can fail
for ordinary reasons. `get.sentinel2_cube()` already applies Sentinel-2's
SCL cloud/shadow/snow mask internally (Tutorial 15); this section adds a
**second, vegetation-specific mask** (`NDVI > 0.5`, denser than Tutorial
15's 0.3 threshold since a closed forest should read distinctly greener
than farmland) so that any road, clearing, or water within the 300m
buffer doesn't dilute the forest signal -- then computes **median and
IQR** across the masked forest pixels for every date, not a single
scene-wide mean:

```{r timeseries, results = 'hide'}
windows <- list(c("2024-03-01","2024-03-31"), c("2024-04-01","2024-04-30"),
                 c("2024-05-01","2024-05-31"), c("2024-06-01","2024-06-30"),
                 c("2024-07-01","2024-07-31"), c("2024-08-01","2024-08-31"),
                 c("2024-09-01","2024-09-30"), c("2024-10-01","2024-10-31"),
                 c("2024-11-01","2024-11-30"))

get_one_date <- function(w) {
  tryCatch({
    sc <- get.satellite_collection(scenario = scenario, collection = "sentinel-2-l2a",
                                    cloud_server = "microsoft", n.limit = 20,
                                    date_range = w, cloud_threshold = 40, buffer_size = 300)
    if (is.null(sc[[1]])) stop("no cloud-free items this window")
    cube <- get.sentinel2_cube(sc[[1]], shape = shape, date_range = w,
                                aggregation_method = "mean", get.dataset = FALSE)
    refl <- cube[[real_names]] / 10000

    ndvi <- (refl[["B08"]] - refl[["B04"]]) / (refl[["B08"]] + refl[["B04"]])
    forest_mask <- ndvi > 0.5  # dense-canopy mask -- excludes roads/clearings/water in the buffer

    pix_df <- as.data.frame(refl, na.rm = FALSE)
    mvals <- values(forest_mask)[, 1]
    ok_rows <- stats::complete.cases(pix_df) & (mvals %in% TRUE)
    if (sum(ok_rows) < 5) stop("fewer than 5 forest pixels this window")
    pix_ok <- pix_df[ok_rows, ]

    ndvi_px <- (pix_ok[["B08"]] - pix_ok[["B04"]]) / (pix_ok[["B08"]] + pix_ok[["B04"]])
    ndre_px <- (pix_ok[["B08"]] - pix_ok[["B05"]]) / (pix_ok[["B08"]] + pix_ok[["B05"]])
    ndmi_px <- (pix_ok[["B08"]] - pix_ok[["B11"]]) / (pix_ok[["B08"]] + pix_ok[["B11"]])

    Cab_sail <- as.numeric(predict(fits_sail$Cab$model, newdata = pix_ok[, real_names]))
    LAI_sail <- as.numeric(predict(fits_sail$LAI$model, newdata = pix_ok[, real_names]))
    EWT_sail <- as.numeric(predict(fits_sail$EWT$model, newdata = pix_ok[, real_names]))
    Cab_inform <- as.numeric(predict(fits_inform$Cab$model, newdata = pix_ok[, real_names]))
    LAI_inform <- as.numeric(predict(fits_inform$LAI$model, newdata = pix_ok[, real_names]))

    # Domain check (Tutorial 15 Section 5): fraction of forest pixels with at
    # least one band outside the fourSAIL LUT's own simulated reflectance range.
    out_of_domain <- rep(FALSE, nrow(pix_ok))
    for (b in real_names) {
      rng <- range(se2a[[b]])
      out_of_domain <- out_of_domain | (pix_ok[[b]] < rng[1] | pix_ok[[b]] > rng[2])
    }

    data.frame(date = as.Date(w[1]), n_forest_px = sum(ok_rows),
               NDVI = median(ndvi_px), NDVI_iqr = IQR(ndvi_px),
               NDRE = median(ndre_px), NDMI = median(ndmi_px),
               Cab_fourSAIL = median(Cab_sail), Cab_fourSAIL_iqr = IQR(Cab_sail),
               LAI_fourSAIL = median(LAI_sail), LAI_fourSAIL_iqr = IQR(LAI_sail),
               EWT_fourSAIL = median(EWT_sail), EWT_fourSAIL_iqr = IQR(EWT_sail),
               Cab_INFORM = median(Cab_inform), LAI_INFORM = median(LAI_inform),
               pct_out_of_domain = round(100 * mean(out_of_domain), 1),
               ok = TRUE, msg = "")
  }, error = function(e) data.frame(date = as.Date(w[1]), n_forest_px = NA_real_,
                                     NDVI = NA_real_, NDVI_iqr = NA_real_, NDRE = NA_real_, NDMI = NA_real_,
                                     Cab_fourSAIL = NA_real_, Cab_fourSAIL_iqr = NA_real_,
                                     LAI_fourSAIL = NA_real_, LAI_fourSAIL_iqr = NA_real_,
                                     EWT_fourSAIL = NA_real_, EWT_fourSAIL_iqr = NA_real_,
                                     Cab_INFORM = NA_real_, LAI_INFORM = NA_real_,
                                     pct_out_of_domain = NA_real_, ok = FALSE, msg = conditionMessage(e)))
}

ts_list <- lapply(windows, get_one_date)
ts_df <- do.call(rbind, ts_list)
```

```{r timeseries-check}
print(ts_df[, c("date", "n_forest_px", "NDVI", "Cab_fourSAIL", "Cab_INFORM", "pct_out_of_domain", "ok")])
```

If `2024-03-01`/`2024-04-01` are missing `ok = TRUE` above, those windows
had no cloud-free forest pixels in the archive at this threshold -- a
real, expected data gap, not silently papered over. November's pixel
count is real too, and much lower than the summer months -- low sun
angle and higher cloud frequency both genuinely reduce usable forest
pixels late in the season; treat that date's numbers as noisier for
that reason, visible directly in its wider IQR below.

## 4. A critical first check: does the training LUT even cover what the real scene looks like?

Before reading anything into the trajectory, check the domain-
applicability numbers already computed above (Tutorial 15 Section 5's
same check, applied per date here):

```{r domain-summary}
knitr::kable(ts_df[ts_df$ok, c("date", "n_forest_px", "pct_out_of_domain")], row.names = FALSE)
```

**Every single forest pixel, on every date, falls outside the fourSAIL
training LUT's simulated reflectance range in at least one band.** That
is a real, load-bearing finding, not a formality: it means every
`Cab_fourSAIL`/`LAI_fourSAIL`/`EWT_fourSAIL` value below is an
extrapolation from the model's training data, not an interpolation
within it. With a 10-dimensional band space and a demonstration-scale
1500-row LUT, this is not entirely surprising (covering 10 correlated
dimensions well enough that *no* real pixel falls outside *any* single
band's simulated range needs a much larger and more carefully
constrained LUT than this page uses) -- but it is exactly the kind of
check that should happen *before* trusting a trajectory, which is why
it's presented here, before Section 5's comparison, not after.

## 5. NDVI/NDRE/NDMI vs. hybrid-inverted traits, through time -- and where fourSAIL and INFORM disagree

```{r timeseries-plot, fig.height = 8}
ts_ok <- subset(ts_df, ok)
ts_long <- do.call(rbind, lapply(
  c("NDVI", "Cab_fourSAIL", "Cab_INFORM", "LAI_fourSAIL", "LAI_INFORM", "EWT_fourSAIL"),
  function(v) data.frame(date = ts_ok$date, variable = v, value = ts_ok[[v]])
))
ts_long$variable <- factor(ts_long$variable,
  levels = c("NDVI", "Cab_fourSAIL", "Cab_INFORM", "LAI_fourSAIL", "LAI_INFORM", "EWT_fourSAIL"))

ggplot(ts_long, aes(x = date, y = value)) +
  geom_line(color = "#2E8B57") + geom_point(color = "#2E8B57", size = 2) +
  facet_wrap(~variable, scales = "free_y", ncol = 2) +
  labs(title = "Speulderbos forest, 2024: indices vs. hybrid-inverted traits (fourSAIL vs. INFORM)",
       x = NULL, y = NULL) +
  theme_bw(base_size = 11)
```

**`Cab_fourSAIL` and `Cab_INFORM` do not agree** -- fourSAIL's retrieved
Cab stays nearly flat all season (roughly 7.2-7.4), while INFORM's rises
sharply from May to a mid-summer peak around August before partially
declining. Same real pixels, same real reflectance, two different
canopy models, two qualitatively different seasonal stories. **This is
exactly the lesson this page is built to demonstrate: a physically
based inversion does not automatically make a retrieved trajectory
correct.** Before reaching for a biological explanation for either
curve's shape, three things need checking first -- and two of them
already have real answers above: the LUT (Section 4 -- 100% of pixels
are extrapolations for the fourSAIL model), the model's own structural
assumptions (Tutorial 14 -- INFORM's crown-geometry parameters interact
with LAI/Cab in ways fourSAIL's homogeneous-canopy assumption can't
replicate, and vice versa), and temporal plausibility (a real Cab
trajectory shouldn't swing as sharply as INFORM's does here without an
ecological event to explain it, which nothing in this dataset confirms).
`LAI_fourSAIL` and `LAI_INFORM` agree much better -- both show a mild
summer decline and a partial autumn recovery -- suggesting LAI is more
robust to canopy-model choice than Cab is for this site, at least under
these two training LUTs.

## 6. Uncertainty through time, not point-to-point lines alone

```{r uncertainty-plot}
ggplot(ts_ok, aes(x = date, y = Cab_fourSAIL)) +
  geom_ribbon(aes(ymin = Cab_fourSAIL - Cab_fourSAIL_iqr / 2, ymax = Cab_fourSAIL + Cab_fourSAIL_iqr / 2),
              fill = "#2E8B57", alpha = 0.25) +
  geom_line(color = "#2E8B57") + geom_point(color = "#2E8B57", size = 2) +
  labs(title = "fourSAIL Cab, with per-date spatial IQR across forest pixels",
       subtitle = "Band width = inter-quartile range across pixels, not a formal retrieval error",
       x = NULL, y = "Cab (fourSAIL)") +
  theme_bw(base_size = 11)
```

The IQR band shown is spatial variability across the forest ROI's
pixels on a given date, not a calibrated retrieval uncertainty (that
would need repeated/ensemble RF predictions or a proper error
propagation from the LUT -- out of scope here) -- but it does show which
dates are noisier due to fewer/more heterogeneous forest pixels (October
and November, matching their lower pixel counts in Section 3).

## 7. A cautious phenology read, and what would make it stronger

**Given Section 4's domain-gap finding and Section 5's fourSAIL/INFORM
disagreement, this page does not claim to have identified a real
phenological signal (e.g. beech senescence) from these retrieved
trajectories alone.** What can be said: NDVI declines through the season
(consistent with some combination of canopy senescence and/or sun-angle/
illumination change -- this dataset alone can't separate those), and the
two RTM-based Cab trajectories disagree enough with each other that
neither should be read as ground truth without independent validation.
Attributing either pattern specifically to Speulderbos' deciduous beech
component would need spatial footprint/species composition data this
page doesn't have -- "consistent with seasonal canopy changes, not yet
attributable to a specific mechanism" is the accurate level of claim
here, not a confirmed biological finding.

```{r divergence-check}
cat("Correlation, NDVI vs LAI (fourSAIL):", round(cor(ts_ok$NDVI, ts_ok$LAI_fourSAIL), 2), "\n")
cat("Correlation, NDVI vs Cab (fourSAIL):", round(cor(ts_ok$NDVI, ts_ok$Cab_fourSAIL), 2), "\n")
cat("(", nrow(ts_ok), "real dates -- enough for a directional read, still too few for a firm\n",
    "statistical claim; treat as suggestive, not confirmatory.)\n")
```

## 8. One figure, every trajectory together

Normalized to each variable's own min-max range across the season, so
timing (not magnitude) can be compared directly -- does the NIR-based
water index move with EWT? Does the red-edge index track Cab better than
plain NDVI does?

```{r final-phenology-figure, fig.height = 5}
norm01 <- function(x) (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
final_long <- do.call(rbind, lapply(
  c("NDVI", "NDRE", "NDMI", "Cab_fourSAIL", "LAI_fourSAIL", "EWT_fourSAIL"),
  function(v) data.frame(date = ts_ok$date, variable = v, value = norm01(ts_ok[[v]]))
))
ggplot(final_long, aes(x = date, y = value, color = variable)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  labs(title = "Speulderbos 2024: every trajectory, normalized 0-1",
       subtitle = "Timing comparison only -- magnitudes are not comparable across variables",
       x = NULL, y = "Normalized (min-max per variable)", color = NULL) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
```

## Series complete

```text
01 Getting Started -> 02 Leaf-to-Canopy -> 03 SPART -> 04 Model Comparison
-> 05 LUTs -> 06 Parallel Simulation -> 07 Sensor Convolution
-> 08 Hyperspectral Sensors -> 09 Vegetation Indices -> 10 Sensitivity
-> 11 Hybrid Inversion -> 12 ML Comparison -> 13 Deep Learning
-> 14 End-to-End Pipeline -> 15 Real EO Application
-> 16 MARMIT + SPART Soil-to-Atmosphere -> 17 Forest Time Series (this page)
```

From one hand-written trait row (Tutorial 01) to a real, multi-date
forest monitoring result built entirely on this package's own functions
-- every stage in between real, runnable, and verified against either
known physics or real satellite data. This closing page's own headline
finding is as much a methodological one as a phenological one: two
structurally different, individually reasonable RTM choices, trained on
the same real pixels, disagree on Cab's seasonal shape -- checking the
LUT's domain coverage and comparing across model structure caught that
before it became an unchecked biological claim.
