---
title: "18. Data-Driven Spatial Index Mapping with getSpatial_index()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{18. Data-Driven Spatial Index Mapping with getSpatial_index()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
```

Tutorial 09 introduced `getSpatial_index()` as the spatial (raster-in,
raster-out) counterpart to the tabular index functions -- one real
GeoTIFF in, one index map out. This closing tutorial uses it for real,
but doesn't guess which index to map: it simulates a LUT, inverts `Cab`
from it (Tutorial 12's own rigor), finds which spectral index actually
correlates best with the *retrieved* trait, and only then maps that
winning index -- and `Cab` itself -- spatially, over a real Sentinel-2
image.

```text
Simulated LUT (500 rows)
      |
      v
Hybrid-invert Cab (RF, Tutorial 12)
      |
      v
Rank every spectral index by correlation with Cab
      |
      v
Winning index (data-driven, not assumed)
      |                                    Real Sentinel-2 image (STAC,
      |                                    NL-Loo/Loobos forest, NL)
      |                                          |
      v                                          v
      +------------------ getSpatial_index() ----+
                            |
                            v
              Winning index map + Cab trait map
```

## 1. Simulate a 500-row LUT and invert Cab

Same rigor as Tutorial 12 -- 500 simulations, Sentinel-2A convolution,
Random Forest, held-out test set:

```{r simulate-invert, results = 'hide'}
n_samples <- 500
LUT <- as.data.frame(getLUT(inputs = ToolsRTM::inputsPROSAIL, nLUT = n_samples, setseed = 1))
wl <- 400:2500; rsoil <- rep(0.15, length(wl))
refl <- t(sapply(seq_len(n_samples), function(i) {
  foursail(inputLUT = LUT[i, ], rsoil = rsoil, LeafModel = "PROSPECT-PRO")$rsot
}))
refl_X <- as.data.frame(refl); colnames(refl_X) <- paste0("X", wl); refl_X <- cbind(id = seq_len(n_samples), refl_X)
se2a_full <- suppressMessages(get.spectra.convolved(rfl = refl_X, sensor = "Sentinel2a", plot.spectra = FALSE))
names(se2a_full) <- c("id","B1","B2","B3","B4","B5","B6","B7","B8","B8A","B9","B10","B11","B12")

keep <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")
real_names <- c("B02","B03","B04","B05","B06","B07","B08","B8A","B11","B12")
se2a <- se2a_full[, keep]; names(se2a) <- real_names

set.seed(1)
train_idx <- sample(seq_len(n_samples), size = round(0.7 * n_samples))
train_df <- cbind(LUT[train_idx, ], se2a[train_idx, ])
fit_cab <- get.inversion(data = train_df, depVar = "Cab", inputs = real_names,
                          algorithm = "RF", n.samples = nrow(train_df), seed = 42)
```

## 2. Which index correlates best with Cab? Computed, not assumed

```{r rank-indices}
se2a_indexed <- se2a
names(se2a_indexed) <- c("B2","B3","B4","B5","B6","B7","B8","B8A","B11","B12")  # getIndicesSE2's own naming
indices_full <- suppressMessages(getIndicesSE2(df = se2a_indexed, sensor = "Sentinel-2a",
                                                df.data = NULL, fast.process = TRUE))

# getSpatial_index() only implements a subset of index names -- only rank
# among indices that can actually be mapped spatially in Section 4 below,
# not the full 70-index tabular set Tutorial 09 covers.
spatial_index_names <- c("ARI","ARV2","ARVI","AVI","BAI","BAIS2","CIg","CIgreen","CIre","CR_SWIR",
                          "EVI","GM1","GM2","GNDVI","Greeness","IRECI","MCARI","MCARI1","MCARI2",
                          "MNDVI","MSAVI","MSR","MTVI1","MTVI2","NBR","NDRE","NDVI","NDWI","NDWI2",
                          "OSAVI","PSSRa","PVI","RDVI","REIP1","REIP2","RVI","RedEg1","RedEg2",
                          "Redness","S2REP","SIPI","SR","TCARI","TCARI_OSAVI","TVI","WDRVI")
candidate_names <- intersect(names(indices_full), spatial_index_names)

cor_table <- data.frame(
  index = candidate_names,
  abs_cor_with_Cab = sapply(candidate_names, function(nm) abs(cor(indices_full[[nm]], LUT$Cab[seq_len(n_samples)], use = "complete.obs")))
)
cor_table <- cor_table[order(-cor_table$abs_cor_with_Cab), ]
knitr::kable(head(cor_table, 10), row.names = FALSE, digits = 3)

winning_index <- cor_table$index[1]
cat("Winning index (highest |correlation| with Cab, among those getSpatial_index() supports):", winning_index, "\n")
```

## 3. Retrieve a real Sentinel-2 image: NL-Loo (Loobos), Netherlands

A real ICOS station: **Loobos (NL-Loo)**, an evergreen needleleaf forest
(Scots pine, *Pinus sylvestris*) near Kootwijk on the Veluwe, Gelderland --
not to be confused with Speulderbos (Tutorials 15/17, a different site a
few km away). Planted around 1909 on sand dunes to fight erosion and left
largely unmanaged since, it's one of the world's longest continuously
running eddy-covariance flux-tower records (ICOS station class 2, PI
Michiel van der Molen). Official station coordinates, from the [ICOS
Carbon Portal station page](https://meta.icos-cp.eu/resources/stations/ES_NL-Loo):
52.166447°N, 5.74355°E, 33 m elevation. Same STAC retrieval code already
verified in Tutorials 15/17, a different real site:

```{r stac-retrieve, results = 'hide'}
library(sf); library(terra)
pt <- st_point(c(5.7436, 52.1666)) |> st_sfc(crs = 4326)
scenario <- st_as_sf(data.frame(id = 1), geometry = st_sfc(pt[[1]], crs = 4326))
bbox <- get_bounding_box(scenario, 500)
shape <- st_as_sf(data.frame(id = 1), geometry = st_sfc(st_polygon(list(rbind(
  c(bbox["xmin"], bbox["ymin"]), c(bbox["xmin"], bbox["ymax"]),
  c(bbox["xmax"], bbox["ymax"]), c(bbox["xmax"], bbox["ymin"]),
  c(bbox["xmin"], bbox["ymin"])))), crs = 4326))

sc <- get.satellite_collection(scenario = scenario, collection = "sentinel-2-l2a",
                                cloud_server = "microsoft", n.limit = 20,
                                date_range = c("2024-07-01", "2024-07-31"),
                                cloud_threshold = 40, buffer_size = 500)
cube <- get.sentinel2_cube(sc[[1]], shape = shape, date_range = c("2024-07-01", "2024-07-31"),
                            aggregation_method = "mean", get.dataset = FALSE)
cat("Real cube retrieved:", paste(dim(cube), collapse = " x "), "(rows x cols x bands), bands:",
    paste(names(cube), collapse = ", "), "\n")
```

## 4. `getSpatial_index()` needs a file on disk, and all 12 nominal bands

`getSpatial_index()` reads a GeoTIFF path (not an in-memory object) and
assumes the full nominal 12-band SMAC order (`B01`...`B12`).
`get.sentinel2_cube()` deliberately excludes the two 60m-only bands
(`B01`, `B09`) that don't carry vegetation signal at this resolution
(same convention as Tutorials 15/17) -- filled here with their nearest
real spectral neighbor as a placeholder so the function can run on
genuinely real reflectance rather than needing bands this pipeline never
collects. Neither `B01` nor `B09` enters the winning index formula for
any of Section 2's top candidates, so this placeholder doesn't affect the
result below:

```{r write-tif}
refl <- cube[[real_names]] / 10000

full12 <- c(refl[["B02"]], refl[["B02"]], refl[["B03"]], refl[["B04"]], refl[["B05"]], refl[["B06"]],
            refl[["B07"]], refl[["B08"]], refl[["B8A"]], refl[["B8A"]], refl[["B11"]], refl[["B12"]])
names(full12) <- c("B01","B02","B03","B04","B05","B06","B07","B08","B8A","B09","B11","B12")

tmp_tif <- tempfile(fileext = ".tif")
terra::writeRaster(full12, tmp_tif, overwrite = TRUE)
```

## 5. The winning index, mapped over a real image

```{r spatial-index-map}
idx_map <- getSpatial_index(rasterFiles = tmp_tif, Sensor = "Sentinel2a",
                             SpectraltoCompute = winning_index, factorR = 1)
plot(idx_map[[1]], main = paste0("Loobos forest, ", winning_index, " (data-driven, highest |cor| with Cab)"))
cat(winning_index, "range over the scene:", paste(round(range(terra::values(idx_map[[1]]), na.rm = TRUE), 3), collapse = " to "), "\n")
```

## 6. Cab itself, mapped -- the same per-pixel pattern as Tutorials 15/17

```{r cab-map}
pix_df <- as.data.frame(refl, xy = TRUE, na.rm = FALSE)
ok_rows <- stats::complete.cases(pix_df[, real_names])
Cab_pixels <- rep(NA_real_, nrow(pix_df))
Cab_pixels[ok_rows] <- as.numeric(predict(fit_cab$model, pix_df[ok_rows, real_names]))

cab_map <- refl[["B04"]]
terra::values(cab_map) <- Cab_pixels
names(cab_map) <- "Cab_pred"

op <- par(mfrow = c(1, 2))
plot(idx_map[[1]], main = winning_index)
plot(cab_map, main = "Retrieved Cab")
par(op)

cat("Correlation between the mapped", winning_index, "and mapped Cab, pixel-by-pixel:",
    round(cor(as.numeric(terra::values(idx_map[[1]])), Cab_pixels, use = "complete.obs"), 2), "\n")
```

Two independent pixel-by-pixel outputs over the same real scene -- one a
plain spectral index, the other a full hybrid-inversion trait retrieval
-- their spatial agreement (or disagreement) is itself informative, not
assumed: a strong correlation here means the simple index is largely
capturing the same signal the more expensive ML model is; a weak one
means the ML model is picking up something the index alone misses.

## Series complete

```text
01 Getting Started -> ... -> 17 Forest Time Series -> 18 Spatial Index Mapping (this page)
```

Two real forest sites now mapped across this series -- Speulderbos
(Tutorials 15, 17) and Loobos (this page) -- both real ICOS/flux-tower
locations in the Netherlands, both retrieved live via STAC, both fed
through this package's own hybrid-inversion machinery end to end.
