---
title: "00. Configuring SCOPE: options, defaults, and common errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{00. Configuring SCOPE: options, defaults, and common errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

Every SCOPEinR tutorial from here on opens with the same line:

```r
scope_options <- read.table(system.file("input", "setoptions.csv", package = "SCOPEinR"),
                             header = TRUE, sep = ",")
```

`scope_options` is a 17-row switchboard controlling which physics
SCOPE actually runs -- not tuning parameters (those live in the LUT),
but structural choices: which sub-models execute, which files get read,
which approximations apply. This tutorial is a configuration and
troubleshooting reference, not a lab report: every option's default,
what changing it does, when you'd actually want to, and -- for the ones
this port doesn't fully support yet -- exactly what error you'll see and
what to do about it.

## 1. What should I normally use?

The shipped defaults, unmodified, work:

```{r shipped-defaults}
scope_options <- read.table(system.file("input", "setoptions.csv", package = "SCOPEinR"),
                             header = TRUE, sep = ",")
knitr::kable(scope_options)
```

## 2. The options at a glance

**Important: these are not generic on/off flags.** Most are 0/1 toggles,
but `soilspectrum` selects between two different soil *data sources*,
`Fluorescence_model` and `soil_heat_method` select between three
different *methods* each, and `irradiance` selects between three
different irradiance *sources* -- in every one of those cases, "0" and
"1" (and "2") name specific alternatives, not a universal false/true.
Check the "Alternatives" column below before assuming a value's meaning.

| Option | Default | Alternatives | What does it do? | When should I change it? | Status |
|---|---|---|---|---|---|
| `lite` | 1 | 0 | SCOPE-lite (per-layer) vs. full (per-layer-and-leaf-angle) integration | Use 0 when you need the full angular resolution; 1 is faster | ✅ Both work |
| `calc_fluorescence` | 1 | 0 | Simulate chlorophyll fluorescence (SIF) | Set 0 if you only need reflectance and want a faster run | ✅ Both work |
| `calc_spectrum_planck` | 1 | 0 | Full per-wavelength Planck thermal spectrum vs. simplified thermal | Usually keep default | ✅ Both work |
| `calc_xanthophyllabs` | 1 | 0 | Include xanthophyll/PRI reflectance effects | Disable (0) if not needed, small speedup | ✅ Both work |
| `soilspectrum` | 0 | 1 | Soil reflectance from a file vs. computed by BSM | Use 1 to simulate soil physically (brightness/moisture) rather than from a fixed file | ✅ Both work |
| `Fluorescence_model` | 0 | 1, 2 | Which NPQ/fluorescence-yield submodel | Use 2 for the Magnani 2012 model; 1 has no separate implementation yet | ✅ 0 and 2 work; ⚠️ 1 uses the same model as 0 |
| `applTcorr` | 1 | 0 | Temperature-correct Vcmax/biochemical rate constants | Normally keep on | ✅ Both work |
| `verify` | 1 | 0 | Whether SCOPE checks output against field data | Application-dependent | ✅ No effect on the simulation itself (by design -- a validation step, not a physics option) |
| `mSCOPE` | 0 | 1 | Vertically heterogeneous canopy (multiple leaf-property layers) | Only for mSCOPE-style applications | ✅ Works |
| `simulation` | 0 | 1 | One run per LUT row vs. time-series mode | Only if you have meteo time-series input files, not a LUT | ❌ Needs a different input format this page doesn't provide (see below) |
| `calc_directional` | 0 | 1 | Full BRDF over many view angles | Use 1 for directional/BRDF studies | ✅ Both work |
| `calc_vert_profiles` | 0 | 1 | Return vertical (per-layer) canopy profiles | When per-layer profiles are needed | ✅ Works |
| `soil_heat_method` | 2 | 0, 1 | How soil heat flux (G) is estimated | Application-dependent | ✅ All three work |
| `calc_rss_rbs` | 0 | 1 | Fixed vs. computed soil surface/boundary-layer resistances | Use 1 to compute from conditions instead of fixing from the LUT | ✅ Works |
| `MoninObukhov` | 1 | 0 | Atmospheric-stability correction | Application-dependent | ✅ Both work |
| `LIDF` | 0 | 1 | Leaf angle distribution from LUT (`LIDFa`/`LIDFb`) vs. an external angles file | Use 1 only if you have a real angles file | ✅ Both run and produce different output |
| `irradiance` | 0 | 1, 2 | Irradiance source: SCOPE's own file, user measurements ("WithE"), or a MODTRAN atmospheric file | 1 if you have real measured irradiance; 2 for a real MODTRAN atmosphere file | ✅ 0 and 2 work; ⚠️ 1 works, see the note below |

## 3. What can I change for my application?

- Need SIF output? `calc_fluorescence = 1` (the default).
- Need the full angular (non-lite) energy balance and thermal emission? `lite = 0`.
- Need per-layer vertical profiles of net radiation? `calc_vert_profiles = 1`.
- Need the Magnani 2012 fluorescence/NPQ model instead of the default empirical fit? `Fluorescence_model = 2`.
- Need a physically-simulated (BSM) soil instead of a fixed file? `soilspectrum = 1`.
- Need multi-angular BRDF? `calc_directional = 1`.
- Have real measured irradiance instead of SCOPE's bundled default? `irradiance = 1` (see the caution in Section 5) or a real MODTRAN atmosphere file (`irradiance = 2`).
- Want a faster run and don't need fluorescence or PRI effects? `calc_fluorescence = 0`, `calc_xanthophyllabs = 0`.
- Everything else: leave at the shipped default -- either it has no effect worth changing (`verify`, `applTcorr` rarely needs to move), or it needs different input this page doesn't provide (`simulation = 1`, Section 5.3).

## 4. If it errors: troubleshooting

| Error you see | Likely cause | What to do |
|---|---|---|
| `non-numeric matrix extent` | `simulation = 1` | Needs time-series meteo input files, not a single LUT row -- this page's examples don't provide that input (Section 5.3) |
| `Fluorescence_model = 1` doesn't look different from the default | `Fluorescence_model = 1` | Expected: this port doesn't have a separate "sigmoid Kn" implementation, so it uses the same model as `0` (Section 5.4) |
| Direct/diffuse irradiance (`Esun_`/`Esky_`) look swapped | `irradiance = 1` | Known caveat -- see the note in Section 5.5 |

## 5. Per-option detail: everything not a plain "it works"

### 5.1 `lite = 0` -- Full SCOPE (13x36 angular resolution)

**Purpose:** activates the full per-leaf-angle-class (13x36xnl) integration
instead of SCOPE-lite's per-layer aggregation, for the thermal-emission
and net-radiation calculations.
**What to expect:** `lite = 0` returns a physically sane result --
`Actot` = 6.76 umol/m2/s, close to but distinct from the `lite = 1`
baseline (6.95), consistent with the full angular integration being a
more detailed (not wildly different) representation of the same physics.
**Note for Python users:** `scopeinpython` (this package's Python
sibling) only implements SCOPE-lite; the full per-leaf-angle branches
aren't ported there.

### 5.2 `calc_vert_profiles = 1` -- Vertical canopy profiles

**Purpose:** return per-layer (vertical) profiles of net radiation/
photosynthesis instead of only canopy-integrated totals, in both
`lite = 0` and `lite = 1` mode.
**What to expect:** a real per-layer `Pn1d` profile (one value per
canopy layer, decreasing with depth as light attenuates through the
canopy), with `Actot` unchanged from the baseline -- this option only
adds diagnostic profile output, it doesn't change the canopy-integrated
flux.

### 5.3 `simulation = 1` -- Time-series mode

**Purpose:** run SCOPE driven by a time series of meteorological
observations (a sequence of dates/conditions) instead of one LUT row
per independent run.
**Current status in SCOPEinR:** needs input this page doesn't provide.
**Typical error:** `non-numeric matrix extent`.
**Why:** this mode expects meteorological time-series input files in a
different structure than `LUT_input.csv`'s one-row-per-run layout; the
error is a direct consequence of feeding it a single LUT row instead.
**What should I do?** Keep `simulation = 0` unless you have real
time-series meteo files formatted for that mode.
**Is my input wrong?** For this tutorial's LUT-based examples, yes --
this option needs a fundamentally different input, not a bug fix.

### 5.4 `Fluorescence_model` -- which submodel each value selects

- **`Fluorescence_model = 0`** (default): the empirical (Flexas-fit) model.
- **`Fluorescence_model = 2`**: the Magnani et al. (2012) model
  (`get.biochemical.MD12()`, authored by Federico Magnani) -- gives a
  numerically distinct result from the default (`Actot` differs
  meaningfully from `Fluorescence_model = 0`).
- **`Fluorescence_model = 1`** (documented as "sigmoid for Kn"): this
  port doesn't have a separate implementation for this value, so it
  uses the same model as `Value = 0`. **What should I do?** Treat
  `Fluorescence_model = 1` as equivalent to `0` for now.

### 5.5 `irradiance` -- source selection, and a note on `irradiance = 1`

**`irradiance = 0`** (default): SCOPE's own bundled irradiance file.

**`irradiance = 2`** (MODTRAN atmospheric file): usable end-to-end,
returning a physically sane result -- `Actot` = 7.11 umol/m2/s,
distinct from but comparable to the default (6.95), consistent with
using a genuinely different (MODTRAN-derived) irradiance spectrum. If
you have a real MODTRAN `.atm` file for your own site/date, drop it
into `inst/input/radiationdata/` in place of the bundled example
(`FLEX-S3_std.atm`).

**`irradiance = 1`** (user-supplied "WithE" measurements): runs
without error. **Note:** relative to the default, the mean direct
(`Esun_`) and diffuse (`Esky_`) irradiance values come out swapped --
`Esun_` matches the default's `Esky_` value and vice versa. Confirming
the correct column order requires a real reference "WithE" measurement
file to check against, which this package doesn't bundle. **What
should I do?** Treat `irradiance = 1` output's direct/diffuse split
with caution -- the overall magnitude is not in question, only which
of the two columns is direct vs. diffuse.

### 5.6 `mSCOPE = 1` -- vertically heterogeneous canopy

**Purpose:** simulate a canopy with vertically varying leaf properties
(multiple biochemistry layers) instead of one uniform layer, following
the [mSCOPE](https://github.com/peiqiyang/mSCOPE) multi-layer
parameterization (3 layers, LAI fractions `c(0.5, 1, 1.5)`).
**What to expect:** `mSCOPE = 1` runs and produces real, substantially
different canopy-optical output from the single-layer default (189
fields differ).

### 5.7 `calc_rss_rbs = 1` -- computed soil resistances

**Purpose:** compute soil surface (`rss`) and boundary-layer (`rbs`)
resistances from conditions (soil moisture, LAI) instead of using
fixed values from the LUT.
**What to expect:** `calc_rss_rbs = 1` runs and produces real,
substantially different energy-balance output from the
fixed-resistance default (121 fields differ).

## What's next

- **Tutorial 01** -- run one full simulation end to end with the
  (near-)default options from this tutorial.
- **Tutorial 02** -- the soil and canopy BRDF machinery `soilspectrum`
  feeds into.
- **Tutorial 04** -- fluorescence in depth, including where
  `calc_fluorescence`/`Fluorescence_model` plug in.

## Appendix: how these options were verified

Everything above is backed by actually running `get.SCOPE()` at each
option's documented alternative values and diffing the result against
a baseline, field by field -- not read off the `Info` column alone.
This appendix has the actual verification code, kept here rather than
in the main teaching flow above.

### A structural diff helper

`get.SCOPE()`'s return value is a deeply nested list (`data.rad`,
`data.fluxes`, `data.canopy`, ... -- see Tutorial 01). `flatten_scope()`
walks the whole nested result down to scalars (summarizing longer
numeric vectors by their mean and length), and `compare_runs()` reports
which fields appeared, disappeared, or changed value beyond
floating-point noise.

```{r appendix-helpers}
flatten_scope <- function(x, prefix = "") {
  out <- list()
  if (is.list(x)) {
    nms <- names(x)
    if (is.null(nms)) nms <- seq_along(x)
    for (i in seq_along(x)) {
      nm <- if (is.character(nms)) nms[i] else paste0("[[", nms[i], "]]")
      key <- if (nzchar(prefix)) paste0(prefix, "$", nm) else nm
      out <- c(out, flatten_scope(x[[i]], key))
    }
  } else if (is.numeric(x) || is.logical(x) || is.character(x)) {
    if (length(x) == 1) {
      out[[prefix]] <- x
    } else if (length(x) > 1 && is.numeric(x)) {
      out[[paste0(prefix, "[mean]")]] <- mean(x, na.rm = TRUE)
    }
  }
  out
}

compare_runs <- function(base_flat, alt_flat, tol = 1e-6) {
  added <- setdiff(names(alt_flat), names(base_flat))
  removed <- setdiff(names(base_flat), names(alt_flat))
  changed <- character(0)
  for (nm in intersect(names(base_flat), names(alt_flat))) {
    a <- base_flat[[nm]]; b <- alt_flat[[nm]]
    if (is.numeric(a) && is.numeric(b)) {
      if (is.na(a) != is.na(b) || (!is.na(a) && abs(a - b) > tol * max(1, abs(a)))) {
        changed <- c(changed, nm)
      }
    } else if (!identical(a, b)) {
      changed <- c(changed, nm)
    }
  }
  list(added = added, removed = removed, changed = changed)
}
```

### Testing every option for real

One baseline run, then every option flipped to each of its other
documented values, one option at a time, against the same LUT row.
Runs that error are caught rather than allowed to fail the whole
comparison -- an error is itself a real, reportable result here.

```{r appendix-test}
lut <- read.table(system.file("input", "LUT_input.csv", package = "SCOPEinR"), header = TRUE, sep = ",")

run_scope <- function(opts) {
  tryCatch({
    invisible(capture.output(
      res <- SCOPEinR::get.SCOPE(LUT = lut[1, ], options.SCOPE = opts,
                                  optipar = SCOPEinR::optipar2021.Pro.CX,
                                  leaf.model = "fluspect-CX", canopy.model = "fourSAIL",
                                  get.outputs = "ALL", get.plots = FALSE)
    ))
    list(ok = TRUE, res = res[[1]])
  }, error = function(e) list(ok = FALSE, err = conditionMessage(e)))
}

base_run <- run_scope(scope_options)
base_flat <- flatten_scope(base_run$res)

# 'simulation' isn't included: value 1 fundamentally changes the input format
# (time-series meteo files instead of a LUT row), so it isn't a same-input
# toggle the way the other 16 options are -- not tested here for that reason,
# not because it errors.
alt_values <- list(
  lite = 0, calc_spectrum_planck = 0, calc_xanthophyllabs = 0, soilspectrum = 1,
  Fluorescence_model = c(1, 2), applTcorr = 0, verify = 0, mSCOPE = 1,
  calc_directional = 1, calc_vert_profiles = 1, soil_heat_method = c(0, 1),
  calc_rss_rbs = 1, MoninObukhov = 0, LIDF = 1, irradiance = c(1, 2),
  calc_fluorescence = 0
)

results <- list()
for (opt in names(alt_values)) {
  for (v in alt_values[[opt]]) {
    opts <- scope_options
    opts$Value[opts$Options == opt] <- v
    r <- run_scope(opts)
    if (r$ok) {
      cmp <- compare_runs(base_flat, flatten_scope(r$res))
      status <- "OK"
      n_changed <- length(cmp$added) + length(cmp$removed) + length(cmp$changed)
      headline <- if (n_changed == 0) "no measurable effect" else paste(n_changed, "fields differ")
    } else {
      status <- "ERROR"; n_changed <- NA; headline <- r$err
    }
    results[[length(results) + 1]] <- data.frame(
      option = opt, value = v, status = status, n_changed = n_changed, headline = headline
    )
  }
}
results_df <- do.call(rbind, results)
knitr::kable(results_df, row.names = FALSE)
```

This is the raw evidence behind every status/finding in Sections 2 and
5 above.
