---
title: "01. Getting Started with SCOPEinR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Getting Started with SCOPEinR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

`SCOPEinR` is a full R port of SCOPE 2.0 (Soil Canopy Observation,
Photochemistry and Energy fluxes, van der Tol et al. 2009; Yang et al.
2021) -- it couples soil reflectance, canopy optical BRDF, leaf
photosynthesis, and a full energy-balance closure loop into one
simulation. `ToolsRTM`'s series (this site's other tutorial set) covers
leaf/canopy radiative transfer alone; this series covers what SCOPE adds
on top: energy balance, fluorescence, and photosynthesis, all coupled
together. See the [Trait & LUT Glossary](trait-glossary.html)
for what every LUT column consumed below actually means, its unit, and
its realistic range.

```text
Leaf/canopy/soil/meteo inputs (one LUT row)
              |
              v
      Fluspect-Cx + fourSAIL (optical BRDF)
              |
              v
        Energy balance solve
     (leaf/soil temperature, iterative)
              |
      +-------+-------+
      |               |
      v               v
TOC reflectance   Fluorescence (SIF)
      |               |
      +-------+-------+
              |
              v
    Photosynthesis (Actot)
```

## 1. Run one full SCOPE simulation

```{r step1}
table.with.opts <- read.table(system.file("input", "setoptions.csv", package = "SCOPEinR"),
                               header = TRUE, sep = ",")
Table.LUT <- read.table(system.file("input", "LUT_input.csv", package = "SCOPEinR"),
                         header = TRUE, sep = ",")

invisible(capture.output(
  db.sim <- SCOPEinR::get.SCOPE(
    LUT           = Table.LUT[1, ],
    options.SCOPE = table.with.opts,
    optipar       = SCOPEinR::optipar2021.Pro.CX,
    leaf.model    = "fluspect-CX",
    canopy.model  = "fourSAIL",
    get.outputs   = "ALL",
    get.plots     = FALSE
  )
))
res <- db.sim[[1]]  # get.SCOPE() always returns a list-of-results, even for one LUT row

cat("Canopy layers:", res$data.canopy$nlayers, "\n")
cat("TOC reflectance at 550nm:", round(res$data.rad$refl[550 - 400 + 1], 4), "\n")
cat("TOC reflectance at 800nm:", round(res$data.rad$refl[800 - 400 + 1], 4), "\n")
cat("Canopy-average leaf temperature (Tcave):", round(res$data.fluxes$Tcave, 2), "degC\n")
cat("Net radiation, total (Rntot):", round(res$data.fluxes$Rntot, 1), "W/m2\n")
cat("Canopy photosynthesis (Actot):", round(res$data.fluxes$Actot, 2), "umol m-2 s-1\n")
cat("Canopy fluorescence flux (EoutF):", round(res$data.rad$EoutF, 4), "W/m2\n")
```

**A note on `get.SCOPE()`'s return structure**: unlike `foursail()`
(Tutorial 01 of the ToolsRTM series, a flat `list(rdot=, rsot=)`),
`get.SCOPE()` returns SCOPE's own internal nested-list structure directly
-- `data.rad`, `data.thermal`, `data.fluxes`, `data.canopy`,
`data.spectral`, `iter.ebal`, ... 19 top-level fields. This is deliberate:
it's exactly what the original MATLAB SCOPE model produces, so it stays
directly comparable to SCOPE documentation and literature.

| What you want | Where it is |
|---|---|
| TOC reflectance (full spectrum) | `res$data.rad$refl` |
| Bi-hemispherical / directional components | `res$data.rad$rdd` / `rsd` / `rdo` / `rso` |
| Wavelength grid | `res$data.spectral$wlS` (400-2400nm @ 1nm, then coarser to 50000nm) |
| Fluorescence spectrum | `res$data.rad$LoF_`, on `res$data.spectral$wlF` (640-850nm @ 1nm) |
| Sunlit/shaded leaf temperature per layer | `res$data.thermal$Tcu` / `Tch` |
| Soil temperature | `res$data.thermal$Tsu` / `Tsh` |
| Energy-balance totals | `res$data.fluxes$Rntot` / `lEtot` / `Htot` / `Gtot` |
| Photosynthesis | `res$data.fluxes$Actot` |
| Solver convergence | `res$iter.ebal$counter` (iterations to converge) |

## 2. Four views of the same simulation

```{r step2, fig.width=11, fig.height=8}
wl_optical <- 400:2400  # the uniform-1nm part of data.spectral$wlS
n <- length(wl_optical)

op <- par(mfrow = c(2, 2))

plot(wl_optical, res$data.rad$refl[1:n], type = "l", col = "black", lwd = 1.5,
     xlab = "Wavelength (nm)", ylab = "Reflectance", main = "TOC reflectance components")
lines(wl_optical, res$data.rad$rdd[1:n], col = "steelblue")
lines(wl_optical, res$data.rad$rso[1:n], col = "firebrick")
legend("topright", c("refl (apparent)", "rdd (bi-hemispherical)", "rso (bidirectional)"),
       col = c("black", "steelblue", "firebrick"), lty = 1, cex = 0.7)

wlF <- res$data.spectral$wlF
plot(wlF, res$data.rad$LoF_, type = "l", col = "red3", lwd = 1.5,
     xlab = "Wavelength (nm)", ylab = "Fluorescence radiance (mW m-2 nm-1 sr-1)",
     main = sprintf("TOC fluorescence (F685=%.2f, F740=%.2f)", res$data.rad$F685, res$data.rad$F740))

layers <- seq_len(res$data.canopy$nlayers)
plot(layers, res$data.thermal$Tcu, type = "o", pch = 16, col = "darkorange",
     ylim = range(c(res$data.thermal$Tcu, res$data.thermal$Tch, res$data.thermal$Tsu, res$data.thermal$Tsh)),
     xlab = "Canopy layer (1 = top)", ylab = "Temperature (degC)",
     main = sprintf("Converged temperature profile (%d iterations)", res$iter.ebal$counter))
lines(layers, res$data.thermal$Tch, type = "o", pch = 16, col = "steelblue")
abline(h = res$data.thermal$Tsu, col = "brown", lty = 2)
abline(h = res$data.thermal$Tsh, col = "tan", lty = 2)
legend("topright", c("Tcu (sunlit)", "Tch (shaded)", "Tsu (soil, sunlit)", "Tsh (soil, shaded)"),
       col = c("darkorange", "steelblue", "brown", "tan"), lty = c(1, 1, 2, 2), cex = 0.7)

fluxes <- c(Rntot = res$data.fluxes$Rntot, lEtot = res$data.fluxes$lEtot,
            Htot = res$data.fluxes$Htot, Gtot = res$data.fluxes$Gtot)
barplot(fluxes, col = c("darkorange", "steelblue", "firebrick", "grey60"),
        ylab = "W m-2", main = "Canopy energy-balance totals")
par(op)
```

Reflectance, fluorescence, the converged temperature profile, and where
the absorbed energy actually goes -- all from one `get.SCOPE()` call.
Tutorials 02-04 look at each of these in more depth; Tutorial 05 onward
builds LUTs and runs many simulations at once.

## What's next

- **Tutorial 02** -- the soil and canopy BRDF machinery behind `refl`,
  and SCOPE's own input-table structure.
- **Tutorial 03** -- the energy-balance solve behind the temperature
  profile above.
- **Tutorial 04** -- fluorescence (SIF) in depth.
