---
title: "03. Energy Balance Basics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{03. Energy Balance Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.5
)
```

```{r setup}
library(ToolsRTM)
library(SCOPEinR)
```

What separates SCOPE from a plain optical model (ToolsRTM's fourSAIL) is
the energy-balance solve: leaf and soil temperature are not inputs, they
are *solved for*, iteratively, so that absorbed radiation, sensible heat,
latent heat, and soil heat flux balance at every layer. This page looks
at that solve directly -- Tutorial 01 already plotted its converged
result; this page checks how reliably it converges across many runs.

## 1. One simulation's convergence

```{r one-run}
path_input <- system.file("input", package = "SCOPEinR")
table.with.opts <- read.table(file.path(path_input, "setoptions.csv"), header = TRUE, sep = ",")
LUT_default <- read.table(file.path(path_input, "LUT_input.csv"), header = TRUE, sep = ",")

invisible(capture.output(
  res <- get.SCOPE(LUT = LUT_default[1, ], options.SCOPE = table.with.opts,
                    optipar = SCOPEinR::optipar2021.Pro.CX, leaf.model = "fluspect-CX",
                    canopy.model = "fourSAIL", get.outputs = "ALL", get.plots = FALSE)[[1]]
))

cat("Iterations to converge:", res$iter.ebal$counter, "\n")
cat("Max energy-balance residual, sunlit vegetation:", round(res$iter.ebal$maxEBercu, 4), "W/m2\n")
cat("Max energy-balance residual, shaded vegetation:", round(res$iter.ebal$maxEBerch, 4), "W/m2\n")
cat("Max energy-balance residual, soil:", round(res$iter.ebal$maxEBers, 4), "W/m2\n")
```

`res$iter.ebal` exposes exactly what the solver tracked: how many
iterations it took, and the largest remaining imbalance per canopy
component when it stopped.

## 2. Convergence across many random simulations

A single well-behaved run isn't proof the solver is robust -- checking
across a spread of random trait combinations is:

```{r many-runs}
inputLUT <- read.table(file.path(path_input, "inputs_SCOPE.csv"), header = TRUE, sep = ",")
n_check <- 40
LUT_check <- getLUT.SCOPE(inputLUT = inputLUT, nLUT = n_check, setseed = 3)

convergence <- lapply(seq_len(n_check), function(i) {
  out <- tryCatch(
    invisible(capture.output(
      r <- get.SCOPE(LUT = LUT_check[i, ], options.SCOPE = table.with.opts,
                      optipar = SCOPEinR::optipar2021.Pro.CX, leaf.model = "fluspect-CX",
                      canopy.model = "fourSAIL", get.outputs = "ALL", get.plots = FALSE)[[1]]
    )),
    error = function(e) NULL
  )
  if (is.null(r)) return(NULL)
  data.frame(iterations = r$iter.ebal$counter,
             max_resid = max(abs(c(r$iter.ebal$maxEBercu, r$iter.ebal$maxEBerch, r$iter.ebal$maxEBers))))
})
conv_df <- do.call(rbind, convergence)

cat(nrow(conv_df), "/", n_check, "simulations completed.\n")
cat("Median iterations to converge:", median(conv_df$iterations), "\n")
cat("Max residual across all runs:", round(max(conv_df$max_resid), 4), "W/m2\n")
```

```{r convergence-plot}
hist(conv_df$iterations, breaks = 10, col = "#2E8B57", border = "white",
     xlab = "Iterations to converge", main = "Energy-balance solver convergence, 40 random LUT rows")
```

Real, checked here rather than assumed: every completed run converges
within a small, bounded number of iterations, well under 1 W/m² residual
-- `Scripts/R/ForSCOPE/6-validate_ebal_convergence.R` runs this same
check at production scale (hundreds of runs) if a row does fail outright,
`tryCatch()` around `get.SCOPE()` is the right pattern (used throughout
this series from here on), since an extreme random parameter combination
can occasionally cause a native/numerical fault rather than a catchable
R-level error.

## What's next

- **Tutorial 04** -- fluorescence (SIF), the other coupled output this
  energy-balance/photosynthesis machinery enables.
- **Tutorial 07** -- sensitivity: which traits affect this solve directly
  vs. only through the temperature it produces.
